---
title: "Gaiti_multiome_SummarizedExperiment_EDA"
output: html_notebook
---
# Introduction

Here we explore the SummarizedExperiment object output from peak calling script operated on the Gaiti multiome. The Granges object derived from the SE object is also explored.

# Intall Libraries
```{r install packages, eval = FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("ComplexHeatmap")
```
# Load Libraries
```{r load libraries}
# Load required libraries
library(SummarizedExperiment)
library(ggplot2)
library(pheatmap)
library(tidyverse)
library(here)
library(ggrepel)
library(ComplexHeatmap)
```

# Load marker features: SummarizedExperiment and GRanges Object
```{r load SE and GRanges from getMarkerFeatures}
```


# Load motifEnrichment Results
Here we load the motifEnrichment results (rds files) from running the following script on the computing cluster: `run_07_Gaiti_multiome_motifEnrichment.sh`. The RDS files hold the SummarizedExperiment object.

Note this script takes three parameters:
1. motif set: eg. homer
2. groupBy variable: eg. PIMO_Region
3. comparison name: eg. "PIMOup_EB_vs_PIMOup_TC"

```{r load motifEnrichment results}
# #####################
# motifSet = homer
# groupBy = PIMO_Region
#######################

# PIMOup_EB_vs_PIMOup_TC
motifsUp_PIMOupEBvsPIMOupTC <- readRDS(
  file = here(
    "data",
    "Gaiti_multiome",
    "motifEnrichment_homer_PIMO_Region_PIMOup_EB_vs_PIMOup_TC",
    "motifsUp_PIMORegion_PIMOupEBvsPIMOupTC_homer.rds"
  )
)
motifsDown_PIMOupEBvsPIMOupTC <- readRDS(
  file = here(
    "data",
    "Gaiti_multiome",
    "motifEnrichment_homer_PIMO_Region_PIMOup_EB_vs_PIMOup_TC",
    "motifsDown_PIMORegion_PIMOupEBvsPIMOupTC_homer.rds"
  )
)

# PIMOup_EB_vs_PIMOdown_EB
motifsUp_PIMOupEBvsPIMOdownEB <- readRDS(
  file = here(
    "data",
    "Gaiti_multiome",
    "motifEnrichment_homer_PIMO_Region_PIMOup_EB_vs_PIMOdown_EB",
    "motifsUp_PIMORegion_PIMOupEBvsPIMOdownEB_homer.rds"
  )
)
motifsDown_PIMOupEBvsPIMOdownEB <- readRDS(
  file = here(
    "data",
    "Gaiti_multiome",
    "motifEnrichment_homer_PIMO_Region_PIMOup_EB_vs_PIMOdown_EB",
    "motifsDown_PIMORegion_PIMOupEBvsPIMOdownEB_homer.rds"
  )
)

# PIMOup_PT_vs_PIMOdown_PT
motifsUp_PIMOupPTvsPIMOdownPT <- readRDS(
  file = here(
    "data",
    "Gaiti_multiome",
    "motifEnrichment_homer_PIMO_Region_PIMOup_PT_vs_PIMOdown_PT",
    "motifsUp_PIMORegion_PIMOupPTvsPIMOdownPT_homer.rds"
  )
)
motifsDown_PIMOupPTvsPIMOdownPT <- readRDS(
  file = here(
    "data",
    "Gaiti_multiome",
    "motifEnrichment_homer_PIMO_Region_PIMOup_PT_vs_PIMOdown_PT",
    "motifsDown_PIMORegion_PIMOupPTvsPIMOdownPT_homer.rds"
  )
)

# PIMOup_TC_vs_PIMOdown_TC
motifsUp_PIMOupTCvsPIMOdownTC <- readRDS(
  file = here(
    "data",
    "Gaiti_multiome",
    "motifEnrichment_homer_PIMO_Region_PIMOup_TC_vs_PIMOdown_TC",
    "motifsUp_PIMORegion_PIMOupTCvsPIMOdownTC_homer.rds"
  )
)
motifsDown_PIMOupTCvsPIMOdownTC <- readRDS(
  file = here(
    "data",
    "Gaiti_multiome",
    "motifEnrichment_homer_PIMO_Region_PIMOup_TC_vs_PIMOdown_TC",
    "motifsDown_PIMORegion_PIMOupTCvsPIMOdownTC_homer.rds"
  )
)
```

# Plot TF Rank vs Significance: PIMOup_vs_PIMOdown motifEnrichment from single cell multiome

To better understand motif enrichment results, we can plot the ranking of each motif and their significance. 

To do so we write two functions: 
1. motifsSEtoDF():  converts the SE object to a data frame
2. plotMotifEnrichments(): that takes the data frame and plots the Significance (-log10padj) vs TF rank. 

```{r plotting function for motif enrichment rank vs Significance }
# convert the SE to df for plotting and EDA
motifSEtoDF <- function(motifs) {
  # Convert to data frame for ggplot
  df <- data.frame(TF = rownames(motifs), mlog10Padj = assay(motifs)[, 1])
  df <- df[order(df$mlog10Padj, decreasing = TRUE), ]
  df$rank <- seq_len(nrow(df))
  df$TF_fixed <- sub("_.*", "", df$TF) # remove the _number suffix
df$TF_family <- sub(".*\\.", "", df$TF_fixed) # add a TF family column based on the last suffix
  return(df)
}

# plot TF rank vs significance
# highlight_TF: provide a vector of TFs to highlight if needed
plotMotifEnrichments <- function(df, highlight_TF = NULL) {
  if (length(highlight_TF) > 0) {
    df$highlight <- ifelse(df$TF_fixed %in% highlight_TF, "highlight", "normal")
    gg <- ggplot(df, aes(rank, mlog10Padj)) +
      geom_point(aes(color = highlight), size = 1.5) +
      scale_color_manual(values = c("highlight" = "red", "normal" = "grey70")) +
      ggrepel::geom_label_repel(
        data = subset(df, TF_fixed %in% highlight_TF),
        aes(x = rank, y = mlog10Padj, label = TF_fixed),
        size = 2.5,
        nudge_x = 150,
        nudge_y = 50,
        hjust = 0.5,
        force = 5,
        direction = "both",
        max.overlaps = 15,
        color = "black"
      ) +
      ylab("-log10(P-adj) Motif Enrichment") +
      xlab("Rank Sorted TFs Enriched")

    return(gg)
    
  }else
    gg <- ggplot(df, aes(rank, mlog10Padj, color = mlog10Padj)) +
      geom_point(size = 1.5) +
      ggrepel::geom_label_repel(
        data = df[rev(seq_len(15)), ],
        aes(x = rank, y = mlog10Padj, label = TF_fixed),
        size = 2.5,
        nudge_x = 150,
        nudge_y = 50,
        hjust = 0.5,
        force = 5,
        direction = "both",
        max.overlaps = 15,
        color = "black"
      ) +
      ylab("-log10(P-adj) Motif Enrichment") +
      xlab("Rank Sorted TFs Enriched")
  
  return(gg)
}
```

## motifSet = homer (PIMO_Region comparisons)
```{r plot}
# convert SE to DF
motifsUp_PIMOupEBvsPIMOupTC.df <- motifSEtoDF(motifsUp_PIMOupEBvsPIMOupTC)
motifsDown_PIMOupEBvsPIMOupTC.df <- motifSEtoDF(motifsDown_PIMOupEBvsPIMOupTC)

motifsUp_PIMOupEBvsPIMOdownEB.df <- motifSEtoDF(motifsUp_PIMOupEBvsPIMOdownEB)
motifsDown_PIMOupEBvsPIMOdownEB.df <- motifSEtoDF(motifsDown_PIMOupEBvsPIMOdownEB)

motifsUp_PIMOupPTvsPIMOdownPT.df <- motifSEtoDF(motifsUp_PIMOupPTvsPIMOdownPT)
motifsDown_PIMOupPTvsPIMOdownPT.df <- motifSEtoDF(motifsDown_PIMOupPTvsPIMOdownPT)

motifsUp_PIMOupTCvsPIMOdownTC.df <- motifSEtoDF(motifsUp_PIMOupTCvsPIMOdownTC)
motifsDown_PIMOupTCvsPIMOdownTC.df <- motifSEtoDF(motifsDown_PIMOupTCvsPIMOdownTC)


# highlight motifs of interest
highlight_TF <- c("HIF1A","ARNT", "SOX2", "SOX3", "SOX4", 
                  "SOX6", "SOX9", "SOX10", "SOX15", "Oct4:SOX17")
plotMotifEnrichments(motifsUp_PIMOupEBvsPIMOupTC.df, highlight_TF) +
  ggtitle("motifsUp homer")
plotMotifEnrichments(motifsDown_PIMOupEBvsPIMOupTC.df, highlight_TF) +
  ggtitle("motifsDown homer")
```
## homer motif cleaning
```{r homer motif cleaning function}
# Function for additional cleaning of motifs (homer multiome syntax needs to match homer bulk-ATAC-seq)
homer_motif_cleaning <- function(df){
  df_cleaned <- df %>%
  mutate(
    TF_cleaned = TF %>%
      # Step 1: Remove the numerical suffix (e.g., _251)
      # "AP.1.bZIP_1" -> "AP.1.bZIP"
      str_remove("_\\d+$") %>%
      
      # Step 2: Handle the specific edge cases first.
      # "Brachyury.T.box" -> "Brachyury"
      # "Brn2-POU" -> "Brn2"
      str_remove("\\.T\\.box$") %>%
      str_remove("-Homeobox$") %>%
      
      # Step 3: Now, handle the general case by removing the last remaining 
      # motif identifier (e.g., .bZIP, .DR1).
      # "AP.1.bZIP" -> "AP.1"
      # "RXR.NR..DR1" -> "RXR.NR.."
      str_remove("\\.[^.]+$") %>%
      
      # Step 4: Replace all remaining periods with hyphens.
      # "AP.1" -> "AP-1"
      # "RXR.NR.." -> "RXR-NR--"
      str_replace_all("\\.", "-") %>%
      
      # Step 5: Final cleanup. Remove any trailing hyphens.
      # "RXR-NR--" -> "RXR-NR"
      str_remove("-+$") %>%
    
      # Step 6: Additional edge cases:
      str_remove("-Paired$") %>%
      str_remove("-NR$") %>%
      str_remove("-Forkhead$") %>%
      str_remove("-Homeobox$") %>%
      str_remove("-POU$") %>%
      str_replace_all("Arnt-Ahr","Arnt:Ahr") %>%
      str_replace_all("bZIP-IRF-bZIP","bZIP:IRF") %>%
      str_replace_all("CTCF-SatelliteElement-Zf", "CTCF-SatelliteElement") %>%
      str_replace_all("EWS-FLI1-fusion","EWS:FLI1-fusion") %>% 
      str_replace_all("Fox-Ebox", "Fox:Ebox") %>%
      str_replace_all("GRE-NR","GRE") %>%
      str_replace_all("Pit1-1bp","Pit1+1bp") %>%
      str_replace_all("Tlx","Tlx?") %>%
      str_replace_all("OCT-OCT","OCT:OCT") %>%
      str_replace_all("RAR-RXR","RAR:RXR") %>%
      str_replace_all("ZNF143-STAF","ZNF143|STAF") %>%
      str_replace_all("OCT4-SOX2-TCF-NANOG-POU-Homeobox","OCT4-SOX2-TCF-NANOG") %>%
      str_replace_all("(Nkx\\d+)-(\\d+)", "\\1.\\2") # for Nkxn-n, where n is any number
  )
return (df_cleaned)    
}

# Improved plotMotifEnrichments function with additional parameters:

plotMotifEnrichments_new <- function(df,
                                     common_tfs_vector,
                                     show_labels = TRUE,
                                     show_labels_qty = 25,
                                     padj_threshold = 0.05) {
  
  # Convert p-adj threshold to -log10 scale
  # e.g., 0.05 becomes ~1.301
  mlog10_thresh <- -log10(padj_threshold)

  # 1. Flag common TFs
  df <- df %>%
    mutate(is_common = TF_cleaned %in% common_tfs_vector)
  
  num_is_common <- sum(df$is_common, na.rm = TRUE)
  message(paste0("Common motifs: ", num_is_common, " out of ", nrow(df)))
  
  # 2. Base Plot
  gg <- ggplot(df, aes(x = rank, y = mlog10Padj)) + 
    # Add a horizontal line for the significance threshold
    geom_hline(yintercept = mlog10_thresh, linetype = "dashed", color = "darkred", alpha = 0.5) +
    geom_point(aes(color = is_common), shape = 16, size = 2, alpha = 0.8) +
    scale_color_manual(
      name = "Common TF", 
      values = c(`TRUE` = "red", `FALSE` = "grey50"),
      labels = c(`TRUE` = "Yes", `FALSE` = "No")
    ) +
    labs(
      y = "-log10(P-adj) Motif Enrichment",
      x = "Rank Sorted TFs Enriched",
      caption = paste0("Dashed line: p-adj < ", padj_threshold)
    ) +
    theme_minimal() +
    guides(color = guide_legend(override.aes = list(size = 4)))

  # 3. Conditional Labeling with Threshold Filter
  if (show_labels) {
    # Filter for significance first
    significant_df <- df %>% filter(mlog10Padj >= mlog10_thresh)
    
    # Determine which data to label (Common TFs vs Top Hits)
    label_data <- if(any(significant_df$is_common)) {
      significant_df %>% filter(is_common) %>% slice_head(n = show_labels_qty)
    } else {
      significant_df %>% slice_head(n = show_labels_qty)
    }
    
    # Only attempt to add labels if there is data passing the threshold
    if(nrow(label_data) > 0) {
      gg <- gg + geom_text_repel(
        data = label_data,
        aes(label = TF_cleaned),
        size = 3,
        force = 10,
        max.overlaps = 15,
        color = "black",
        segment.color = "grey50",
        direction = "both",
        nudge_x = 100,
        nudge_y = 10,
        box.padding = 0.5,
        point.padding = 0.3
      )
    } else {
      message("No motifs passed the padj_threshold for labeling.")
    }
  }
  
  return(gg)
}
```

## Test motif cleaning and new plotting function 
```{r}
motifsUp_PIMOupEBvsPIMOupTC.df %>%
            homer_motif_cleaning() %>%
  plotMotifEnrichments_new(., common_tfs_vector=NULL) +
  ggtitle("Gaiti Multiome: MotifsUP in PIMOup vs PIMOdown", subtitle = "Top 25 Significant Motifs")
```

## Define final plotting function
Here we improve on the plotting function, building upon plotMotifEnrichments(), and plotMotifEnrichments_new().

```{r plotMotifEnrichments_final}
plotMotifEnrichments_final <- function(df,
                                       common_tfs_vector = NULL,
                                       show_labels = TRUE,
                                       show_labels_qty = 25,
                                       padj_threshold = 0.05) {
  
  # Convert p-adj threshold to -log10 scale
  mlog10_thresh <- -log10(padj_threshold)

  # 1. Flag common TFs and handle NULL vector
  if(is.null(common_tfs_vector)) common_tfs_vector <- character(0)
  df <- df %>% mutate(is_common = TF_cleaned %in% common_tfs_vector)
  
  # 2. Base Plot
  gg <- ggplot(df, aes(x = rank, y = mlog10Padj)) + 
    geom_hline(yintercept = mlog10_thresh, linetype = "dashed", color = "darkred", alpha = 0.4) +
    geom_point(aes(color = is_common), shape = 16, size = 2, alpha = 0.7) +
    scale_color_manual(
      name = "Common TF", 
      values = c(`TRUE` = "red", `FALSE` = "grey60"),
      labels = c(`TRUE` = "Yes", `FALSE` = "No")
    ) +
    scale_x_continuous(expand = expansion(mult = c(0.05, 0.25))) + # Extra space for labels
    labs(y = "-log10(P-adj) Enrichment", x = "Rank Sorted TFs") +
    theme_minimal() +
    theme(panel.grid.minor = element_blank())

  # 3. Text Labeling (Shifted Right)
  if (show_labels) {
    # Only label those passing threshold
    significant_df <- df %>% filter(mlog10Padj >= mlog10_thresh)
    
    label_data <- if(any(significant_df$is_common)) {
      significant_df %>% filter(is_common) %>% slice_head(n = show_labels_qty)
    } else {
      significant_df %>% slice_head(n = show_labels_qty)
    }
    
    if(nrow(label_data) > 0) {
      gg <- gg + geom_text_repel(
        data = label_data,
        aes(label = TF_cleaned),
        size = 3.2,
        color = "black",
        # Push to the right sidebar
        force = 5,
        nudge_x = max(df$rank, na.rm = TRUE) * 0.15,
        direction = "y",
        hjust = 0,
        max.overlaps = 30,
        box.padding = 0.3,
        point.padding = 0.5,
        min.segment.length = 0,
        segment.color = "grey70"
      )
    }
  }
  return(gg)
}

```
## plot all the conditions
```{r plot gg plots}
# Up-regulated
p1 <- motifsUp_PIMOupEBvsPIMOupTC.df %>%
  homer_motif_cleaning() %>%
  plotMotifEnrichments_final(common_tfs_vector = NULL) +
  ggtitle("Gaiti Multiome: Motifs UP (PIMOup_EB vs PIMOup_TC)", subtitle = "Top 25 Significantly enriched motifs")

# Down-regulated
p2 <- motifsDown_PIMOupEBvsPIMOupTC.df %>%
  homer_motif_cleaning() %>%
  plotMotifEnrichments_final(common_tfs_vector = NULL) +
  ggtitle("Gaiti Multiome: Motifs Down (PIMOup_EB vs PIMOup_TC)", subtitle = "Top 25 Significantly enriched motifs")

# EB Comparison
p3 <- motifsUp_PIMOupEBvsPIMOdownEB.df %>%
  homer_motif_cleaning() %>%
  plotMotifEnrichments_final(common_tfs_vector = NULL) +
  ggtitle("Gaiti Multiome: Motifs UP (PIMOup vs PIMOdown in EB)", subtitle = "Top 25 Significantly enriched motifs")

p4 <- motifsDown_PIMOupEBvsPIMOdownEB.df %>%
  homer_motif_cleaning() %>%
  plotMotifEnrichments_final(common_tfs_vector = NULL) +
  ggtitle("Gaiti Multiome: Motifs DOWN (PIMOup vs PIMOdown in EB)", subtitle = "Top 25 Significantly enriched motifs")

# PT Comparison
p5 <- motifsUp_PIMOupPTvsPIMOdownPT.df %>%
  homer_motif_cleaning() %>%
  plotMotifEnrichments_final(common_tfs_vector = NULL) +
  ggtitle("Gaiti Multiome: Motifs UP (PIMOup vs PIMOdown in PT)", subtitle = "Top 25 Significantly enriched motifs")

p6 <- motifsDown_PIMOupPTvsPIMOdownPT.df %>%
  homer_motif_cleaning() %>%
  plotMotifEnrichments_final(common_tfs_vector = NULL) +
  ggtitle("Gaiti Multiome: Motifs DOWN (PIMOup vs PIMOdown in PT)", subtitle = "Top 25 Significantly enriched motifs")

# TC Comparison
p7 <- motifsUp_PIMOupTCvsPIMOdownTC.df %>%
  homer_motif_cleaning() %>%
  plotMotifEnrichments_final(common_tfs_vector = NULL) +
  ggtitle("Gaiti Multiome: Motifs UP (PIMOup vs PIMOdown in TC)", subtitle = "Top 25 Significantly enriched motifs")

p8 <- motifsDown_PIMOupTCvsPIMOdownTC.df %>%
  homer_motif_cleaning() %>%
  plotMotifEnrichments_final(common_tfs_vector = NULL) +
  ggtitle("Gaiti Multiome: Motifs DOWN (PIMOup vs PIMOdown in TC)", subtitle = "Top 25 Significantly enriched motifs")

p1
p2
p3
p4
p5
p6
p7
p8

```
# export pdfs of plots
```{r export pdfs of plots, eval = FALSE}
library(patchwork)
here()
pdf(here("Plots","2025-12-19_Gaiti_multiome_PIMOup_vs_PIMOdown_PIMO_Region_homer_motif_ranking.pdf"))

p1
p2
p1+p2
p3
p4
p3+p4
p5
p6
p5+p6
p7
p8
p7+p8

dev.off()
```

## export motifs for supplemental table
```{r export motifs for supplement, eval = FALSE}
# convert SE to DF
motifsUp_PIMOupEBvsPIMOupTC.df_cleaned <- motifsUp_PIMOupEBvsPIMOupTC.df %>%
  homer_motif_cleaning()
motifsDown_PIMOupEBvsPIMOupTC.df_cleaned <- motifsDown_PIMOupEBvsPIMOupTC.df %>%
  homer_motif_cleaning()

motifsUp_PIMOupEBvsPIMOdownEB.df_cleaned <- motifsUp_PIMOupEBvsPIMOdownEB.df %>%
  homer_motif_cleaning()
motifsDown_PIMOupEBvsPIMOdownEB.df_cleaned <- motifsDown_PIMOupEBvsPIMOdownEB.df %>%
  homer_motif_cleaning()

motifsUp_PIMOupPTvsPIMOdownPT.df_cleaned <- motifsUp_PIMOupPTvsPIMOdownPT.df %>%
  homer_motif_cleaning()
motifsDown_PIMOupPTvsPIMOdownPT.df_cleaned <- motifsDown_PIMOupPTvsPIMOdownPT.df %>%
  homer_motif_cleaning()

motifsUp_PIMOupTCvsPIMOdownTC.df_cleaned <- motifsUp_PIMOupTCvsPIMOdownTC.df %>%
  homer_motif_cleaning()
motifsDown_PIMOupTCvsPIMOdownTC.df_cleaned <- motifsDown_PIMOupTCvsPIMOdownTC.df %>%
  homer_motif_cleaning()

write_csv(
  motifsUp_PIMOupTCvsPIMOdownTC.df_cleaned,
  here(
    "data",
    "Gaiti_multiome",
    "motifEnrichment_homer_PIMO_Region_PIMOup_TC_vs_PIMOdown_TC",
    "motifsUp_PIMORegion_PIMOupTCvsPIMOdownTC_homer.csv"
  )
)
write_csv(
  motifsDown_PIMOupTCvsPIMOdownTC.df_cleaned,
  here(
    "data",
    "Gaiti_multiome",
    "motifEnrichment_homer_PIMO_Region_PIMOup_TC_vs_PIMOdown_TC",
    "motifsDown_PIMORegion_PIMOupTCvsPIMOdownTC_homer.csv"
  )
)

write_csv(
  motifsUp_PIMOupEBvsPIMOdownEB.df_cleaned,
  here(
    "data",
    "Gaiti_multiome",
    "motifEnrichment_homer_PIMO_Region_PIMOup_EB_vs_PIMOdown_EB",
    "motifsUp_PIMORegion_PIMOupEBvsPIMOdownEB_homer.csv"
  )
)
write_csv(
  motifsDown_PIMOupEBvsPIMOdownEB.df_cleaned,
  here(
    "data",
    "Gaiti_multiome",
    "motifEnrichment_homer_PIMO_Region_PIMOup_EB_vs_PIMOdown_EB",
    "motifsDown_PIMORegion_PIMOupEBvsPIMOdownEB_homer.csv"
  )
)
```

# Summarize significant motifs and compare across PIMO_Region conditions
```{r motif summary}
motifsUp_PIMOupEBvsPIMOupTC.df %>%
  homer_motif_cleaning() %>% nrow()

motif_summary <- function(df, padj_threshold=0.05){
# Convert p-adj threshold to -log10 scale
  # e.g., 0.05 becomes ~1.301
  mlog10_thresh <- -log10(padj_threshold)
  
  df <- homer_motif_cleaning(df) # clean motifs
  num_motif <- nrow(df) # count motifs

print(paste("The number of motifs in this df is:", num_motif))

num_sig_motif <- df %>%
  filter(mlog10Padj > mlog10_thresh) %>% nrow

print(paste("The number of significant motifs padj<",padj_threshold, "is", num_sig_motif))

# return the significant motifs

sig_motif <- df %>% 
  filter(mlog10Padj > mlog10_thresh) %>%
  pull(TF_cleaned)

return(sig_motif) # returns a vector of the significant motifs

}



# significant motifs for each comparison padj < 0.05
motif_summary(motifsUp_PIMOupEBvsPIMOupTC.df) # 14
motif_summary(motifsDown_PIMOupEBvsPIMOupTC.df) # 36

motif_summary(motifsUp_PIMOupEBvsPIMOdownEB.df) # 26
motif_summary(motifsDown_PIMOupEBvsPIMOdownEB.df) # 0

motif_summary(motifsUp_PIMOupPTvsPIMOdownPT.df) # 0
motif_summary(motifsDown_PIMOupPTvsPIMOdownPT.df) # 0

motif_summary(motifsUp_PIMOupTCvsPIMOdownTC.df) # 158
motif_summary(motifsDown_PIMOupTCvsPIMOdownTC.df) # 49
```
## Venn Diagram:
### motifsUp
```{r venn diagram for motifs}
library(VennDiagram)
library(RColorBrewer)
myColours <- brewer.pal(3,"Pastel2") # colors

# list of significant motifs
motifsUp_lt <- list(
  PIMOupEBvsPIMOupTC = motif_summary(motifsUp_PIMOupEBvsPIMOupTC.df),
  PIMOupEBvsPIMOdownEB = motif_summary(motifsUp_PIMOupEBvsPIMOdownEB.df),
  PIMOupTCvsPIMOdownTC = motif_summary(motifsUp_PIMOupTCvsPIMOdownTC.df)
)

motifsDown_lt <- list(
  PIMOupEBvsPIMOupTC = motif_summary(motifsDown_PIMOupEBvsPIMOupTC.df),
  PIMOupEBvsPIMOdownEB = motif_summary(motifsDown_PIMOupEBvsPIMOdownEB.df),
  PIMOupTCvsPIMOdownTC = motif_summary(motifsDown_PIMOupTCvsPIMOdownTC.df)
)

grid.newpage()
pushViewport(viewport(width=unit(0.8, "npc"), height = unit(0.8, "npc")))
venn.diagram(
  motifsUp_lt,
  NULL,
  fill = myColours[1:3],
  cex = 2,
  cat.cex = 0.6,
  lty = "blank",
  category.names = c(
    "PIMOupEBvsPIMOupTC",
    "PIMOupEBvsPIMOdownEB",
    "PIMOupTCvsPIMOdownTC"
  ),
  cat.default.pos = "outer",
  main = "Gaiti_multiome: PIMO_Region motifsUp"
) %>%
grid.draw()

grid.newpage()
pushViewport(viewport(width=unit(0.8, "npc"), height = unit(0.8, "npc")))
venn.diagram(
  motifsDown_lt,
  NULL,
  fill = myColours[1:3],
  cex = 2,
  cat.cex = 0.6,
  lty = "blank",
  category.names = c(
    "PIMOupEBvsPIMOupTC",
    "PIMOupEBvsPIMOdownEB",
    "PIMOupTCvsPIMOdownTC"
  ),
  cat.default.pos = "outer",
  main = "Gaiti_multiome: PIMO_Region motifsDown"
) %>%
grid.draw()
  

```
### motifsDown


### Export pdf of venn diagram
```{r export pdf venn, eval = FALSE}
pdf(here("Plots","2026-01-05_Gaiti_multiome_motifsUP_PIMOup_vs_PIMOdown_PIMO_Region_Venn.pdf"))
grid.newpage()
pushViewport(viewport(width=unit(0.8, "npc"), height = unit(0.8, "npc")))
venn.diagram(
  motifsUp_lt,
  NULL,
  fill = myColours[1:3],
  cex = 2,
  cat.cex = 0.6,
  lty = "blank",
  category.names = c(
    "PIMOupEBvsPIMOupTC",
    "PIMOupEBvsPIMOdownEB",
    "PIMOupTCvsPIMOdownTC"
  ),
  cat.default.pos = "outer",
  main = "Gaiti_multiome: PIMO_Region motifsUp"
) %>%
grid.draw()


dev.off()

pdf(here("Plots","2026-01-08_Gaiti_multiome_motifsDown_PIMOup_vs_PIMOdown_PIMO_Region_Venn.pdf"))
grid.newpage()
pushViewport(viewport(width=unit(0.8, "npc"), height = unit(0.8, "npc")))
venn.diagram(
  motifsDown_lt,
  NULL,
  fill = myColours[1:3],
  cex = 2,
  cat.cex = 0.6,
  lty = "blank",
  category.names = c(
    "PIMOupEBvsPIMOupTC",
    "PIMOupEBvsPIMOdownEB",
    "PIMOupTCvsPIMOdownTC"
  ),
  cat.default.pos = "outer",
  main = "Gaiti_multiome: PIMO_Region motifsDown"
) %>%
grid.draw()


dev.off()
```
## Extract Overlaps from Venn diagram

```{r extract overlaps}
overlaps <- calculate.overlap(motifsUp_lt)
print(overlaps)

calculate.overlap(motifsDown_lt) %>%
print()
```


