---
title: "human_multiome_deeptools_heatmap"
author: "Ronald Wu"
date: "2026-03-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Purpose
To uncover the identity of the regions in a heatmap of pseudo-bulked chromatin accessibility generated by deeptools. Some of these regions show high accessible signal on the heatmap and warrent further analysis. To do so, we output the files used to generate the heatmap and calculate the pseudo-bulked scATAC-seq signal from the set of regions (bed) used to generate heatmaps in deeptools. 

The dataset we are using here is from the human multiome from the PIMO project, specifically on the subset of malignant tumour cells assigned both PIMOhigh (expressing the high PIMO signature score) and also assigned one of three metaprograms derived from cNMF analysis (k3_program).
# Load Libraries
```{r load libraries}
# Load required libraries
library(SummarizedExperiment)
library(tidyverse)
library(here)
```

# Load deeptools computeMatrix Results
Here we load the output from deeptools computeMatrix(), which is normally passed to plotHeatmap() in deeptools.
There are two output files:
1. A matrix.gz file used for plotProfile() and plotHeatmap() within deeptools
2. A matrix IndividualValues.tab that can be loaded into R.
3. A bed file, with regions are saved after skiping zeros or min/max threshold values. The order of the regions in the file follows the sorting order selected.

```{r load compute matrix}
# Define the file path
tab_path <- here("Plots", 
                    "human_multiome",
                    "human_multiome_hmmp_k3_program",
                    "deeptools_k3_program_PIMO_MP",
                    "matrix_refpt_k3_program_PIMO_MP.tab")
bed_path <- here("Plots", 
                    "human_multiome",
                    "human_multiome_hmmp_k3_program",
                    "deeptools_k3_program_PIMO_MP",
                    "regions_refpt_k3_program_PIMO_MP.bed")

# 1. Load the BED file (contains gene names and coordinates)
# The file includes a header starting with '#'
regions_bed <- read_tsv(
   bed_path,
  show_col_types = FALSE
) %>%
  rename(chr = `#chrom`) # Clean up the header symbol

# 2. Load the Signal Matrix (.tab)
# We skip the first 3 lines which contain deepTools metadata and column headers
# PIMO_MP_1_genes.bed, etc. 
signal_matrix <- read_tsv(
  tab_path,
  skip = 3,
  col_names = FALSE,
  show_col_types = FALSE
)

# EDA
regions_bed %>% dim() # [1] 142  13
signal_matrix %>% dim() # [1]  142 1200

# 3. Align and Rank
# We use bind_cols because deepTools preserves the row order between these files
ranked_genes <- regions_bed %>%
  bind_cols(signal_matrix) %>%
  # Calculate the mean across all signal columns (everything after the BED columns)
  # rowMeans is used here for computational efficiency
  mutate(avg_signal = rowMeans(select(., starts_with("X")), na.rm = TRUE)) %>%
  # Select relevant metadata and our new calculation
  select(chr, start, end, name, deepTools_group, avg_signal) %>%
  # Sort by highest signal
  arrange(desc(avg_signal))

# 4. View top 20 genes
head(ranked_genes, 20)

# Optional: Save the ranked list for downstream ArchR/Seurat analysis
# write_csv(ranked_genes, "GBM_ranked_signal_genes.csv")

```
Note: data have not been split for the program1, program2 and program3.

# calculate average signal per group per gene per program
Recall that only a subset of malignant cells were used for this analysis, and they were seperated into three programs:
program1, program2, and program3.

Some genes like VEGFA are only found in PIMO_MP_1, but their signal is investigated in all the cells assigned to the three programs. 

Some genes like MALAT1 are present in all three metaprograms PIMO_MP_1, PIMO_MP_2, and PIMO_MP_3, and their signal is investigated in all the cells assigned to all three programs.

```{r calculate average signal per group per gene}
library(tidyverse)
library(ggrepel)

# 1. Load the BED file for group assignments and gene names
regions_bed <- read_tsv(
  bed_path, 
  show_col_types = FALSE
) %>%
  rename(chr = `#chrom`)

# 2. Load the Matrix using line 3 as column names
# skip = 2 skips lines 1-2; line 3 becomes the header
matrix_tab <- read_tsv(
  tab_path, 
  skip = 2, 
  show_col_types = FALSE
)

# 3. Clean and Align Data
# We remove the first 3 columns of the matrix (which contain group size metadata)
# then bind it with the BED metadata
combined_data <- regions_bed %>%
  bind_cols(select(matrix_tab, -1:-3)) %>%
  # Pivot signal columns into long format
  pivot_longer(
    cols = -colnames(regions_bed), 
    names_to = "sample_group", 
    values_to = "signal"
  ) %>%
  # Strip the numeric suffix read_tsv adds to duplicate headers (e.g. program_1...10)
  mutate(sample_group = str_remove(sample_group, "\\.\\.\\.\\d+$")) %>%
  mutate(sample_group = str_extract(sample_group, "^[^-]+")) %>%
  # Calculate the single average signal per gene
  group_by(name, deepTools_group, sample_group) %>%
  summarize(avg_signal = mean(signal, na.rm = TRUE), .groups = "drop") %>%
  # Rank within each group
  group_by(deepTools_group) %>%
  mutate(rank = min_rank(desc(avg_signal))) %>%
  ungroup()

# 4. Identify top 5 labels for ggrepel

top_labels <- combined_data %>%
  group_by(deepTools_group, sample_group) %>%
  slice_max(avg_signal, n = 5) %>%
  ungroup()

# 5. Plot results

ggplot(combined_data, aes(x = rank, y = avg_signal)) +
  geom_point(aes(color = deepTools_group), alpha = 0.5, size = 1.2) +
  geom_text_repel(
    data = top_labels, 
    aes(label = name),
    size = 3.5,
    fontface = "bold",
    box.padding = 0.5,
    max.overlaps = 20,
    segment.color = "grey50"
  ) +
  facet_wrap(~ deepTools_group + sample_group) +
  theme_minimal() +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold", size = 11),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Ranked Pseudo-bulk ATAC Signal by Program",
    subtitle = "Top 5 Genes for each program assignment",
    x = "Signal Rank (1 = Highest)",
    y = "Average Signal Intensity (TSS +/- 2kb)"
  )
```

# Save Ranked Average Signal plots

```{r save pdf plots, eval = FALSE}
pdf(here("Plots","2026-03-02_human_multiome_hmmp_refpt_k3_program_PIMO_MP_Ranked_Signal_Plot.pdf"))
ggplot(combined_data, aes(x = rank, y = avg_signal)) +
  geom_point(aes(color = deepTools_group), alpha = 0.5, size = 1.2) +
  geom_text_repel(
    data = top_labels, 
    aes(label = name),
    size = 3.5,
    fontface = "bold",
    box.padding = 0.5,
    max.overlaps = 20,
    segment.color = "grey50"
  ) +
  facet_wrap(~ deepTools_group + sample_group) +
  theme_minimal() +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold", size = 11),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Ranked Pseudo-bulk ATAC Signal by Program",
    subtitle = "Top 5 Genes for each program assignment",
    x = "Signal Rank (1 = Highest)",
    y = "Average Signal Intensity (TSS +/- 2kb)"
  )
dev.off()

```

# export combined_data of average signals
```{r export combined_data, eval = FALSE}
combined_data
write_csv(combined_data,
          here("Plots","2026-03-02_human_multiome_hmmp_refpt_k3_program_PIMO_MP_Ranked_Signal_table.csv")
          )
```