---
title: "human_multiome_SummarizedExperiment_EDA"
output: html_notebook
---

# Introduction

Here we explore the SummarizedExperiment object output from peak calling script operated on the human multiome. The Granges object derived from the SE object is also explored.

# Intall Libraries
```{r install packages, eval = FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")


#if (!require("BiocManager", quietly = TRUE)) install.packages("BiocManager")

BiocManager::install(c("SummarizedExperiment", "pheatmap", "tidyverse", "here", "ggrepel", "ComplexHeatmap"))
```
# Load Libraries
```{r load libraries}
# Load required libraries
library(SummarizedExperiment)
library(ggplot2)
library(pheatmap)
library(tidyverse)
library(here)
library(ggrepel)
library(ComplexHeatmap)
```
# Load marker features: SummarizedExperiment and GRanges Object
Here we load the results of differential peak analysis conducted using ArchR on the HPC cluster. The results are from ArchR::getMarkerFeatures, which returns a SE object. Also, we load the derived the SimpleList of GRanges object, which was obtained using ArchR::getMarkers() on the SE object. The cutOff param was set so there was no threshold cut off to get all the features.

```{r load SE and GRanges from getMarkerFeatures}
markersPeak <- readRDS(
  here("Plots","human_multiome","human_multiome_harmony_merged_malig_peak",
       "markersPeaks_PIMO_up_status.rds")
)

# SimpleList of GRanges result from peakCalling
# This is extracted from above via ArchR::getMarkers()
markersPeak_gr <- readRDS(
  here("Plots","human_multiome","human_multiome_harmony_merged_malig_peak",
       "markersPeaks_GR_PIMO_up_status.rds")
)

# For PIMOup vs PIMOdown (useGroups=PIMOup, bgdGroups = PIMOdown)
# SE
markersTest <- readRDS(
  here("Plots","human_multiome","human_multiome_harmony_merged_malig_peak",
       "PIMOup_vs_PIMOdown",
       "markersTest_PIMO_up_status_PIMOup_vs_PIMOdown.rds")
)
#GR
markersTest_gr <- readRDS(
  here("Plots","human_multiome","human_multiome_harmony_merged_malig_peak",
       "PIMOup_vs_PIMOdown",
       "markersTest_GR_PIMO_up_status_PIMOup_vs_PIMOdown.rds")
)

# For k3_program (non-pairwise)
markersPeak_k3_program <- readRDS(
  here("data","human_multiome","human_multiome_hmmp_k3_program",
       "PeakCalls",
       "markersPeaks_k3_program.rds")
)
```

# Load motifEnrichment Results
Here we load the the output of the motif enrichment script done on the H4h HPC for further EDA and plotting. Specifically, the output is from the script: `run_07_human_multiome_motifEnrichment.sh`, specifically from the ArchR::peakAnnoEnrichment() function.

Each of the results loaded below are derived from different motif sets. A different motif set was assigned to the ArchR object, followed by peakAnnoEnrichment. The default is cisbp, but we also tested homer, which is most closely related to previous in vitro analysis.

```{r load motif enrichment results}
# motifEnrichment

# motifSet = cisbp
motifsUp <- readRDS(file = here("data", "human_multiome", "motifEnrichment", "motifsUp_PIMOup_vs_PIMOdown.rds"))
motifsDown <- readRDS(file = here("data", "human_multiome", "motifEnrichment", "motifsDown_PIMOup_vs_PIMOdown.rds"))
chromVarDev <- readRDS(file = here("data", "human_multiome", "motifEnrichment", "chromVarDeviations_PIMOup_vs_PIMOdown.rds"))

# motifSet = homer (# see below for updated with PIMO70_b)
#motifsUp_homer <- readRDS(file = here("data", "human_multiome", "motifEnrichment_homer", "motifsUp_PIMOup_vs_PIMOdown_homer.rds"))
#motifsDown_homer <- readRDS(file = here("data", "human_multiome", "motifEnrichment_homer", "motifsDown_PIMOup_vs_PIMOdown_homer.rds"))

chromVarDev_homer <- readRDS(file = here("data", "human_multiome", "motifEnrichment_homer", "chromVarDeviations_PIMOup_vs_PIMOdown_homer.rds"))

# check class of objects
class(motifsUp) # SummarizedExperiment
class(chromVarDev) # DFrame

# mouse example:
ms_motifsUp_homer <- readRDS(file = here("data", "mouse_multiome", "motifEnrichment_homer_PIMO_up_status", "motifsUp_PIMOupstatus_homer.rds"))
```

## load updated results (PIMO70_b added)
```{r load updated motif results}
# updated motif results from PIMO_up_status PIMOup_vs_PIMOdown (pairwise)

# motifSet = homer
motifsUp_homer <- readRDS(file = here("data", "human_multiome", "motifEnrichment_homer_PIMO_up_status", "motifsUp_PIMOupstatus_homer.rds"))
motifsDown_homer <- readRDS(file = here("data", "human_multiome", "motifEnrichment_homer_PIMO_up_status", "motifsDown_PIMOupstatus_homer.rds"))

# chromVar deviations
chromVarDev_homer_old <- readRDS(
  file = here(
    "data",
    "human_multiome",
    "motifEnrichment_homer",
    "chromVarDeviations_PIMOup_vs_PIMOdown_homer.rds"
  )
)

chromVarDev_homer <- readRDS(
  file = here(
    "data",
    "human_multiome",
    "chromVarDeviations_homer_PIMO_up_status",
    "2025-12-01",
    "chromVarDeviations_PIMOupstatus_homer.rds"
  )
)

# chromVar Positive-TF-Regulators
chromVarDev_pos_TF_reg <- readRDS(
  file = here(
    "data",
    "human_multiome",
    "chromVarDeviations_homer_PIMO_up_status",
    #"2025-12-01",
    "2025-12-18",
    "Positive-TF-Regulators_PIMOupstatus_homer.rds"
  )
)
```
## load scPOS-MP (k3 program) motifs
```{r load scPos-MP k3 program motifs}

k3_motifsUp <- readRDS(file = here("data", "human_multiome", "human_multiome_hmmp_k3_program",
                                "motifEnrichment_homer_k3_program",
                                "motifsUp_k3program_homer.rds"))

k3_motifsDown <- readRDS(file = here("data", "human_multiome", "human_multiome_hmmp_k3_program",
                                "motifEnrichment_homer_k3_program",
                                "motifsDown_k3program_homer.rds"))
```

# load nomura states (meta_module) motifs
```{r load nomura states (meta_module) motifs}
nomura_motifsUp <- readRDS(
  file = here(
    "data",
    "human_multiome",
    "human_multiome_hmmp_nomura",
    "motifEnrichment_homer_meta_module",
    "motifsUp_metamodule_homer.rds"
  )
)

nomura_motifsDown <- readRDS(
  file = here(
    "data",
    "human_multiome",
    "human_multiome_hmmp_nomura",
    "motifEnrichment_homer_meta_module",
    "motifsDown_metamodule_homer.rds"
  )
)
```

## load custom enrichment with bulk ATAC in multiome
Here we load the result of the custom enrichment analysis where peaks from bulk ATAC-seq were used to compare to multiome results. This was processed by `run_07b_human_multiome_CustomMotifEnrichment.sh`

```{r load custom motif enrichment results}
# load custom enrichment (from ArchR::peakAnnoEnrichment()
# (FDR <= 0.1 & Log2FC >= 0.5)
# bulk_ATAC_DA_peaks
enrichRegions <- readRDS(file = here("data", "human_multiome", "CustomMotifEnrichment", "enrichRegions_customPeaks_bulk_ATAC_DA_peaks_vs_PIMO_up_status.rds"))

class(enrichRegions) # SummarizedExperiment

```

# Exploratory Data Analysis
## SE

```{r eda SE}
#se <- markersPeak
se <- markersTest
# Basic info about the object
dim(se)  # dimensions (Features x samples)
assayNames(se)  # available assay data ("Log2FC"   "Mean"     "FDR"      "Pval"     "MeanDiff" "AUC"      "MeanBGD" )
colnames(se)  # sample names
rownames(se)  # feature

# Metadata exploration
colData(se)  # sample metadata
rowData(se)  # feature metadata
metadata(se)  # experiment metadata

#
rr <- GRanges(rowData(se)$seqnames, IRanges(rowData(se)$start, rowData(se)$end))
rr

# Get the different assays
log2fc <- assay(se, "Log2FC")
mean_expr <- assay(se, "Mean") 
fdr <- assay(se, "FDR")
pval <- assay(se, "Pval")
auc <- assay(se, "AUC")  # Area Under Curve (often from ROC analysis)

# Basic statistics
summary(log2fc)
summary(fdr)

# count NAs in FDR
na_per_comparison <- colSums(is.na(fdr))
print(na_per_comparison)
sum(is.na(fdr)) # 761
```

ArchR has a built in function to draw both volcano and MA plots, and we were interested in reproducing these plots in R from the SE object.

Note, the number of features reported in the MA plot from ArchR, which is 167905, seems to be in conflict with the number of rows or features reported in the GRanges object downstream.

We wondered if this is due to NAs in the data?
We observed 761 NAs in FDR column total among all the groups within PIMO_up_status. Recall, we start with a total of 167905 features unfiltered.
167905 - 761 = 167144

Below is a breakdown for each group of PIMO_up_status:
PIMOdown: 167905 - 35 = 167870
PIMOinter: 167905 - 49 = 167856
PIMOup: 167905 - 677 = 167228
The above features for each group match the final Granges dimensions, suggesting that NAs are filtered out.

Thus, we can conclude the MA plots generated by ArchR show each feature (points) even if their FDRs are NA. The MA plots are made based on the SE object, not the derived GRanges object, which filters these out. These are some of the grey dots labelled as Not-differential (see legend).

## Granges
```{r eda GR}
names(markersPeak_gr)
markersPeak_gr[["PIMOdown"]] %>% length # 167870
markersPeak_gr[["PIMOinter"]] %>% length # 167856
markersPeak_gr[["PIMOup"]] %>% length # 167228


markersPeak_gr[["PIMOup"]] %>% 
  subset(Log2FC > 0 & FDR < 1) %>% length # 66921

# cutoff used previous for MA plot
# "FDR <= 0.1 & abs(Log2FC) >= 0.5"
cutOff <- "FDR <= 0.1 & abs(Log2FC) >= 0.5"

markersPeak_gr[["PIMOup"]] %>%
  subset(FDR <= 0.1 & Log2FC >= 0.5) %>% length # 1268
markersPeak_gr[["PIMOup"]] %>%
  subset(FDR <= 0.1 & Log2FC <= -0.5) %>% length # 2059

# count upregulated and downregulated features based on cutOff
for (i in names(markersPeak_gr)) {
  print(paste("For the PIMO_up_status category:", i))
  print(paste("Using threshold: ", cutOff))
  
  up <- markersPeak_gr[[i]] %>%
    subset(FDR <= 0.1 & Log2FC >= 0.5) %>% length
  down <- markersPeak_gr[[i]] %>%
    subset(FDR <= 0.1 & Log2FC <= -0.5) %>% length
  
  print(paste("The number of upregulated features:",up))
  print(paste("The number of downregulated features:",down))
}

# intersect all three granges
# Assume gr1, gr2, gr3 are your GRanges objects
result <- BiocGenerics::Reduce(GenomicRanges::intersect, list(markersPeak_gr[["PIMOup"]], markersPeak_gr[["PIMOup"]], markersPeak_gr[["PIMOup"]]))

GenomicRanges::intersect(markersPeak_gr[["PIMOup"]],markersPeak_gr[["PIMOup"]])

# count NAs
markersPeak_gr[["PIMOup"]]$FDR %>% is.na %>% table # FALSE 167228
markersPeak_gr[["PIMOup"]]$FDR %>% length() # 167228

  
```
When we intersect the GRanges from the three categories, we end up with 166949 features. When the PIMOup GRanges (167228) is intersected with itself, we end up with 166949 as well, suggesting that some features are lost. It's unclear where the features went, and we also checked for NAs.

# Plot TF Rank vs Significance: PIMOup_vs_PIMOdown motifEnrichment from single cell multiome

To better understand motif enrichment results, we can plot the ranking of each motif and their significance. Here we write two functions: motifsSEtoDF() which converts the SE object to a data frame, and plotMotifEnrichments() that takes the data frame and plots the Significance (-log10padj) vs TF rank. 

```{r plotting function for motif enrichment rank vs Significance }
# convert the SE to df for plotting and EDA
#motifSEtoDF <- function(motifs) {
  # Convert to data frame for ggplot
#   df <- data.frame(TF = rownames(motifs), mlog10Padj = assay(motifs)[, 1])
#   df <- df[order(df$mlog10Padj, decreasing = TRUE), ]
#   df$rank <- seq_len(nrow(df))
#   df$TF_fixed <- sub("_.*", "", df$TF) # remove the _number suffix
# df$TF_family <- sub(".*\\.", "", df$TF_fixed) # add a TF family column based on the last suffix
#   return(df)
# }
motifSEtoDF <- function(motifs) {
  # Extract the underlying matrix from the SummarizedExperiment object.
  # as.matrix() ensures it's a standard dense matrix (this safely handles 
  # underlying sparse dgCMatrix assays from ArchR/Signac)
  motif_mat <- as.matrix(assay(motifs))
  
  # Convert to a data frame and add rownames as a column using tibble
  df_base <- motif_mat %>%
    as.data.frame() %>%
    rownames_to_column("TF")
  
  # Check the matrix columns using ncol()
  if (ncol(motif_mat) > 1) {
    print("More than 1 discrete cell grouping within this groupBy")
    
    # Use tidyr, dplyr, and stringr for pivoting, ranking, and cleaning
    df <- df_base %>%
      pivot_longer(cols = -TF, names_to = "group", values_to = "mlog10Padj") %>%
      group_by(group) %>%
      mutate(rank = row_number(desc(mlog10Padj))) %>%
      ungroup() %>%
      arrange(group, rank) %>%
      mutate(
        TF_fixed = str_remove(TF, "_.*"),
        TF_family = str_remove(TF_fixed, ".*\\.")
      )
      
    return(df)
    
  } else {
    # Single grouping condition
    df <- df_base %>%
      rename(mlog10Padj = 2) %>%
      mutate(rank = row_number(desc(mlog10Padj))) %>%
      arrange(rank) %>%
      mutate(
        TF_fixed = str_remove(TF, "_.*"),
        TF_family = str_remove(TF_fixed, ".*\\.")
      )
      
    return(df)
  }
}

# plot TF rank vs significance
# highlight_TF: provide a vector of TFs to highlight if needed
plotMotifEnrichments <- function(df, highlight_TF = NULL) {
  if (length(highlight_TF) > 0) {
    df$highlight <- ifelse(df$TF_fixed %in% highlight_TF, "highlight", "normal")
    gg <- ggplot(df, aes(rank, mlog10Padj)) +
      geom_point(aes(color = highlight), size = 1.5) +
      scale_color_manual(values = c("highlight" = "red", "normal" = "grey70")) +
      ggrepel::geom_label_repel(
        data = subset(df, TF_fixed %in% highlight_TF),
        aes(x = rank, y = mlog10Padj, label = TF_fixed),
        size = 2.5,
        nudge_x = 150,
        nudge_y = 50,
        hjust = 0.5,
        force = 5,
        direction = "both",
        max.overlaps = 15,
        color = "black"
      ) +
      ylab("-log10(P-adj) Motif Enrichment") +
      xlab("Rank Sorted TFs Enriched")

    return(gg)
    
  }else
    gg <- ggplot(df, aes(rank, mlog10Padj, color = mlog10Padj)) +
      geom_point(size = 1.5) +
      ggrepel::geom_label_repel(
        data = df[rev(seq_len(15)), ],
        aes(x = rank, y = mlog10Padj, label = TF_fixed),
        size = 2.5,
        nudge_x = 150,
        nudge_y = 50,
        hjust = 0.5,
        force = 5,
        direction = "both",
        max.overlaps = 15,
        color = "black"
      ) +
      ylab("-log10(P-adj) Motif Enrichment") +
      xlab("Rank Sorted TFs Enriched")
  
  return(gg)
}
```
## motifSet = cisbp (PIMOup_vs_PIMOdown)
Here we have the cisdb motif results of the enrichment between PIMOup_vs_PIMOdown malignant populations.
```{r plot cisbp}
# SE to use
motifsUp
motifsDown

# convert SE to DF for plotting
motifsUp.df <- motifSEtoDF(motifsUp)
motifsDown.df <- motifSEtoDF(motifsDown)

# plot enriched motifs
ggUp <- plotMotifEnrichments(motifsUp.df)
ggDown <- plotMotifEnrichments(motifsDown.df)

ggUp + ggtitle("Enriched Motifs PIMOup vs PIMOdown", subtitle = "top 15 motifsup cisbp")
ggDown + ggtitle("Enriched Motifs PIMOup vs PIMOdown", subtitle = "top 15 motifsDown cisbp")

# highlight motifs of interest
highlight_TF <- c("HIF1A","ARNT", "SOX2", "SOX3", "SOX4", 
                  "SOX6", "SOX9", "SOX10", "SOX15", "Oct4:SOX17")
plotMotifEnrichments(motifsUp.df, highlight_TF) +
  ggtitle("motifsUp cisbp")
plotMotifEnrichments(motifsDown.df, highlight_TF) +
  ggtitle("motifsDown cisbp")
```
## motifSet = homer (PIMOup_vs_PIMOdown)
```{r plot}

motifsUp_homer.df <- motifSEtoDF(motifsUp_homer)
motifsDown_homer.df <- motifSEtoDF(motifsDown_homer)
#ms_motifsUp.df <- motifSEtoDF(ms_motifsUp_homer)

ggUp_homer <- plotMotifEnrichments(motifsUp_homer.df)
ggDown_homer <- plotMotifEnrichments(motifsDown_homer.df)
#ggup_ms_homer <- plotMotifEnrichments(ms_motifsUp.df)

ggUp_homer + 
  ggtitle("Enriched Motifs PIMOup vs PIMOdown", 
          subtitle = "top 15 motifsUp homer")
ggDown_homer +
  ggtitle("Enriched Motifs PIMOup vs PIMOdown", 
          subtitle = "top 15 motifsDown homer")

# mouse multiome (move later)
#ggup_ms_homer + ggtitle("mouse multiome")

# highlight motifs of interest
highlight_TF <- c("HIF1A","ARNT", "SOX2", "SOX3", "SOX4", 
                  "SOX6", "SOX9", "SOX10", "SOX15", "Oct4:SOX17")
plotMotifEnrichments(motifsUp_homer.df, highlight_TF) +
  ggtitle("motifsUp homer")
plotMotifEnrichments(motifsDown_homer.df, highlight_TF) +
  ggtitle("motifsDown homer")
```

# Table of top motifs

Here we filter the data frames we created earlier for motifs of interest. This is a bit tricky since each motif database names motifs differently in terms of syntax. 

The chromVar DFrame is also converted to data frame for easier access.

```{r top motifs}
# data frames
motifsUp.df
motifsDown.df

# Function: convert chromvar DFrame
motifDRametoDF <- function(DF){
  DF <- as_tibble(as.data.frame(DF))
  DF$TF <- sub("_.*", "", DF$name) # remove the _number suffix # add a TF column
  return(DF)
}
# Convert Chromvar DFrame to dataframe
chromVarDev_homer_old.df <-motifDRametoDF(chromVarDev_homer_old)
chromVarDev_homer.df <- motifDRametoDF(chromVarDev_homer)
chromVarDev.df <- motifDRametoDF(chromVarDev) # cisbp

# Function: check motifs of interest in df
checkMotifsInterest <- function(df, motifs){
df_sub <- df %>%
   filter(str_detect(TF, regex(paste(motifs, collapse = "|"), ignore_case = TRUE)))
return(df_sub) # returns subset of df
}

# motifs of interest to check
# motifs <- c("SOX", "HIF", "ARNT", "NF1", "NFI", "FOS", "JUN","AP-1")
# motifs <- c("ID1","ID2","ID3","ID4")
motifs <- c("HIF",
            "AP","JUN","FOS","FRA",
            "NRF",
            "OLIG")

# check motifs of interest
checkMotifsInterest(motifsUp.df, motifs)
checkMotifsInterest(motifsDown.df, motifs)
checkMotifsInterest(motifsUp_homer.df, motifs)
checkMotifsInterest(motifsDown_homer.df, motifs)

# check motifs of interest: chromVar
checkMotifsInterest(chromVarDev_homer_old.df, motifs)
checkMotifsInterest(chromVarDev_homer.df, motifs)

# join tables
# join motifsup from cisdb and homer: can't join b/c diff naming style
full_join(motifsUp_homer.df, motifsDown_homer.df, by = "TF_fixed") 
```
Note: mlog10Padj means minus log10(adjusted p-value). For Padj < 0.05, we expect the mlog10padj value to be > 1.3.

HIFs are ranked in the bottom 200s/332 from the chromVarDev_homer

# Summarize significant motifs and compare across conditions
```{r motif summary}

motif_summary <- function(df, padj_threshold=0.05){
# Convert p-adj threshold to -log10 scale
  # e.g., 0.05 becomes ~1.301
  mlog10_thresh <- -log10(padj_threshold)
  
  df <- homer_motif_cleaning(df) # clean motifs
  num_motif <- nrow(df) # count motifs

print(paste("The number of motifs in this df is:", num_motif))

num_sig_motif <- df %>%
  filter(mlog10Padj > mlog10_thresh) %>% nrow

print(paste("The number of significant motifs padj<",padj_threshold, "is", num_sig_motif))

# return the significant motifs

sig_motif <- df %>% 
  filter(mlog10Padj > mlog10_thresh) %>%
  pull(TF_cleaned)

return(sig_motif) # returns a vector of the significant motifs

}

# significant motifs for each comparison padj < 0.05

```

# Plot: MA and Volcano

Here we plot the marker peaks using the function implementated from ArchR
```{r plot MA volcano}
plotMarkers_standalone <- function(
  seMarker = NULL,
  name = NULL,
  cutOff = "FDR <= 0.1 & abs(Log2FC) >= 0.5",
  plotAs = "Volcano",
  point_size = 0.5,
  alpha = 0.6,
  rasterize = FALSE
) {
  
  # Input validation
  if (!is(seMarker, "SummarizedExperiment")) {
    stop("seMarker must be a SummarizedExperiment object")
  }
  
  if (is.null(name)) {
    name <- colnames(seMarker)[1]
    message(paste0("No 'name' provided. Using first column: ", name))
  }
  
  if (!name %in% colnames(seMarker)) {
    stop(paste0("'", name, "' not found in seMarker columns. Available columns: ", 
                paste(colnames(seMarker), collapse = ", ")))
  }
  
  if (!plotAs %in% c("Volcano", "MA", "volcano", "ma")) {
    stop("plotAs must be either 'Volcano' or 'MA'")
  }
  
  plotAs <- tolower(plotAs)
  
  # Get available assay names
  assayNames <- names(assays(seMarker))
  
  # Check for required assays
  required_assays <- c("Log2FC", "FDR")
  if (!all(required_assays %in% assayNames)) {
    stop(paste0("Required assays missing. Need: ", paste(required_assays, collapse = ", ")))
  }
  
  # Create environment with assay data for evaluation
  eval_env <- new.env()
  for(an in assayNames) {
    assign(an, assays(seMarker)[[an]], envir = eval_env)
  }
  
  # Evaluate the cutOff expression
  passMat <- eval(parse(text = cutOff), envir = eval_env)
  passMat[is.na(passMat)] <- FALSE
  
  # Extract values for the specified column/group
  FDR <- assays(seMarker)$FDR[, name]
  FDR <- as.numeric(FDR)
  FDR[is.na(FDR)] <- 1
  
  LFC <- assays(seMarker)$Log2FC[, name]
  LFC <- as.numeric(LFC)
  
  # Calculate axis values based on plot type
  if (plotAs == "ma") {
    # Check for required assays for MA plot
    if (!"Mean" %in% assayNames || !"MeanBGD" %in% assayNames) {
      stop("MA plot requires 'Mean' and 'MeanBGD' assays in seMarker")
    }
    
    Mean <- assays(seMarker)$Mean[, name]
    Mean <- as.numeric(Mean)
    MeanBGD <- assays(seMarker)$MeanBGD[, name]
    MeanBGD <- as.numeric(MeanBGD)
    LM <- log2((Mean + MeanBGD)/2 + 1)
    
    x_vals <- LM
    y_vals <- LFC
    x_lab <- "Log2 Mean"
    y_lab <- "Log2 Fold Change"
    
    # Calculate y-axis limits
    qLFC <- max(quantile(abs(LFC), probs = 0.999, na.rm = TRUE), 4) * 1.05
    y_lims <- c(-qLFC, qLFC)
    x_lims <- NULL
    
  } else {  # Volcano plot
    x_vals <- LFC
    y_vals <- -log10(FDR)
    x_lab <- "Log2 Fold Change"
    y_lab <- "-Log10 FDR"
    
    # Calculate x-axis limits
    qLFC <- max(quantile(abs(LFC), probs = 0.999, na.rm = TRUE), 4) * 1.05
    x_lims <- c(-qLFC, qLFC)
    y_lims <- NULL
  }
  
  # Classify features
  color <- ifelse(passMat[, name], "Differential", "Not-Differential")
  color[color == "Differential"] <- ifelse(
    LFC[color == "Differential"] > 0, 
    "Up-Regulated", 
    "Down-Regulated"
  )
  
  # Create plot title with statistics
  n_total <- nrow(seMarker)
  n_up <- sum(color == "Up-Regulated")
  n_down <- sum(color == "Down-Regulated")
  pct_up <- round(100 * n_up / n_total, 2)
  pct_down <- round(100 * n_down / n_total, 2)
  
  plot_title <- paste0(
    "Group: ", name, "\n",
    "Total features: ", n_total, " | ",
    "Up: ", n_up, " (", pct_up, "%) | ",
    "Down: ", n_down, " (", pct_down, "%)"
  )
  
  # Create data frame for plotting (non-differential first, then differential)
  idx <- c(which(color == "Not-Differential"), 
           which(color %in% c("Up-Regulated", "Down-Regulated")))
  
  plot_df <- data.frame(
    x = x_vals[idx],
    y = y_vals[idx],
    color = factor(color[idx], 
                   levels = c("Not-Differential", "Down-Regulated", "Up-Regulated"))
  )
  
  # Define colors
  pal <- c(
    "Up-Regulated" = "#DC143C",      # firebrick3
    "Not-Differential" = "#D3D3D3",   # lightgrey
    "Down-Regulated" = "#1E90FF"      # dodgerblue3
  )
  
  # Create base plot
  p <- ggplot(plot_df, aes(x = x, y = y, color = color)) +
    theme_classic() +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      legend.position = "right",
      plot.title = element_text(hjust = 0.5, size = 11)
    ) +
    labs(
      x = x_lab,
      y = y_lab,
      title = plot_title,
      color = "Status"
    ) +
    scale_color_manual(values = pal)
  
  # Add points (with or without rasterization)
  if (rasterize && requireNamespace("ggrastr", quietly = TRUE)) {
    p <- p + ggrastr::geom_point_rast(size = point_size, alpha = alpha)
  } else {
    if (rasterize) {
      message("ggrastr package not available. Plotting without rasterization.")
    }
    p <- p + geom_point(size = point_size, alpha = alpha)
  }
  
  # Add reference line and adjust axes based on plot type
  if (plotAs == "ma") {
    p <- p + 
      geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
      scale_y_continuous(breaks = seq(-100, 100, 2), 
                        limits = y_lims, 
                        expand = c(0, 0))
  } else {  # Volcano
    p <- p + 
      geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
      scale_x_continuous(breaks = seq(-100, 100, 2), 
                        limits = x_lims, 
                        expand = c(0, 0))
  }
  
  return(p)
}
```

```{r plot MA}
# here we plot for PIMOup
# cutOff = "FDR <= 0.1 & Log2FC >= 0.5"
# "FDR <= 1 & Log2FC >= 0"
# markers[["PIMOup"]] %>% dim() # 75393     7
#markers[["PIMOup"]] %>% nrow # 75393
# sum(markers[["PIMOup"]]$FDR < 0.1)

# For PIMOup vs PIMOdown (useGroups=PIMOup, bgdGroups = PIMOdown)
markersTest
plotMarkers_standalone(seMarker = markersTest, name = "PIMOup", plotAs = "Volcano")
plotMarkers_standalone(seMarker = markersTest, name = "PIMOup", plotAs = "MA")
```
Total features from MA/volcano: 166329
# Custom Motif Enrichment
```{r SE EDA custom motif enrichment}

se_eda <- function(se){
# Basic info about the object
dim(se)  # dimensions (Features x samples)
assayNames(se)  # available assay data ("Log2FC"   "Mean"     "FDR"      "Pval"     "MeanDiff" "AUC"      "MeanBGD" )
colnames(se)  # sample names
rownames(se)  # feature

# Metadata exploration
colData(se)  # sample metadata
rowData(se)  # feature metadata
metadata(se)  # experiment metadata
}
se_eda(enrichRegions)

assay(enrichRegions, "mlog10Padj")
```
## ComplexHeatmap
```{r complexheatmap}
# Row-wise scaling function (recreated from ArchR's .rowScale)
rowScale <- function(mat, min = 0, max = NULL) {
  # Store original rownames and colnames
  rn <- rownames(mat)
  cn <- colnames(mat)
  
  # Calculate row-wise min and max
  row_mins <- apply(mat, 1, min, na.rm = TRUE)
  row_maxs <- apply(mat, 1, max, na.rm = TRUE)
  
  # Scale each row to 0-1 range
  mat_scaled <- t(apply(mat, 1, function(x) {
    row_min <- min(x, na.rm = TRUE)
    row_max <- max(x, na.rm = TRUE)
    
    # Avoid division by zero
    if (row_max == row_min) {
      rep(0, length(x))
    } else {
      (x - row_min) / (row_max - row_min)
    }
  }))
  
  # Apply min/max constraints if specified
  if (!is.null(min)) {
    mat_scaled <- mat_scaled * (1 - min) + min
  }
  
  if (!is.null(max)) {
    mat_scaled <- mat_scaled * max
  }
  
  # Restore names
  rownames(mat_scaled) <- rn
  colnames(mat_scaled) <- cn
  
  # Return list with scaled matrix and original max values
  return(list(
    scaled = mat_scaled,
    max = row_maxs,
    min = row_mins
  ))
}



set.seed(42)
mat <- assay(enrichRegions, "mlog10Padj")
mat <- as.matrix(mat)
scaled_result <- rowScale(mat, min = 0)
mat_scaled <- scaled_result$scaled * 100  # Convert to 0-100 scale

Heatmap(t(mat))
Heatmap((mat_scaled))
```

# Motif Bubble plot 
## From bulk ATAC
```{r load motif bubble plot.r}
# load script with bubble plot functions for motifs
source(here("scripts","makeMotifBubblePlot.R"))
# load manually wrangled dataset containing motifs and other statistics from HOMER analysis

motifs_df <- read_csv(here("data","bulkATAC","2021-10-10_2021-10-08_G523_N_VEH_vs_N_BMP_FDR_HOMER_motifs_bubbleplot.csv"),
                      col_names=TRUE)
```

```{r make bubble plot}
MotifBubblePlot(motifs_df, cond = "knownmotif_N_BMP_edgeR")
# iterate thru each dataset
datasets <- motifs_df %>% pull(dataset) %>% unique()
for (i in datasets){
print(i)
  MotifBubblePlot(motifs_df, cond = i) %>% print()
}

```

# Motif Compare: bulk ATAC-seq vs multiome

First, let's export the homer motifs from the multiome. We will reload this and load the bulk ATAC-seq homer motif results.
```{r export multiome homer motifs, eval = FALSE}
# write_csv(motifsUp_homer.df, here("data","bulkATAC","2025_10-17_human_multiome_PIMOup_vs_PIMOdown_HOMER_motifsUP.csv"),
#                       col_names=TRUE)
# write_csv(motifsDown_homer.df, here("data","bulkATAC","2025_10-17_human_multiome_PIMOup_vs_PIMOdown_HOMER_motifsDown.csv"),
#                       col_names=TRUE)

write_csv(motifsUp_homer.df, here("data","bulkATAC","2025_12-08_human_multiome_PIMOup_vs_PIMOdown_HOMER_motifsUP.csv"),
                      col_names=TRUE)
write_csv(motifsDown_homer.df, here("data","bulkATAC","2025_12-08_human_multiome_PIMOup_vs_PIMOdown_HOMER_motifsDown.csv"),
                      col_names=TRUE)
```

Here we load the bulk atac-seq results from G523, focusing on the comparison of N_BMP vs H_BMP. Motifs enriched in each of those two conditions are identified.

## load bulk ATAC-seq: G523
```{r load GSC G523 bulk ATAC-seq}
# from bulk N_BMP_vs_H_BMP
motifs_df <- read_csv(here("data","bulkATAC","2021-10-09_G523_N_BMP_vs_H_BMP_FDR_HOMER_motifs_bubbleplot.csv"),
                      col_names=TRUE)

motifs_df_H_BMP <- motifs_df %>%
  filter(dataset == "knownmotif_H_BMP_deseq2") # deseq2 was used in thesis

motifs_df_N_BMP <- motifs_df %>%
  filter(dataset == "knownmotif_N_BMP_deseq2") # deseq2 was used in thesis

G523_N_VEH_vs_H_VEH_motifs_df <- read_csv(
here("data","bulkATAC","2021-04-17_G523_N_VEH_vs_H_VEH_FDR_HOMER_motifs_bubbleplot.csv")
)

G523_N_VEH_vs_N_BMP_motifs_df <- read_csv(
here("data","bulkATAC","2021-10-10_2021-10-08_G523_N_VEH_vs_N_BMP_FDR_HOMER_motifs_bubbleplot.csv")
)

# quick EDA
G523_motif_lt <- list(
  G523_N_BMP_vs_H_BMP = motifs_df,
  G523_N_VEH_vs_H_VEH = G523_N_VEH_vs_H_VEH_motifs_df,
  G523_N_VEH_vs_N_BMP = G523_N_VEH_vs_N_BMP_motifs_df
)

lapply(G523_motif_lt, function (x){
  x$TF %>% unique() %>%length()
}
)

# 2. Combine the list into one dataframe
# .id = "condition" automatically creates a column using the list names
combined_motifs <- G523_motif_lt %>%
  bind_rows(.id = "condition") %>%
  mutate(dataset = condition) %>%        # Create dataset column as a copy of condition
  select(dataset, condition, everything()) # Reorder to match CSV structure (dataset, condition first)

# View the result
head(combined_motifs)

# combine list into single table
```
```{r G523, eval = FALSE}
# 1. Define the specific files
G523_files <- c(
  here("data","bulkATAC","2021-10-09_G523_N_BMP_vs_H_BMP_FDR_HOMER_motifs_bubbleplot.csv"),
  here("data","bulkATAC","2021-04-17_G523_N_VEH_vs_H_VEH_FDR_HOMER_motifs_bubbleplot.csv"),
here("data","bulkATAC","2021-10-10_2021-10-08_G523_N_VEH_vs_N_BMP_FDR_HOMER_motifs_bubbleplot.csv")
)

# 2. Define a processing function for these specific CSVs
process_bubbleplot_csv <- function(path) {
  fname <- basename(path)
  
  # Extract metadata from filename
  c_line <- str_extract(fname, "G\\d{3}")
  comp_name <- str_extract(fname, paste0("(?<=", c_line, "_).*(?=_FDR)"))
  
  read_csv(path, show_col_types = FALSE) %>%
    mutate(
      cell_line = c_line,
      # STEP A: Create condition using the original 'dataset' column values
      # We use str_remove on 'dataset' before it is modified
      condition = str_remove(dataset, "^knownmotif_"),
      
      # STEP B: Now overwrite 'dataset' with the comparison name from the filename
      dataset = comp_name,
      
      # Ensure numeric values are handled correctly
      `% of targets` = as.character(`% of targets`),
      `% of background` = as.character(`% of background`)
    ) %>%
    select(
      cell_line, dataset, condition, Rank, TF, Motif, `p-value`, `log pvalue`, 
      `q-value/FDR`, `% of targets`, `% of background`, 
      `Fold Enrichment`, `log2(FE)`
    )
}

# 3. Combine them into a single table
G523_final_table <- G523_files %>%
  map_dfr(process_bubbleplot_csv)

# 4. Save or Inspect
glimpse(G523_final_table)
write_csv(G523_final_table, 
          here("data","bulkATAC","2026-04-20_G523_pairwise_HOMER_motifs.csv"),
                      col_names=TRUE)
# reload
G523_final_table <-
  read_csv(here("data","bulkATAC","2026-04-20_G523_pairwise_HOMER_motifs.csv"),
                      col_names=TRUE)
```

## load bulk ATAC-seq: G361+G411 (n=2, all, normoxia vs hypoxia)
```{r load GSC G361+G411 all normoxia vs hypoxia}
# previously n=1
# 2021-03-07_G361_G411_N_v_H_FDR_HOMER_motifs_bubbleplot.csv

G361_G411_all_motifs_df <- read_csv(here("data",
                                         "bulkATAC",
                                         "2026-01-29_G361_G411_all_normoxia_vs_hypoxia_deseq2_block",
                                         "combined_homer_results.csv"),
                                          col_names=TRUE)

G361_G411_all_motifs_H <- G361_G411_all_motifs_df %>%
  filter(dataset == "hypoxia_enrich") # deseq2 was used in thesis

G361_G411_all_motifs_N <- G361_G411_all_motifs_df %>%
  filter(dataset == "normoxia_enrich") # deseq2 was used in thesis

# 2. clean motifs again so it matches multiome, see G523 for clues how todo it

```
## load bulk ATAC-seq: G361 & G411 (n=2, pairwise)
```{r load GSC G361+G411 pairwise}
# library(tidyverse)
# library(fs)
# 
# # --- 1. Updated Processing Function ---
# # loads the knownResults.txt from folders
# process_homer <- function(path, label) {
#   read_tsv(path, show_col_types = FALSE) %>%
#     mutate(
#       dataset = label,
#       # NEW: Extract the condition by removing the suffix
#       # This looks for "_hypoxia" or "_normoxia" at the end of the string ($)
#       condition = str_remove(dataset, "_(hypoxia|normoxia)$"),
#       
#       Rank = row_number(),
#       TF = str_extract(`Motif Name`, "^[^(/]+"), 
#       Motif = Consensus,
#       `p-value` = `P-value`,
#       `log pvalue` = `Log P-value`,
#       `q-value/FDR` = `q-value (Benjamini)`,
#       `% of targets` = `% of Target Sequences with Motif`,
#       `% of background` = `% of Background Sequences with Motif`,
#       target_num = parse_number(`% of targets`),
#       bg_num = parse_number(`% of background`),
#       `Fold Enrichment` = target_num / bg_num,
#       `log2(FE)` = log2(`Fold Enrichment`)
#     ) %>%
#     select(
#       dataset, condition, Rank, TF, Motif, `p-value`, `log pvalue`, 
#       `q-value/FDR`, `% of targets`, `% of background`, 
#       `Fold Enrichment`, `log2(FE)`
#     )
# }
# 
# # --- 2. Robust Path and Label Generation ---
# base_path <- here("data","bulkATAC")
# 
# # Get folders and filter for the specific date/prefix
# folders <- dir_ls(base_path, type = "directory", regexp = "2023-01-10_G")
# 
# # Create a flat vector of files with clean names
# # Each name will look like: G361_N_72_vs_A_24_hypoxia
# samplefiles <- folders %>%
#   map(function(f) {
#     # Extract the condition name from the folder (removing the date prefix)
#     cond_name <- str_remove(basename(f), "^\\d{4}-\\d{2}-\\d{2}_")
#     
#     # Define the two file paths per folder
#     paths <- c(
#       path(f, "hypoxia_enrich", "knownResults.txt"),
#       path(f, "normoxia_enrich", "knownResults.txt")
#     )
#     
#     # Assign names to these paths for imap to use as the 'label'
#     names(paths) <- c(paste0(cond_name, "_hypoxia_enrich"), paste0(cond_name, "_normoxia_enrich"))
#     return(paths)
#   }) %>%
#   flatten_chr()
# 
# # Filter for existing files only
# samplefiles <- samplefiles[file_exists(samplefiles)]
# 
# # --- 3. Run and Combine ---
# final_table <- imap_dfr(samplefiles, ~process_homer(.x, .y))
# 
# # Check the first few rows
# head(final_table)
library(tidyverse)
library(fs)

# --- 1. Redefined Processing Function ---
process_homer <- function(path, label) {
  read_tsv(path, show_col_types = FALSE) %>%
    mutate(
      # 1. cell_line: Everything before the first underscore (e.g., G361)
      cell_line = str_extract(label, "^[^_]+"),
      
      # 2. condition: The enrichment suffix (e.g., hypoxia_enrich)
      condition = str_extract(label, "(hypoxia|normoxia)_enrich$"),
      
      # 3. dataset: The comparison part (e.g., G361_N_72_vs_A_24)
      dataset = str_remove(label, "_(hypoxia|normoxia)_enrich$"),
      
      Rank = row_number(),
      TF = str_extract(`Motif Name`, "^[^(/]+"), 
      Motif = Consensus,
      `p-value` = `P-value`,
      `log pvalue` = `Log P-value`,
      `q-value/FDR` = `q-value (Benjamini)`,
      `% of targets` = `% of Target Sequences with Motif`,
      `% of background` = `% of Background Sequences with Motif`,
      
      target_num = parse_number(`% of targets`),
      bg_num = parse_number(`% of background`),
      `Fold Enrichment` = target_num / bg_num,
      `log2(FE)` = log2(`Fold Enrichment`)
    ) %>%
    # Reordered columns to start with your new metadata
    select(
      cell_line, dataset, condition, Rank, TF, Motif, `p-value`, `log pvalue`, 
      `q-value/FDR`, `% of targets`, `% of background`, 
      `Fold Enrichment`, `log2(FE)`
    )
}

# --- 2. Updated File Path and Label Generation ---
base_path <- "C:/Users/wuronald/WKLab Dropbox/Ronald Wu/_wouters lab/Data Ronald/scATAC-seq/data/bulkATAC"
folders <- dir_ls(base_path, type = "directory", regexp = "2023-01-10_G")

samplefiles <- folders %>%
  map(function(f) {
    # Remove date prefix to get the base comparison name (e.g., G361_N_72_vs_A_24)
    comparison_name <- str_remove(basename(f), "^\\d{4}-\\d{2}-\\d{2}_")
    
    paths <- c(
      path(f, "hypoxia_enrich", "knownResults.txt"),
      path(f, "normoxia_enrich", "knownResults.txt")
    )
    
    # Create labels that match your new decomposition logic
    names(paths) <- c(
      paste0(comparison_name, "_hypoxia_enrich"), 
      paste0(comparison_name, "_normoxia_enrich")
    )
    return(paths)
  }) %>%
  flatten_chr()

# Safety filter for existing files
samplefiles <- samplefiles[file_exists(samplefiles)]

# --- 3. Run and Combine ---
final_table <- imap_dfr(samplefiles, ~process_homer(.x, .y))

# View a summary of the new column structure
glimpse(final_table)
```
### export G361/G411 pairwise homer motif csv
```{r export G361/G411 pairwise homer motif csv, eval=FALSE}

write_csv(final_table, 
          here("data","bulkATAC","2026-04-19_G361_G411_RW492_RW555_pairwise_HOMER_motifs.csv"),
                      col_names=TRUE)

# reload
final_table <- read_csv( 
          here("data","bulkATAC","2026-04-19_G361_G411_RW492_RW555_pairwise_HOMER_motifs.csv"),
                      col_names=TRUE)
```
## HOMER motif cleaning and pre-processing
```{r motif cleaning homer}

motifsUp_homer.df<- read_csv(here("data","bulkATAC","2025_12-08_human_multiome_PIMOup_vs_PIMOdown_HOMER_motifsUP.csv"),
                      col_names=TRUE)

# newest version of homer_motif_cleaning
homer_motif_cleaning_v2 <- function(df){
  df_cleaned <- df %>%
  mutate(
    TF_cleaned = TF %>%
      # Step 1: Remove the numerical suffix (e.g., _251)
      # "AP.1.bZIP_1" -> "AP.1.bZIP"
      str_remove("_\\d+$") %>%
      
      # Step 2: Handle the specific edge cases first.
      # "Brachyury.T.box" -> "Brachyury"
      # "Brn2-POU" -> "Brn2"
      str_remove("\\.T\\.box$") %>%
      str_remove("-Homeobox$") %>%
      
      # Step 3: Now, handle the general case by removing the last remaining 
      # motif identifier (e.g., .bZIP, .DR1).
      # "AP.1.bZIP" -> "AP.1"
      # "RXR.NR..DR1" -> "RXR.NR.."
      str_remove("\\.[^.]+$") %>%
      
      # Step 4: Replace all remaining periods with hyphens.
      # "AP.1" -> "AP-1"
      # "RXR.NR.." -> "RXR-NR--"
      str_replace_all("\\.", "-") %>%
      
      # Step 5: Final cleanup. Remove any trailing hyphens.
      # "RXR-NR--" -> "RXR-NR"
      str_remove("-+$") %>%
    
      # Step 6: Additional edge cases:
      str_remove("-Paired$") %>%
      str_remove("-NR$") %>%
      str_remove("-Forkhead$") %>%
      str_remove("-Homeobox$") %>%
      str_remove("-POU$") %>%
      str_remove("-Ebox$") %>%
      str_remove("-Zf$") %>%
      str_remove("-EBV$") %>%
      str_remove("-ETS$") %>%
      str_replace_all("NFkB-p50-p52","NFkB-p50,p52") %>%
      str_replace_all("CEBP-AP1","CEBP:AP1") %>%
      str_replace_all("CEBP-CEBP","CEBP:CEBP") %>%
      str_replace_all("ETS-E-box","ETS:E-box") %>%
      str_replace_all("ETS-RUNX","ETS:RUNX") %>%
      str_replace_all("EWS-ERG-fusion","EWS:ERG-fusion") %>%
      str_replace_all("FOXA1-AR","FOXA1:AR") %>%
      str_replace_all("GATA-SCL","GATA:SCL") %>%
      str_replace_all("IRF-BATF-IRF","IRF:BATF") %>%
      str_replace_all("NF1-FOXA1-CTF","NF1:FOXA1") %>%
      str_replace_all("NFAT-AP1-RHD","NFAT:AP1") %>%
      str_replace_all("Oct4-Sox17","Oct4:Sox17") %>%
      str_replace_all("PAX3-FKHR-fusion","PAX3:FKHR-fusion") %>%
      str_replace_all("FOXA1-AR","FOXA1:AR") %>%
      str_replace_all("\\bFox\\b","Fox:Ebox") %>%
      str_replace_all("PU-1","PU.1") %>%
      str_replace_all("PU-1-IRF-ETS","PU.1-IRF") %>%
      str_replace_all("PU.1-IRF8","PU.1:IRF8") %>%
      str_replace_all("RBPJ","RBPJ:Ebox") %>%
      str_replace_all("Stat3-il21","Stat3+il21") %>%
      str_replace_all("Tbox-Smad-T-box","Tbox:Smad") %>%
    
      
      
      str_replace_all("Arnt-Ahr","Arnt:Ahr") %>%
      str_replace_all("bZIP-IRF-bZIP","bZIP:IRF") %>%
      str_replace_all("CTCF-SatelliteElement-Zf", "CTCF-SatelliteElement") %>%
      str_replace_all("EWS-FLI1-fusion","EWS:FLI1-fusion") %>% 
      str_replace_all("Fox-Ebox", "Fox:Ebox") %>%
      str_replace_all("GRE-NR","GRE") %>%
      str_replace_all("Pit1-1bp","Pit1+1bp") %>%
      str_replace_all("Tlx","Tlx?") %>%
      str_replace_all("OCT-OCT","OCT:OCT") %>%
      str_replace_all("RAR-RXR","RAR:RXR") %>%
      str_replace_all("ZNF143-STAF","ZNF143|STAF") %>%
      str_replace_all("OCT4-SOX2-TCF-NANOG-POU-Homeobox","OCT4-SOX2-TCF-NANOG") %>%
      str_replace_all("(Nkx\\d+)-(\\d+)", "\\1.\\2") # for Nkxn-n, where n is any number
  )
return (df_cleaned)    
}

motifsDown_homer.df_cleaned_temp<-homer_motif_cleaning_v2(motifsUp_homer.df)

# old cleaning
setdiff(motifsDown_homer.df_cleaned$TF_cleaned %>% unique(),
G361_G411_G523_final_table$TF %>% unique()) %>% sort()
# new cleaning
setdiff(motifsDown_homer.df_cleaned_temp$TF_cleaned %>% unique(),
G361_G411_G523_final_table$TF %>% unique()) %>% sort

 # what is in the GSC only
setdiff(G361_G411_G523_final_table$TF %>% unique(),
        motifsDown_homer.df_cleaned$TF_cleaned %>% unique()) %>% sort()

setdiff(G361_G411_G523_final_table$TF %>% unique(),
        motifsDown_homer.df_cleaned_temp$TF_cleaned %>% unique()) %>% sort()
```

The following 16 motifs after version 2 cleaning are unique to human multiome HOMER (absent in GSC HOMER; not even related motifs in syntax)
1. CDX4
2. COUP-TFII
3. Cux2
4. E2A-bHLH--near_PU (Note: E2A alone is present in both, but this one is special)
5. FOXK2
6. Foxo3
7. Fra2 (Fra1 exists in GSC though)
8. Gata6 (other GATAs are available, but not this one)
9. KLF3 (other Klfs available but not this specific one)
10. NPAS (NPAS2 available not NPAS)
11. PAX5-Paired (Pax5 already has a match, this one is condensed version)
12. Pax7-Paired (Pax7 already has a match, this one is longest)
13. Ronin
14. SF1
15. Six2 (Six1 available but not Six2)
16. THRb


## Define motif compare plotting functions
```{r motif compare plotting functions}
# from human multiome
motifsUp_homer.df
motifsDown_homer.df

# Function for additional cleaning of motifs (homer multiome syntax needs to match homer bulk-ATAC-seq)
homer_motif_cleaning <- function(df){
  df_cleaned <- df %>%
  mutate(
    TF_cleaned = TF %>%
      # Step 1: Remove the numerical suffix (e.g., _251)
      # "AP.1.bZIP_1" -> "AP.1.bZIP"
      str_remove("_\\d+$") %>%
      
      # Step 2: Handle the specific edge cases first.
      # "Brachyury.T.box" -> "Brachyury"
      # "Brn2-POU" -> "Brn2"
      str_remove("\\.T\\.box$") %>%
      str_remove("-Homeobox$") %>%
      
      # Step 3: Now, handle the general case by removing the last remaining 
      # motif identifier (e.g., .bZIP, .DR1).
      # "AP.1.bZIP" -> "AP.1"
      # "RXR.NR..DR1" -> "RXR.NR.."
      str_remove("\\.[^.]+$") %>%
      
      # Step 4: Replace all remaining periods with hyphens.
      # "AP.1" -> "AP-1"
      # "RXR.NR.." -> "RXR-NR--"
      str_replace_all("\\.", "-") %>%
      
      # Step 5: Final cleanup. Remove any trailing hyphens.
      # "RXR-NR--" -> "RXR-NR"
      str_remove("-+$") %>%
    
      # Step 6: Additional edge cases:
      str_remove("-Paired$") %>%
      str_remove("-NR$") %>%
      str_remove("-Forkhead$") %>%
      str_remove("-Homeobox$") %>%
      str_remove("-POU$") %>%
      str_replace_all("Arnt-Ahr","Arnt:Ahr") %>%
      str_replace_all("bZIP-IRF-bZIP","bZIP:IRF") %>%
      str_replace_all("CTCF-SatelliteElement-Zf", "CTCF-SatelliteElement") %>%
      str_replace_all("EWS-FLI1-fusion","EWS:FLI1-fusion") %>% 
      str_replace_all("Fox-Ebox", "Fox:Ebox") %>%
      str_replace_all("GRE-NR","GRE") %>%
      str_replace_all("Pit1-1bp","Pit1+1bp") %>%
      str_replace_all("Tlx","Tlx?") %>%
      str_replace_all("OCT-OCT","OCT:OCT") %>%
      str_replace_all("RAR-RXR","RAR:RXR") %>%
      str_replace_all("ZNF143-STAF","ZNF143|STAF") %>%
      str_replace_all("OCT4-SOX2-TCF-NANOG-POU-Homeobox","OCT4-SOX2-TCF-NANOG") %>%
      str_replace_all("(Nkx\\d+)-(\\d+)", "\\1.\\2") # for Nkxn-n, where n is any number
  )
return (df_cleaned)    
}

df1_cleaned <- homer_motif_cleaning(motifsUp_homer.df)
df2_cleaned <- motifs_df_H_BMP %>%
  mutate(TF_cleaned = str_extract(TF, "[^|]+"))
df3_cleaned <- motifs_df_N_BMP %>%
  mutate(TF_cleaned = str_extract(TF, "[^|]+"))
df4_cleaned <- homer_motif_cleaning(motifsDown_homer.df)

# human multiome motifsUP vs G523 motifs_H_BMP enriched
common_tfs_1 <- intersect(df1_cleaned$TF_cleaned, df2_cleaned$TF_cleaned)
# human multiome motifsUP vs G523 motifs_N_BMP enriched
common_tfs_2 <- intersect(df1_cleaned$TF_cleaned, df3_cleaned$TF_cleaned)

# human multiome motifsDown vs G523 motifs_H_BMP enriched
common_tfs_3 <- intersect(df4_cleaned$TF_cleaned, df2_cleaned$TF_cleaned)
# human multiome motifsDown vs G523 motifs_N_BMP enriched
common_tfs_4 <- intersect(df4_cleaned$TF_cleaned, df3_cleaned$TF_cleaned)

# sanity check for mismatched/ambiguous motifs between datasets
motifs_df$TF[!motifs_df$TF %in% df1_cleaned$TF_cleaned] %>% sort

# Updated plotting function
# show_labels_qty shows the top 25 motifs (default), based on the rank column
plotMotifEnrichments_new <- function(df, common_tfs_vector, show_labels = TRUE, show_labels_qty = 25) {
  padj_threshold = 0.05 # threshold for y-axis horizontal line
  # 1. Create a new column to flag common TFs
  df <- df %>%
    mutate(is_common = TF_cleaned %in% common_tfs_vector)
  
  num_is_common <- df %>% pull(is_common) %>% sum(na.rm = TRUE)
  print(paste0("The number of motis in common is:", num_is_common, " from a total of ",nrow(df)," motifs"))
  
  if(num_is_common > show_labels_qty){
    print(paste0("Number of common motifs exceeds the number to requested to label:", show_labels_qty))
  }
  
  # 2. Create the base plot
  gg <- ggplot(df, aes(x = rank, y = mlog10Padj)) + 
    geom_point(aes(color = is_common, shape = is_common), size = 2, alpha = 0.8) +
    scale_color_manual(name = "Common TF", values = c(`TRUE` = "red", `FALSE` = "grey50")) +
    scale_shape_manual(name = "Common TF", values = c(`TRUE` = 16, `FALSE` = 16)) +
    ylab("-log10(P-adj) Motif Enrichment") + 
    xlab("Rank Sorted TFs Enriched") +
    theme_minimal() +
    guides(color = guide_legend(override.aes = list(size = 4)))
  
  # 3. Conditionally add labels
  # This 'if' statement checks if show_labels is TRUE
  
  if (show_labels & length(common_tfs_vector) > 0) {
    ylimit <- round(max(df$mlog10Padj))
    
    gg <- gg + geom_label_repel(
      data = filter(df, is_common) %>% slice_head(n = show_labels_qty),
      aes(label = TF_cleaned),
      size = 3,
      force = 10,
      max.overlaps = Inf,
      color = "black",
      segment.color = "grey50",
      xlim = c(100, Inf),
      ylim = c(50, ylimit),
      nudge_x = 5,
      nudge_y = 10
    )
  }else{
    ylimit <- round(max(df$mlog10Padj))
    
    gg <- gg + geom_label_repel(
      data = filter(df) %>% slice_head(n = show_labels_qty),
      aes(label = TF_cleaned),
      size = 3,
      force = 10,
      max.overlaps = Inf,
      color = "black",
      segment.color = "grey50",
      xlim = c(100, Inf),
      ylim = c(50, ylimit),
      nudge_x = 5,
      nudge_y = 10
    )
  }
  
  return(gg)
}
```
## Define final plotting function
Here we improve on the plotting function, building upon plotMotifEnrichments(), and plotMotifEnrichments_new().

```{r plotMotifEnrichments_final}
plotMotifEnrichments_final <- function(df,
                                       common_tfs_vector = NULL,
                                       show_labels = TRUE,
                                       show_labels_qty = 25,
                                       padj_threshold = 0.05) {
  
  # Convert p-adj threshold to -log10 scale
  mlog10_thresh <- -log10(padj_threshold)

  # 1. Flag common TFs and handle NULL vector
  if(is.null(common_tfs_vector)) common_tfs_vector <- character(0)
  df <- df %>% mutate(is_common = TF_cleaned %in% common_tfs_vector)
  
  # 2. Base Plot
  gg <- ggplot(df, aes(x = rank, y = mlog10Padj)) + 
    geom_hline(yintercept = mlog10_thresh, linetype = "dashed", color = "darkred", alpha = 0.4) +
    geom_point(aes(color = is_common), shape = 16, size = 2, alpha = 0.7) +
    scale_color_manual(
      name = "Common TF", 
      values = c(`TRUE` = "red", `FALSE` = "grey60"),
      labels = c(`TRUE` = "Yes", `FALSE` = "No")
    ) +
    scale_x_continuous(expand = expansion(mult = c(0.05, 0.25))) + # Extra space for labels
    labs(y = "-log10(P-adj) Enrichment", x = "Rank Sorted TFs") +
    theme_minimal() +
    theme(panel.grid.minor = element_blank())

  # 3. Text Labeling (Shifted Right)
  if (show_labels) {
    # Only label those passing threshold
    significant_df <- df %>% filter(mlog10Padj >= mlog10_thresh)
    
    label_data <- if(any(significant_df$is_common)) {
      significant_df %>% filter(is_common) %>% slice_head(n = show_labels_qty)
    } else {
      significant_df %>% slice_head(n = show_labels_qty)
    }
    
    if(nrow(label_data) > 0) {
      gg <- gg + geom_text_repel(
        data = label_data,
        aes(label = TF_cleaned),
        size = 3.2,
        color = "black",
        # Push to the right sidebar
        force = 5,
        nudge_x = max(df$rank, na.rm = TRUE) * 0.15,
        direction = "y",
        hjust = 0,
        max.overlaps = 30,
        box.padding = 0.3,
        point.padding = 0.5,
        min.segment.length = 0,
        segment.color = "grey70"
      )
    }
  }
  return(gg)
}

```

```{r plotting}
gg_compare1 <-plotMotifEnrichments(df1_cleaned) # [1] 332   6
gg_compare2 <-plotMotifEnrichments_new(df1_cleaned, common_tfs_1)
gg_compare3 <-plotMotifEnrichments_new(df1_cleaned, common_tfs_1,FALSE)
gg_compare4 <-plotMotifEnrichments_new(df1_cleaned, common_tfs_2)
gg_compare5 <-plotMotifEnrichments_new(df1_cleaned, common_tfs_2,FALSE)

gg_compare6 <-plotMotifEnrichments(df4_cleaned) # [1] 332   6
gg_compare7 <-plotMotifEnrichments_new(df4_cleaned, common_tfs_3)
gg_compare8 <-plotMotifEnrichments_new(df4_cleaned, common_tfs_3,FALSE)
gg_compare9 <-plotMotifEnrichments_new(df4_cleaned, common_tfs_4)
gg_compare10 <-plotMotifEnrichments_new(df4_cleaned, common_tfs_4,FALSE)

# MotifsUP plots
gg_compare1 + ggtitle("Human Multiome: MotifsUP in PIMOup vs PIMOdown")
gg_compare2 + ggtitle("Human Multiome: MotifsUP in PIMOup vs PIMOdown", subtitle = "H_BMP highlighted from G523 H_BMP_vs_N_BMP")
gg_compare3 + ggtitle("Human Multiome: MotifsUP in PIMOup vs PIMOdown", subtitle = "H_BMP highlighted from G523 H_BMP_vs_N_BMP")
gg_compare4 + ggtitle("Human Multiome: MotifsUP in PIMOup vs PIMOdown", subtitle = "N_BMP highlighted from G523 H_BMP_vs_N_BMP")
gg_compare5 + ggtitle("Human Multiome: MotifsUP in PIMOup vs PIMOdown", subtitle = "N_BMP highlighted from G523 H_BMP_vs_N_BMP")

# MotifsDown Plots
gg_compare6 + ggtitle("Human Multiome: MotifsDown in PIMOup vs PIMOdown")
gg_compare7 + ggtitle("Human Multiome: MotifsDown in PIMOup vs PIMOdown", subtitle = "H_BMP highlighted from G523 H_BMP_vs_N_BMP")
gg_compare8 + ggtitle("Human Multiome: MotifsDown in PIMOup vs PIMOdown", subtitle = "H_BMP highlighted from G523 H_BMP_vs_N_BMP")
gg_compare9 + ggtitle("Human Multiome: MotifsDown in PIMOup vs PIMOdown", subtitle = "N_BMP highlighted from G523 H_BMP_vs_N_BMP")
gg_compare10 + ggtitle("Human Multiome: MotifsDown in PIMOup vs PIMOdown", subtitle = "N_BMP highlighted from G523 H_BMP_vs_N_BMP")


plotMotifEnrichments_new(df1_cleaned, NULL)
```
## Export PDFs
```{r export multiome vs bulk-ATAC motif comparisons, eval = FALSE}
# pdf(here("Plots","2025-10-17_human_multiome_PIMOup_vs_PIMOdown_motifUP_vs_G523_H_BMP_vs_N_BMP_motif_ranking_bulk.pdf"))

pdf(here("Plots","2025-12-08_human_multiome_PIMOup_vs_PIMOdown_motifUp_vs_G523_H_BMP_vs_N_BMP_motif_ranking_bulk.pdf"))
plotMotifEnrichments_new(df1_cleaned, NULL) + ggtitle("Human Multiome: MotifsUP in PIMOup vs PIMOdown", subtitle = "Top 25 Motifs")
gg_compare1 + ggtitle("Human Multiome: MotifsUP in PIMOup vs PIMOdown")
gg_compare2 + ggtitle("Human Multiome: MotifsUP in PIMOup vs PIMOdown", subtitle = "H_BMP highlighted from G523 H_BMP_vs_N_BMP")
gg_compare3 + ggtitle("Human Multiome: MotifsUP in PIMOup vs PIMOdown", subtitle = "H_BMP highlighted from G523 H_BMP_vs_N_BMP")
gg_compare4 + ggtitle("Human Multiome: MotifsUP in PIMOup vs PIMOdown", subtitle = "N_BMP highlighted from G523 H_BMP_vs_N_BMP")
gg_compare5 + ggtitle("Human Multiome: MotifsUP in PIMOup vs PIMOdown", subtitle = "N_BMP highlighted from G523 H_BMP_vs_N_BMP")
dev.off()

# pdf(here("Plots","2025-10-17_human_multiome_PIMOup_vs_PIMOdown_motifDown_vs_G523_H_BMP_vs_N_BMP_motif_ranking_bulk.pdf"))
pdf(here("Plots","2025-12-08_human_multiome_PIMOup_vs_PIMOdown_motifDown_vs_G523_H_BMP_vs_N_BMP_motif_ranking_bulk.pdf"))
plotMotifEnrichments_new(df4_cleaned, NULL) + ggtitle("Human Multiome: MotifsDown in PIMOup vs PIMOdown", subtitle = "Top 25 Motifs")
gg_compare6 + ggtitle("Human Multiome: MotifsDown in PIMOup vs PIMOdown")
gg_compare7 + ggtitle("Human Multiome: MotifsDown in PIMOup vs PIMOdown", subtitle = "H_BMP highlighted from G523 H_BMP_vs_N_BMP")
gg_compare8 + ggtitle("Human Multiome: MotifsDown in PIMOup vs PIMOdown", subtitle = "H_BMP highlighted from G523 H_BMP_vs_N_BMP")
gg_compare9 + ggtitle("Human Multiome: MotifsDown in PIMOup vs PIMOdown", subtitle = "N_BMP highlighted from G523 H_BMP_vs_N_BMP")
gg_compare10 + ggtitle("Human Multiome: MotifsDown in PIMOup vs PIMOdown", subtitle = "N_BMP highlighted from G523 H_BMP_vs_N_BMP")
dev.off()

```
## Automated Motif Compare Report

```{r motif compare report}
# human multiome
motifsUp_homer.df
motifsDown_homer.df
# Load motif bubble plot.csv
motifs_df <- read_csv(here("data","bulkATAC","2021-10-09_G523_N_BMP_vs_H_BMP_FDR_HOMER_motifs_bubbleplot.csv"),
                      col_names=TRUE)
motifs_df2 <- read_csv(here("data","bulkATAC","2021-04-17_G523_N_VEH_vs_H_VEH_FDR_HOMER_motifs_bubbleplot.csv"),
                      col_names=TRUE)
motifs_df3 <- read_csv(here("data","bulkATAC","2021-10-10_2021-10-08_G523_N_VEH_vs_N_BMP_FDR_HOMER_motifs_bubbleplot.csv"),
                      col_names=TRUE)

motifs_df_H_BMP <- motifs_df %>%
  filter(dataset == "knownmotif_H_BMP_deseq2") # deseq2 was used in thesis

motifs_df_N_BMP <- motifs_df %>%
  filter(dataset == "knownmotif_N_BMP_deseq2") # deseq2 was used in thesis

motifs_df2_H_VEH <- motifs_df2 %>%
  filter(dataset == "knownmotif_H_VEH_deseq2") # deseq2 was used in thesis

motifs_df2_N_VEH <- motifs_df2 %>%
  filter(dataset == "knownmotif_N_VEH_deseq2") # deseq2 was used in thesis

motifs_df3_N_BMP <- motifs_df3 %>%
  filter(dataset == "knownmotif_N_BMP_deseq2") # deseq2 was used in thesis

motifs_df3_N_VEH <- motifs_df3 %>%
  filter(dataset == "knownmotif_N_VEH_deseq2") # deseq2 was used in thesis

# Function: motifCompareReport
# This function compares the motifs from multiome with those enriched in two other
# datasets, such as from bulk ATAC-seq from GSCs
# multiome_df dataframe with human multiome homer motif results 
# gsc_motif_df_1 dataframe from GSC # 1
# gsc_motif_df_2 dataframe from GSC # 2
motifCompareReport <- function(multiome_motif_df, gsc_motif_df_1, gsc_motif_df_2,
                               gsc_title1="", gsc_title2="", title="Ranked Motif Plot", subtitle=""){
  # clean multiome motifs
  multiome_motif_df <- homer_motif_cleaning(multiome_motif_df)
  # clean GSC motifs
  gsc_motif_df_1 <- gsc_motif_df_1 %>% mutate(TF_cleaned = str_extract(TF, "[^|]+"))
  gsc_motif_df_2 <- gsc_motif_df_2 %>% mutate(TF_cleaned = str_extract(TF, "[^|]+"))
  
  # determine motifs in common for each GSC comparison
  common_tfs_1 <- intersect(multiome_motif_df$TF_cleaned, gsc_motif_df_1$TF_cleaned)
  common_tfs_2 <- intersect(multiome_motif_df$TF_cleaned, gsc_motif_df_2$TF_cleaned)
  
  # Make plots
  # 1. No labels
  p1 <- plotMotifEnrichments_new(multiome_motif_df, common_tfs_1, show_labels = FALSE)
  p2 <- plotMotifEnrichments_new(multiome_motif_df, common_tfs_2, show_labels = FALSE)
  # 2. Show labels (default)
  p3 <- plotMotifEnrichments_new(multiome_motif_df, common_tfs_1, show_labels = TRUE)
  p4 <- plotMotifEnrichments_new(multiome_motif_df, common_tfs_2, show_labels = TRUE)
  # 3. custom label amount
  return(list(
    p1 + ggtitle(title, subtitle = paste(gsc_title1, subtitle)),
    p2 + ggtitle(title, subtitle = paste(gsc_title2, subtitle)),
    p3 + ggtitle(title, subtitle = paste(gsc_title1, subtitle)),
    p4 + ggtitle(title, subtitle = paste(gsc_title2, subtitle))
  ))
}

# Run reports function
# 1. H_BMP_vs_N_BMP_deseq2
motifCompareReport(motifsUp_homer.df, motifs_df_H_BMP, motifs_df_N_BMP, gsc_title1 = "H_BMP", gsc_title2 = "N_BMP",
                   title="Human Multiome: MotifsUP in PIMOup vs PIMOdown", 
                   subtitle="highlighted from G523 H_BMP_vs_N_BMP_deseq2")
motifCompareReport(motifsDown_homer.df, motifs_df_H_BMP, motifs_df_N_BMP, gsc_title1 = "H_BMP", gsc_title2 = "N_BMP",
                   title="Human Multiome: MotifsDown in PIMOup vs PIMOdown", 
                   subtitle="highlighted from G523 H_BMP_vs_N_BMP_deseq2")
# 2. N_VEH_vs_H_VEH_deseq2
motifCompareReport(motifsUp_homer.df, motifs_df2_H_VEH, motifs_df2_N_VEH, gsc_title1 = "H_VEH", gsc_title2 = "N_VEH",
                   title="Human Multiome: MotifsUP in PIMOup vs PIMOdown", 
                   subtitle="highlighted from G523 H_VEH_vs_N_VEH_deseq2")
motifCompareReport(motifsDown_homer.df, motifs_df2_H_VEH, motifs_df2_N_VEH, gsc_title1 = "H_VEH", gsc_title2 = "N_VEH",
                   title="Human Multiome: MotifsDown in PIMOup vs PIMOdown", 
                   subtitle="highlighted from G523 H_VEH_vs_N_VEH_deseq2")
# 3. N_VEH_vs_N_BMP_deseq2
motifCompareReport(motifsUp_homer.df, motifs_df3_N_BMP, motifs_df3_N_VEH, gsc_title1 = "N_BMP", gsc_title2 = "N_VEH",
                   title="Human Multiome: MotifsUP in PIMOup vs PIMOdown", 
                   subtitle="highlighted from G523 N_BMP_vs_N_VEH_deseq2")
motifCompareReport(motifsDown_homer.df, motifs_df3_N_BMP, motifs_df3_N_VEH, gsc_title1 = "N_BMP", gsc_title2 = "N_VEH",
                   title="Human Multiome: MotifsDown in PIMOup vs PIMOdown", 
                   subtitle="highlighted from G523 N_BMP_vs_N_VEH_deseq2")
```

### export reports 

```{r export reports, eval = FALSE}
#pdf(here("Plots","2025-11-14_human_multiome_PIMOup_vs_PIMOdown_motifs_vs_G523_motifs_motif_ranking_bulk.pdf"))
pdf(here("Plots","2025-12-08_human_multiome_PIMOup_vs_PIMOdown_motifs_vs_G523_motifs_motif_ranking_bulk.pdf"))
# 1. H_BMP_vs_N_BMP_deseq2
motifCompareReport(motifsUp_homer.df, motifs_df_H_BMP, motifs_df_N_BMP, gsc_title1 = "H_BMP", gsc_title2 = "N_BMP",
                   title="Human Multiome: MotifsUP in PIMOup vs PIMOdown", 
                   subtitle="highlighted from G523 H_BMP_vs_N_BMP_deseq2")
motifCompareReport(motifsDown_homer.df, motifs_df_H_BMP, motifs_df_N_BMP, gsc_title1 = "H_BMP", gsc_title2 = "N_BMP",
                   title="Human Multiome: MotifsDown in PIMOup vs PIMOdown", 
                   subtitle="highlighted from G523 H_BMP_vs_N_BMP_deseq2")
# 2. N_VEH_vs_H_VEH_deseq2
motifCompareReport(motifsUp_homer.df, motifs_df2_H_VEH, motifs_df2_N_VEH, gsc_title1 = "H_VEH", gsc_title2 = "N_VEH",
                   title="Human Multiome: MotifsUP in PIMOup vs PIMOdown", 
                   subtitle="highlighted from G523 H_VEH_vs_N_VEH_deseq2")
motifCompareReport(motifsDown_homer.df, motifs_df2_H_VEH, motifs_df2_N_VEH, gsc_title1 = "H_VEH", gsc_title2 = "N_VEH",
                   title="Human Multiome: MotifsDown in PIMOup vs PIMOdown", 
                   subtitle="highlighted from G523 H_VEH_vs_N_VEH_deseq2")
# 3. N_VEH_vs_N_BMP_deseq2
motifCompareReport(motifsUp_homer.df, motifs_df3_N_BMP, motifs_df3_N_VEH, gsc_title1 = "N_BMP", gsc_title2 = "N_VEH",
                   title="Human Multiome: MotifsUP in PIMOup vs PIMOdown", 
                   subtitle="highlighted from G523 N_BMP_vs_N_VEH_deseq2")
motifCompareReport(motifsDown_homer.df, motifs_df3_N_BMP, motifs_df3_N_VEH, gsc_title1 = "N_BMP", gsc_title2 = "N_VEH",
                   title="Human Multiome: MotifsDown in PIMOup vs PIMOdown", 
                   subtitle="highlighted from G523 N_BMP_vs_N_VEH_deseq2")
dev.off()

```

# Motif Compare old vs new (PIMO70_b added)

```{r motif compare old vs new, eval = FALSE}

# write_csv(motifsUp_homer.df, here("data","bulkATAC","2025_10-17_human_multiome_PIMOup_vs_PIMOdown_HOMER_motifsUP.csv"),
#                       col_names=TRUE)
# write_csv(motifsDown_homer.df, here("data","bulkATAC","2025_10-17_human_multiome_PIMOup_vs_PIMOdown_HOMER_motifsDown.csv"),
#                       col_names=TRUE)

write_csv(motifsUp_homer.df, here("data","bulkATAC","2025_12-08_human_multiome_PIMOup_vs_PIMOdown_HOMER_motifsUP.csv"),
                      col_names=TRUE)
write_csv(motifsDown_homer.df, here("data","bulkATAC","2025_12-08_human_multiome_PIMOup_vs_PIMOdown_HOMER_motifsDown.csv"),
                      col_names=TRUE)

# old vs new: motifsUP
read_csv(here("data","bulkATAC","2025_10-17_human_multiome_PIMOup_vs_PIMOdown_HOMER_motifsUP.csv"),col_names=TRUE)
read_csv(here("data","bulkATAC","2025_12-08_human_multiome_PIMOup_vs_PIMOdown_HOMER_motifsUp.csv"),col_names=TRUE)

# old vs new:  motifsDown
read_csv(here("data","bulkATAC","2025_10-17_human_multiome_PIMOup_vs_PIMOdown_HOMER_motifsDown.csv"),col_names=TRUE)
read_csv(here("data","bulkATAC","2025_12-08_human_multiome_PIMOup_vs_PIMOdown_HOMER_motifsDown.csv"),col_names=TRUE)
```
Looks like the motifs enriched are similar but more significant in the new one. Slight variations in ranking too, at least for the top motifs.
## export motifs for supplemental table
```{r export motifs for supplemental table, eval = FALSE}
# export the pairwise PIMO_up_status motifs (PIMOup_vs_PIMOdown)
motifsUp_homer.df_cleaned <- homer_motif_cleaning(motifsUp_homer.df)
motifsDown_homer.df_cleaned <- homer_motif_cleaning(motifsDown_homer.df)

write_csv(
  motifsUp_homer.df_cleaned,
  here(
    "data",
    "human_multiome",
    "motifEnrichment_homer_PIMO_up_status",
    "motifsUp_PIMO_up_status_PIMOup_vs_PIMOdown_homer.csv"
  )
)

write_csv(
  motifsDown_homer.df_cleaned,
  here(
    "data",
    "human_multiome",
    "motifEnrichment_homer_PIMO_up_status",
    "motifsDown_PIMO_up_status_PIMOup_vs_PIMOdown_homer.csv"
  )
)

# EDA
motifsUp_homer.df_cleaned$TF_cleaned %>% length # 332
motifsUp_homer.df_cleaned$TF_fixed %>% unique %>% length # 324
motifsUp_homer.df_cleaned$TF_cleaned %>% unique %>% length # 320

# what the duplicates? 8 after fixing, 12 after cleaning
motifsUp_homer.df_cleaned$TF_cleaned[duplicated(motifsUp_homer.df_cleaned$TF_cleaned)]

```
1. FOXA1 (before merging: FOXA1.Forkhead_88, FOXA1.Forkhead_87)
2. GATA-Zf
3. GATA3-Zf
4. GRE
5. HRE
6. Nr5a2
7. OCT:OCT
8. Pax7-Paired
9. RAR:RXR
10. STAT6
11. c-Myc
12. p53

# Motif Compare: Mouse Multiome
Here we load the significant motifs (padj < 0.05) from pairwise comparison of homer motifs for PIMO_up_status (PIMOup vs PIMOdown).
## Motif Rank plot
```{r compare to mouse multiome}
# load mouse multiome sig motifs (from PIMOup vs PIMOdown PIMO_up_status)
ms_motifsUp_homer_pw.sig <- read_csv(
     here(
    "data",
    "mouse_multiome",
    "motifEnrichment_homer_PIMO_up_status_PIMOup_vs_PIMOdown",
    "motifsUp_PIMOupstatus_PIMOupvsPIMOdown_homer_padj_lt_05.csv"
     ),
    col_names = FALSE
) %>% pull()
ms_motifsDown_homer_pw.sig <- read_csv(
     here(
    "data",
    "mouse_multiome",
    "motifEnrichment_homer_PIMO_up_status_PIMOup_vs_PIMOdown",
    "motifsDown_PIMOupstatus_PIMOupvsPIMOdown_homer_padj_lt_05.csv"
     ),
    col_names = FALSE
) %>% pull()


# plot
motifsUp_homer.df %>%
  homer_motif_cleaning() %>%
  plotMotifEnrichments_final(common_tfs_vector = ms_motifsUp_homer_pw.sig) +
    ggtitle("Human Multiome: Motifs Up (PIMOup vs PIMOdown) in common with Mouse Multiome", 
          subtitle = "Top 25 Significantly enriched motifs"
          )

motifsDown_homer.df %>%
  homer_motif_cleaning() %>%
  plotMotifEnrichments_final(common_tfs_vector = ms_motifsDown_homer_pw.sig) +
  ggtitle("Human Multiome: Motifs Down (PIMOup vs PIMOdown) in common with Mouse Multiome", 
          subtitle = "Top 25 Significantly enriched motifs"
          )

# intersect motifs
intersect(motifsUp_homer.df %>%
  homer_motif_cleaning() %>% pull(TF_cleaned),
  ms_motifsUp_homer_pw.sig
  )

intersect(motifsUp_homer.df %>%
  homer_motif_cleaning() %>% pull(TF_cleaned),
  ms_motifsUp_homer_pw.sig
  )
```
## Venn Diagram:
### motifsUp and motifsDown
```{r venn diagram for motifs}
library(VennDiagram)
library(RColorBrewer)
myColours <- brewer.pal(2,"Pastel2") # colors

# list of significant motifs
motifsUp_lt <- list(
  human_multiome_motifsUp = motif_summary(motifsUp_homer.df),
  mouse_multiome_motifsUp = ms_motifsUp_homer_pw.sig
)
motifsDown_lt <- list(
  human_multiome_motifsUp = motif_summary(motifsDown_homer.df),
  mouse_multiome_motifsUp = ms_motifsDown_homer_pw.sig
)

grid.newpage()
pushViewport(viewport(width=unit(0.8, "npc"), height = unit(0.8, "npc")))
venn.diagram(
  motifsUp_lt,
  NULL,
  fill = myColours[1:2],
  cex = 2,
  cat.cex = 0.6,
  lty = "blank",
  category.names = c(
    "human_motifsUp",
    "mouse_motifsUp"
  ),
  cat.default.pos = "outer",
  main = "Human Multiome vs Mouse Multiome: PIMOup_vs_PIMOdown PIMO_up_status motifsUp",
  sub = "Significant motifs padj <0.05"
) %>%
grid.draw()

grid.newpage()
pushViewport(viewport(width=unit(0.8, "npc"), height = unit(0.8, "npc")))
venn.diagram(
  motifsDown_lt,
  NULL,
  fill = myColours[1:2],
  cex = 2,
  cat.cex = 0.6,
  lty = "blank",
  category.names = c(
    "human_motifsDown",
    "mouse_motifsDown"
  ),
  cat.default.pos = "outer",
  main = "Human Multiome vs Mouse Multiome: PIMOup_vs_PIMOdown PIMO_up_status motifsDown",
  sub = "Significant motifs padj <0.05"
) %>%
grid.draw()
  

```
### extract overlaps
```{r extract overlaps}

calculate.overlap(motifsUp_lt) %>%
print()

calculate.overlap(motifsDown_lt) %>%
print()
```

## Export to PDF
```{r export to pdf}
pdf(here("Plots","2026-01-19_human_multiome_PIMOup_vs_PIMOdown_motifs_vs_mouse_multiome_homer_motif_ranking.pdf"))
motifsUp_homer.df %>%
  homer_motif_cleaning() %>%
  plotMotifEnrichments_final(common_tfs_vector = ms_motifsUp_homer_pw.sig) +
    ggtitle("Human Multiome: Motifs Up (PIMOup vs PIMOdown) in common with Mouse Multiome", 
          subtitle = "Top 25 Significantly enriched motifs"
          )

motifsDown_homer.df %>%
  homer_motif_cleaning() %>%
  plotMotifEnrichments_final(common_tfs_vector = ms_motifsDown_homer_pw.sig) +
  ggtitle("Human Multiome: Motifs Down (PIMOup vs PIMOdown) in common with Mouse Multiome", 
          subtitle = "Top 25 Significantly enriched motifs"
          )
dev.off()


# venn 
pdf(here("Plots","2026-01-19_human_multiome_PIMOup_vs_PIMOdown_motifs_vs_mouse_multiome_homer_venn.pdf"))
grid.newpage()
pushViewport(viewport(width=unit(0.8, "npc"), height = unit(0.8, "npc")))
venn.diagram(
  motifsUp_lt,
  NULL,
  fill = myColours[1:2],
  cex = 2,
  cat.cex = 0.6,
  lty = "blank",
  category.names = c(
    "human_motifsUp",
    "mouse_motifsUp"
  ),
  cat.default.pos = "outer",
  main = "Human Multiome vs Mouse Multiome: PIMOup_vs_PIMOdown PIMO_up_status motifsUp",
  sub = "Significant motifs padj <0.05"
) %>%
grid.draw()

grid.newpage()
pushViewport(viewport(width=unit(0.8, "npc"), height = unit(0.8, "npc")))
venn.diagram(
  motifsDown_lt,
  NULL,
  fill = myColours[1:2],
  cex = 2,
  cat.cex = 0.6,
  lty = "blank",
  category.names = c(
    "human_motifsDown",
    "mouse_motifsDown"
  ),
  cat.default.pos = "outer",
  main = "Human Multiome vs Mouse Multiome: PIMOup_vs_PIMOdown PIMO_up_status motifsDown",
  sub = "Significant motifs padj <0.05"
) %>%
grid.draw()
dev.off()
```

# Motif Compare: Human Multiome k3_program scPOS-MP
```{r}
# SE objects from ArchR output
markersPeak_k3_program
k3_motifsUp
k3_motifsDown

# EDA on SE object
assay(markersPeak_k3_program)

# Convert motifs SE to df
k3_motifsUp.df <- motifSEtoDF(k3_motifsUp)
k3_motifsDown.df <- motifSEtoDF(k3_motifsDown)

k3_motifsUp.df
k3_motifsDown.df

# additional homer motifs cleaning so motifs match b/w multiome and bulk ATAC 
k3_motifsUp.df_cleaned <- homer_motif_cleaning(k3_motifsUp.df)
k3_motifsDown.df_cleaned <- homer_motif_cleaning(k3_motifsDown.df)

# plotMotifEnrichments for each group
# plotMotifEnrichments_new <- function(df, common_tfs_vector, show_labels = TRUE, show_labels_qty = 25) 

# default
k3_motifsUp.df_cleaned %>% 
    filter(group == "program_1") %>%  
    plotMotifEnrichments_final() +
  ggtitle("Human Multiome: motifsUp scPOS-MP k3_program (program_1)",
          subtitle = "Top 25 motifs padj<0.05")

# compare to motifsUp homer (original dataset PIMOup_vs_PIMOdown)

homer_motif_cleaning(motifsUp_homer.df) %>% 
  filter(mlog10Padj > 1.30103) %>%
  pull(TF_cleaned)

homer_motif_cleaning(motifsDown_homer.df) %>% 
  filter(mlog10Padj > 1.30103) %>%
  pull(TF_cleaned)

# human multiome motifsUp (k3_program) vs human multiome motifsUp (original dataset PIMOup_vs_PIMOdown) 
common_OG_vs_k3_motifsUp <- 
  homer_motif_cleaning(motifsUp_homer.df) %>%
    filter(mlog10Padj > 1.30103) %>%
    pull(TF_cleaned)

common_OG_vs_k3_motifsDown <- 
  homer_motif_cleaning(motifsDown_homer.df) %>%
    filter(mlog10Padj > 1.30103) %>%
    pull(TF_cleaned)


```
## export PDF
```{r export pdf 1, eval = FALSE}
pdf(here("Plots","2026-03-09_human_multiome_k3_program_motifs_homer_motif_ranking.pdf"))

# default
k3_motifsUp.df_cleaned %>% 
    filter(group == "program_1") %>%  
    plotMotifEnrichments_final() +
  ggtitle("Human Multiome: motifsUp scPOS-MP k3_program (program_1)",
          subtitle = "Top 25 motifs padj<0.05")

k3_motifsUp.df_cleaned %>% 
    filter(group == "program_2") %>%  
    plotMotifEnrichments_final() +
  ggtitle("Human Multiome: motifsUp scPOS-MP k3_program (program_2)",
          subtitle = "Top 25 motifs padj<0.05")

k3_motifsUp.df_cleaned %>% 
    filter(group == "program_3") %>%  
    plotMotifEnrichments_final() +
  ggtitle("Human Multiome: motifsUp scPOS-MP k3_program (program_3)",
          subtitle = "Top 25 motifs padj<0.05")


k3_motifsDown.df_cleaned %>% 
    filter(group == "program_1") %>%  
    plotMotifEnrichments_final() +
  ggtitle("Human Multiome: motifsDown scPOS-MP k3_program (program_1)",
          subtitle = "Top 25 motifs padj<0.05")

k3_motifsDown.df_cleaned %>% 
    filter(group == "program_2") %>%  
    plotMotifEnrichments_final() +
  ggtitle("Human Multiome: motifsDown scPOS-MP k3_program (program_2)",
          subtitle = "Top 25 motifs padj<0.05")

k3_motifsDown.df_cleaned %>% 
    filter(group == "program_3") %>%  
    plotMotifEnrichments_final() +
  ggtitle("Human Multiome: motifsDown scPOS-MP k3_program (program_3)",
          subtitle = "Top 25 motifs padj<0.05")
dev.off()



pdf(here("Plots","2026-03-09_human_multiome_k3_program_vs_PIMOup_vs_PIMOdown_motifs_homer_motif_ranking.pdf"))

# compared to human_multiome PIMOup_vs_PIMOdown

k3_motifsUp.df_cleaned %>% 
    filter(group == "program_1") %>%  
    plotMotifEnrichments_final(common_tfs_vector = common_OG_vs_k3_motifsUp) +
  ggtitle("Human Multiome: motifsUp scPOS-MP k3_program (program_1)",
          subtitle = "Compared to motifsUp PIMOup_vs_PIMOdown padj<0.05")

k3_motifsUp.df_cleaned %>% 
    filter(group == "program_2") %>%  
    plotMotifEnrichments_final(common_tfs_vector = common_OG_vs_k3_motifsUp) +
  ggtitle("Human Multiome: motifsUp scPOS-MP k3_program (program_2)",
          subtitle = "Compared to motifsUp PIMOup_vs_PIMOdown padj<0.05")

k3_motifsUp.df_cleaned %>% 
    filter(group == "program_3") %>%  
    plotMotifEnrichments_final(common_tfs_vector = common_OG_vs_k3_motifsUp) +
  ggtitle("Human Multiome: motifsUp scPOS-MP k3_program (program_3)",
          subtitle = "Compared to motifsUp PIMOup_vs_PIMOdown padj<0.05")

k3_motifsDown.df_cleaned %>% 
    filter(group == "program_1") %>%  
    plotMotifEnrichments_final(common_tfs_vector = common_OG_vs_k3_motifsDown) +
  ggtitle("Human Multiome: motifsDown scPOS-MP k3_program (program_1)",
          subtitle = "Compared to motifsDown PIMOup_vs_PIMOdown padj<0.05")

k3_motifsDown.df_cleaned %>% 
    filter(group == "program_2") %>%  
    plotMotifEnrichments_final(common_tfs_vector = common_OG_vs_k3_motifsDown) +
  ggtitle("Human Multiome: motifsDown scPOS-MP k3_program (program_2)",
          subtitle = "Compared to motifsDown PIMOup_vs_PIMOdown padj<0.05")

k3_motifsDown.df_cleaned %>% 
    filter(group == "program_3") %>%  
    plotMotifEnrichments_final(common_tfs_vector = common_OG_vs_k3_motifsDown) +
  ggtitle("Human Multiome: motifsDown scPOS-MP k3_program (program_3)",
          subtitle = "Compared to motifsDown PIMOup_vs_PIMOdown padj<0.05")

dev.off()

```
## export motifs for supplemental table
```{r export motifs for supplement, eval = FALSE}

write_csv(
  k3_motifsUp.df_cleaned,
  here(
    "data",
    "human_multiome",
    "human_multiome_hmmp_k3_program",
    "motifEnrichment_homer_k3_program",
    "motifsUp_k3program_homer.csv"
  )
)
write_csv(
  k3_motifsDown.df_cleaned,
  here(
    "data",
    "human_multiome",
    "human_multiome_hmmp_k3_program",
    "motifEnrichment_homer_k3_program",
    "motifsDown_k3program_homer.csv"
  )
)

```


# Motif Compare: Human Multiome nomura (meta_module) 
```{r}
# SE objects from ArchR output

nomura_motifsUp
nomura_motifsDown

# EDA on SE object
assay(nomura_motifsUp)

# Convert motifs SE to df
nomura_motifsUp.df <- motifSEtoDF(nomura_motifsUp)
nomura_motifsDown.df <- motifSEtoDF(nomura_motifsDown)

nomura_motifsUp.df
nomura_motifsDown.df

# additional homer motifs cleaning so motifs match b/w multiome and bulk ATAC 
nomura_motifsUp.df_cleaned <- homer_motif_cleaning(nomura_motifsUp.df)
nomura_motifsDown.df_cleaned <- homer_motif_cleaning(nomura_motifsDown.df)

# Plotting default (eg. AC.like)
nomura_motifsUp.df_cleaned %>% 
    filter(group == "AC.like") %>%  
    plotMotifEnrichments_final() +
  ggtitle("Human Multiome: motifsUp nomura meta_module: AC.like",
          subtitle = "Top 25 motifs padj<0.05")
```
## export PDF
```{r export pdf, eval = FALSE}
# default
# 1. Get a vector of unique groups
groups <- unique(nomura_motifsUp.df_cleaned$group)

# 2. Iterate and create a list of plots
all_plots_up <- groups %>% set_names() %>% map(function(g) {
  
  nomura_motifsUp.df_cleaned %>% 
    filter(group == g) %>%  
    plotMotifEnrichments_final() +
    labs(
      title = paste("Human Multiome: motifsUp nomura meta_module:", g),
      subtitle = "Top 25 motifs padj<0.05"
    )
})
all_plots_down <- groups %>% set_names() %>% map(function(g) {
    nomura_motifsDown.df_cleaned %>% 
    filter(group == g) %>%  
    plotMotifEnrichments_final() +
    labs(
      title = paste("Human Multiome: motifsDown nomura meta_module:", g),
      subtitle = "Top 25 motifs padj<0.05"
    )
})

pdf(here("Plots","2026-03-19_human_multiome_nomura_meta_module_motifs_homer_motif_ranking.pdf"))
all_plots_up
all_plots_down
dev.off()


```
## export motifs for supplemental table
```{r export motifs for supplement 2, eval = FALSE}

write_csv(
  nomura_motifsUp.df_cleaned,
  here(
    "data",
    "human_multiome",
    "human_multiome_hmmp_nomura",
    "motifEnrichment_homer_meta_module",
    "motifsUp_meta_module_homer.csv"
  )
)
write_csv(
  nomura_motifsDown.df_cleaned,
  here(
    "data",
    "human_multiome",
    "human_multiome_hmmp_nomura",
    "motifEnrichment_homer_meta_module",
    "motifsDown_meta_module_homer.csv"
  )
)

```

```{r nomura motifs EDA}
# here we check the number of HOMER motifs for nomura, and number of unique motifs
nomura_motifsUp.df_cleaned$TF %>% unique() %>% length
nomura_motifsUp.df_cleaned$TF_cleaned %>% unique() %>% length

```
So HOMER has 332 motifs examined, but after we cleaned the motif naming, only 320 remain. There were 12 pairs of motifs that were consolidated as the following:
1. FOXA1 (before merging: FOXA1.Forkhead_88, FOXA1.Forkhead_87)
2. GATA-Zf
3. GATA3-Zf
4. GRE
5. HRE
6. Nr5a2
7. OCT:OCT
8. Pax7-Paired
9. RAR:RXR
10. STAT6
11. c-Myc
12. p53

# Motif Compare heatmap:
## load homer motifs from multiome
```{r motif compare heatmap}
# Load required libraries
library(circlize)
library(RColorBrewer) # For automatic color generation
# Load the datasets
meta_module <- read.csv(
  here(
    "data",
    "human_multiome",
    "human_multiome_hmmp_nomura",
    "motifEnrichment_homer_meta_module",
    "motifsUp_meta_module_homer.csv"
  )
)

k3_program <- read.csv(
  here(
    "data",
    "human_multiome",
    "human_multiome_hmmp_k3_program",
    "motifEnrichment_homer_k3_program",
    "motifsUp_k3program_homer.csv"
  )
)

# 1. Scalable Data Loading
# Create a named list of your file paths
files <- list(
  "MetaModule" =  here(
    "data",
    "human_multiome",
    "human_multiome_hmmp_nomura",
    "motifEnrichment_homer_meta_module",
    "motifsUp_meta_module_homer.csv"
  ),
  "K3Program"  =  here(
    "data",
    "human_multiome",
    "human_multiome_hmmp_k3_program",
    "motifEnrichment_homer_k3_program",
    "motifsUp_k3program_homer.csv"
  )
  # Add more here: "NewDataset" = "path/to/file.csv"
)
```

```{r motif compare heatmap continued}

df_list <- lapply(files, read.csv)

prepare_multi_data <- function(data_list) {
  bind_rows(data_list, .id = "origin") %>%
    group_by(TF_cleaned, group, origin) %>%
    summarise(mlog10Padj = max(mlog10Padj, na.rm = TRUE), .groups = "drop")
}

combined_df <- prepare_multi_data(df_list)

# 2. Updated Plotting Function with Printing Feature
plot_tf_heatmap <- function(data, 
                            groups_to_include = NULL, 
                            top_n_tfs = 50, 
                            normalize_to_max = TRUE,
                            filter_by_padj = TRUE,      
                            min_mlog10padj = 1.30103) {     
  
  # Step A: Filter for user-specified groups
  plot_data <- data
  if (!is.null(groups_to_include)) {
    plot_data <- plot_data %>% filter(group %in% groups_to_include)
  }
  
  # Step B: Pivot to wide matrix
  heatmap_mat <- plot_data %>%
    select(TF_cleaned, group, mlog10Padj) %>%
    pivot_wider(names_from = group, values_from = mlog10Padj, values_fill = 0) %>%
    column_to_rownames("TF_cleaned") %>%
    as.matrix()
  
  # Step C: Significance Filtering
  if (filter_by_padj) {
    keep_rows <- apply(heatmap_mat, 1, function(x) any(x >= min_mlog10padj))
    heatmap_mat <- heatmap_mat[keep_rows, , drop = FALSE]
    
    if (nrow(heatmap_mat) == 0) {
      stop(paste("No motifs passed the threshold of -log10Padj >=", min_mlog10padj))
    }
  }
  
  # Step D: Normalization (Max Scaling to 100)
  if (normalize_to_max) {
    heatmap_mat_display <- apply(heatmap_mat, 2, function(x) {
      m <- max(x, na.rm = TRUE)
      if(m == 0) return(x)
      return((x / m) * 100)
    })
    col_fun <- colorRamp2(c(0, 100), c("white", "red"))
    legend_label <- "Relative Enrichment (0-100)"
  } else {
    heatmap_mat_display <- heatmap_mat
    col_fun <- colorRamp2(c(0, max(heatmap_mat)), c("white", "red"))
    legend_label <- "-log10(Padj)"
  }
  
  # Step E: Select Top N TFs
  importance <- apply(heatmap_mat_display, 1, max)
  top_tfs <- names(sort(importance, decreasing = TRUE))[1:min(top_n_tfs, nrow(heatmap_mat_display))]
  heatmap_mat_display <- heatmap_mat_display[top_tfs, , drop = FALSE]
  
  # --- NEW: Print Summary to Console ---
  cat("\n--- Heatmap Summary ---\n")
  cat("Number of motifs displayed:", nrow(heatmap_mat_display), "\n")
  cat("Motifs included:\n")
  print(rownames(heatmap_mat_display))
  cat("------------------------\n\n")
  # -------------------------------------
  
  # Step F: Dynamic Column Annotation
  group_origin_map <- plot_data %>% select(group, origin) %>% distinct() %>% deframe()
  unique_origins <- unique(plot_data$origin)
  origin_colors <- setNames(brewer.pal(max(3, length(unique_origins)), "Set1")[1:length(unique_origins)], 
                            unique_origins)
  
  col_ann <- columnAnnotation(
    Dataset = group_origin_map[colnames(heatmap_mat_display)],
    col = list(Dataset = origin_colors),
    show_annotation_name = TRUE
  )
  
  # Step G: Generate Heatmap
  Heatmap(heatmap_mat_display,
          name = legend_label,
          col = col_fun,
          top_annotation = col_ann,
          cluster_columns = TRUE,
          cluster_rows = TRUE,
          row_names_gp = gpar(fontsize = 8),
          column_title = paste0("Top TFs (Filter: ", 
                                ifelse(filter_by_padj, paste0("-log10P >= ", min_mlog10padj), "None"), ")"))
}

# 3. Usage Examples:

# Scenario 1. Default significance filter (mlog10padj > 1.30103) with specific groups
selected <- c("AC.like", "MES.like", "Hypoxia","program_1", "program_2", "program_3")
ht1 <- plot_tf_heatmap(combined_df, groups_to_include = selected, filter_by_padj = TRUE)
draw(ht1)

# Scenario 2: Strict filter (Only show TFs where at least one sample is highly significant)
ht2 <- plot_tf_heatmap(combined_df, groups_to_include = selected, filter_by_padj = TRUE, min_mlog10padj = 20)
draw(ht2)
```
## export PDF

# check h5 file integrity
```{r check h5 files}
library(Seurat)
library(Signac)
counts1 <- Read10X_h5(
  here("data","Gaiti_multiome", "pimo66","filtered_feature_bc_matrix.h5")
)

counts2 <- Read10X_h5(
  here("data","Gaiti_multiome","GSM8820702_6419_A" ,"filtered_feature_bc_matrix.h5")
)

counts3 <- Read10X_h5(
  here("data","Gaiti_multiome","GSM8820703_6419_B" ,"filtered_feature_bc_matrix.h5")
)


```

Here we compare the original PIMO .h5 file versus the ones we generated from barcodes, matrix, and features.tsv.gz using dropletutils.

The original PIMO .h5 shows a warning, but not the others: `Genome matrix has multiple modalities, returning a list of matrices for this genome`, thus we should mimic this behavior.

```{r}
class(counts1) # list
counts1

class(counts2) # #[1] "dgCMatrix"

counts2

class(counts3)
counts3

library(DropletUtils)

# compare rownames 

nrow(counts2) # 201635
nrow(counts3) # 98740

counts2
counts3
```
# Motif compare GSC pairwise bubble plot dot plot

```{r GSC pairwise bubbleplot dotplot}
library(tidyverse)

plot_homer_dots_binned <- function(df, fdr_cutoff = 0.05, top_n_rank = 60) {
  
  # 1. Pre-process and categorize significance levels
  plot_data <- df %>%
    filter(`q-value/FDR` < fdr_cutoff) %>%
    filter(Rank <= top_n_rank) %>%
    mutate(
      # Categorize FDR into discrete bins
      fdr_bin = case_when(
        `q-value/FDR` < 0.001 ~ "< 0.001",
        `q-value/FDR` >= 0.001 & `q-value/FDR` < 0.01 ~ "0.001 - 0.01",
        `q-value/FDR` >= 0.01  & `q-value/FDR` < 0.05 ~ "0.01 - 0.05",
        TRUE ~ "> 0.05"
      ),
      # Set factor levels to ensure the legend displays in the correct order
      fdr_bin = factor(fdr_bin, levels = c("< 0.001", "0.001 - 0.01", "0.01 - 0.05")),
      
      # Ensure TF is a factor sorted by its average Rank for a cleaner plot
      #TF = fct_reorder(TF, Rank, .fun = mean, .desc = TRUE)
    )

  # 2. Generate the plot
  ggplot(plot_data, aes(x = dataset, y = TF)) +
    geom_point(aes(size = fdr_bin, color = `Fold Enrichment`)) +
    # Define custom sizes for the bins (Smallest bin = size 2, Largest = size 6)
    scale_size_manual(
      values = c("< 0.001" = 6, "0.001 - 0.01" = 3.5, "0.01 - 0.05" = 1.5),
      name = "q-value/FDR"
    ) +
    # Color scale (Magma highlights high enrichment well)
    scale_color_viridis_c(option = "magma", direction = -1) +
    theme_bw() +
    labs(
      title = paste("Top", top_n_rank, "Motifs"),
      subtitle = paste("Filtered for Rank <=", top_n_rank, "and FDR <", fdr_cutoff),
      x = "Dataset",
      y = "Transcription Factor"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
      axis.text.y = element_text(size = 7),
      panel.grid.major = element_line(color = "grey95"),
      legend.key.height = unit(0.5, "cm")
    )
}

# --- To use this function: ---
final_plot <- plot_homer_dots_binned(final_table)
final_plot
print(final_plot)

# --- Example Usage ---
# Assuming your table is loaded as 'final_table'
pdf(here("data","bulkATAC",
         "2026-04-19_G361_G411_RW492_RW555_pairwise_HOMER_motifs_bubbleplot.pdf"), 
    width = 10, height = 15)
plot_homer_dots_binned(final_table, fdr_cutoff = 0.05, top_n_rank = 10)
plot_homer_dots_binned(final_table, fdr_cutoff = 0.05, top_n_rank = 25)
plot_homer_dots_binned(final_table, fdr_cutoff = 0.05, top_n_rank = 60)
dev.off()
```

# Binary Heatmap for GSC motifs:
```{r binary heatmap for GSC motifs, eval = FALSE}
library(tidyverse)
library(ComplexHeatmap)
library(circlize)

plot_homer_heatmap_annotated <- function(df, fdr_cutoff = 0.05, top_n_rank = 60) {
  
  # 1. Identify passing TFs
  passing_tfs <- df %>%
    filter(`q-value/FDR` < fdr_cutoff, Rank <= top_n_rank) %>%
    pull(TF) %>%
    unique()

  # 2. Prepare the matrix
  heatmap_mat <- df %>%
    filter(TF %in% passing_tfs) %>%
    mutate(pass = if_else(`q-value/FDR` < fdr_cutoff & Rank <= top_n_rank, 1, 0)) %>%
    group_by(TF, dataset) %>%
    summarize(pass = max(pass, na.rm = TRUE), .groups = "drop") %>%
    pivot_wider(names_from = dataset, values_from = pass, values_fill = 0) %>%
    column_to_rownames("TF") %>%
    as.matrix()

  # 3. Decompose dataset names for Metadata
  # This creates a dataframe where each row matches a column in our heatmap
  col_metadata <- data.frame(dataset_full = colnames(heatmap_mat)) %>%
    mutate(
      GSC_line = str_extract(dataset_full, "^[^_]+"),
      enrichment = str_extract(dataset_full, "hypoxia_enrich|normoxia_enrich"),
      # Extract the middle part as the condition
      condition = str_replace(dataset_full, paste0("^", GSC_line, "_"), ""),
      condition = str_replace(condition, paste0("_", enrichment, "$"), "")
    ) %>%
    column_to_rownames("dataset_full")

  # 4. Create Heatmap Annotations
  # We define distinct colors for the different metadata categories
  ha <- HeatmapAnnotation(
    GSC_line = col_metadata$GSC_line,
    Condition = col_metadata$condition,
    Enrichment = col_metadata$enrichment,
    col = list(
      GSC_line = c("G361" = "#E41A1C", "G411" = "#377EB8"),
      Enrichment = c("hypoxia_enrich" = "black", "normoxia_enrich" = "grey80"),
      Condition = c("N_72_vs_A_24" = "#4DAF4A", "N_72_vs_H_24" = "#984EA3", 
                    "N_72_vs_H_72" = "#FF7F00")
    ),
    annotation_name_side = "left"
  )

  # 5. Draw the Heatmap
  Heatmap(heatmap_mat, 
          name = "Enriched", 
          col = c("0" = "white", "1" = "firebrick"),
          top_annotation = ha, # Add the metadata here
          rect_gp = gpar(col = "grey95", lwd = 0.5),
          
          #column_title = "HOMER Motif Enrichment Pattern",
          column_title = paste0("FDR< ", fdr_cutoff, " and Top ",top_n_rank, " motifs"),
          cluster_rows = TRUE, 
          cluster_columns = TRUE,
          show_column_names = FALSE, # Hidden because annotations explain the columns
          
          row_names_gp = gpar(fontsize = 7),
          column_names_gp = gpar(fontsize = 8))
}

# --- Usage ---
# hm <- plot_homer_heatmap_annotated(final_table)
# draw(hm, merge_legend = TRUE)

pdf(here("data","bulkATAC",
         "2026-04-19_G361_G411_RW492_RW555_pairwise_HOMER_motifs_binaryHeatmap.pdf"), 
    width = 10, height = 15)
hm <- plot_homer_heatmap_annotated(final_table, fdr_cutoff=0.05, top_n_rank = 10)
draw(hm, merge_legend = TRUE)
hm <- plot_homer_heatmap_annotated(final_table, fdr_cutoff=0.05, top_n_rank = 25)
draw(hm, merge_legend = TRUE)
hm <- plot_homer_heatmap_annotated(final_table)
draw(hm, merge_legend = TRUE)


hm <- plot_homer_heatmap_annotated(final_table, fdr_cutoff=0.005, top_n_rank = 25)
draw(hm, merge_legend = TRUE)
dev.off()

# for all n=2, normoxia vs hypoxia; non-pairwise
library(tidyverse)
library(ComplexHeatmap)

# This function works on both pairwise and all analysis
plot_homer_heatmap_flexible <- function(df, 
                                        fdr_cutoff = 0.05, 
                                        top_n_rank = 60, 
                                        annotation_colors = NULL) {
  
  # 1. Identify passing TFs
  passing_tfs <- df %>%
    filter(`q-value/FDR` < fdr_cutoff, Rank <= top_n_rank) %>%
    pull(TF) %>%
    unique()

  if(length(passing_tfs) == 0) return(NULL)

  # 2. Prepare the binary matrix
  heatmap_mat <- df %>%
    filter(TF %in% passing_tfs) %>%
    mutate(pass = if_else(`q-value/FDR` < fdr_cutoff & Rank <= top_n_rank, 1, 0)) %>%
    group_by(TF, dataset) %>%
    summarize(pass = max(pass, na.rm = TRUE), .groups = "drop") %>%
    pivot_wider(names_from = dataset, values_from = pass, values_fill = 0) %>%
    column_to_rownames("TF") %>%
    as.matrix()

  # 3. Decomposition Logic
  col_metadata <- data.frame(dataset_full = colnames(heatmap_mat)) %>%
    mutate(
      enrichment = str_extract(dataset_full, "hypoxia_enrich|normoxia_enrich"),
      remaining = str_remove(dataset_full, paste0("_?", enrichment)),
      GSC_line = if_else(str_detect(remaining, "_"), 
                         str_extract(remaining, "^[^_]+"), 
                         NA_character_),
      condition = if_else(!is.na(GSC_line), 
                          str_remove(remaining, paste0("^", GSC_line, "_?")), 
                          if_else(remaining == "", NA_character_, remaining))
    ) %>%
    mutate(across(everything(), ~replace_na(.x, "N/A"))) %>%
    column_to_rownames("dataset_full")

  # Select existing annotation columns
  anno_df <- col_metadata %>% select(GSC_line, condition, enrichment)
  anno_df <- anno_df[, colSums(anno_df == "N/A") < nrow(anno_df), drop = FALSE]

  # 4. Create Annotations with Custom Colors
  # If annotation_colors is provided, HeatmapAnnotation will use it.
  # If it's NULL, it will use default random colors.
  ha <- NULL
  if(ncol(anno_df) > 0) {
    ha <- HeatmapAnnotation(
      df = anno_df, 
      col = annotation_colors, # This is where the magic happens
      annotation_name_side = "left"
    )
  }

  # 5. Draw Heatmap
  Heatmap(heatmap_mat, 
          name = "Enriched", 
          col = c("0" = "white", "1" = "firebrick"),
          top_annotation = ha,
          rect_gp = gpar(col = "grey95", lwd = 0.5),
          column_title = paste0("FDR< ", fdr_cutoff, " and Top ",top_n_rank, " motifs"),
          cluster_rows = TRUE, 
          cluster_columns = ncol(heatmap_mat) > 1,
          row_names_gp = gpar(fontsize = 6))
}
# Define your preferred color schemes
my_colors <- list(
  GSC_line = c(
    "G361" = "#E41A1C", 
    "G411" = "#377EB8", 
    "N/A"  = "grey90"
  ),
  enrichment = c(
    "hypoxia_enrich" = "#000000",   # Black for hypoxia
    "normoxia_enrich" = "#D9D9D9"  # Light grey for normoxia
  ),
  condition = c(
    "N_72_vs_A_24" = "#4DAF4A", 
    "N_72_vs_H_24" = "#984EA3", 
    "N_72_vs_H_72" = "#FF7F00",
    "N/A"          = "white"
  )
)

# --- Usage with your new CSV ---
# df_combined <- read_csv("combined_homer_results.csv")
# hm <- plot_homer_heatmap_flexible(df_combined)
# draw(hm)

pdf(here("data","bulkATAC",
         "2026-04-19_G361_G411_RW492_RW555_all_normoxia_vs_hypoxia_HOMER_motifs_binaryHeatmap.pdf"), 
    width = 14, height = 10)

plot_homer_heatmap_flexible(G361_G411_all_motifs_df, fdr_cutoff=0.05, top_n_rank = 10, annotation_colors = my_colors)
plot_homer_heatmap_flexible(G361_G411_all_motifs_df, fdr_cutoff=0.05, top_n_rank = 25, annotation_colors = my_colors)
plot_homer_heatmap_flexible(G361_G411_all_motifs_df, fdr_cutoff=0.05, top_n_rank = 60, annotation_colors = my_colors)
plot_homer_heatmap_flexible(G361_G411_all_motifs_df, fdr_cutoff=0.005, top_n_rank = 25, annotation_colors = my_colors)

dev.off()
```
## New binary heatmap functions for homer motifs
```{r new homer heatmap function, eval = FALSE}
library(tidyverse)
library(ComplexHeatmap)

# Define colours used in heatmap
my_colors <- list(
  # 1. Cell Lines: Use distinct bold colors
  cell_line = c(
    "G361" = "cyan",   # teal
    "G411" = "orange",   # orange
    "G523" = "#4DAF4A",   # Green
    "G521" = "#984EA3",   # Purple (adding a buffer for future lines)
    "N/A"  = "grey90"
  ),
  
condition = c(
    # --- Hypoxia (Cool Blues) ---
    "hypoxia_enrich"  = "#081d58", # Midnight Blue
    "H_VEH_edgeR"     = "#084594", # Deep Blue
    "H_VEH_deseq2"    = "#2171b5", # Medium Blue
    "H_BMP_edgeR"     = "#4292c6", # Steel Blue
    "H_BMP_deseq2"    = "#6baed6", # Sky Blue
    
    # --- Normoxia (Warm Reds/Oranges) ---
    "normoxia_enrich" = "#fff5f0", # Warm Off-white
    "N_VEH_edgeR"     = "#99000d", # Blood Red
    "N_VEH_deseq2"    = "#cb181d", # Bright Red
    "N_BMP_edgeR"     = "#ef3b2c", # Orange-Red
    "N_BMP_deseq2"    = "#fb6a4a", # Coral/Salmon
    
    "N/A"             = "white"
  ),
  
  # 3. Dataset (Specific Comparisons)
  dataset = c(
    # 2023 Comparisons
    "G361_N_72_vs_A_24" = "#8DD3C7",
    "G361_N_72_vs_H_24" = "#FFFFB3",
    "G361_N_72_vs_H_72" = "#BEBADA",
    "G411_N_72_vs_A_24" = "#80B1D3",
    "G411_N_72_vs_H_24" = "#FDB462",
    "G411_N_72_vs_H_72" = "#B3DE69",
    
    # 2021 Comparisons
    "N_VEH_vs_H_VEH"    = "#FCCDE5",
    "N_BMP_vs_H_BMP"    = "#D9D9D9",
    "N_VEH_vs_N_BMP"    = "#BC80BD",
    "N/A"               = "white"
  )
)

plot_homer_heatmap <- function(df, 
                               fdr_cutoff = 0.05, 
                               top_n_rank = 60, 
                               annotation_colors = NULL) {
  
  # 1. Identify and Print passing TFs
  passing_tfs <- df %>%
    filter(`q-value/FDR` < fdr_cutoff, Rank <= top_n_rank) %>%
    pull(TF) %>%
    unique() %>%
    sort() # Alphabetical sorting for the console printout

  if(length(passing_tfs) == 0) {
    message("No motifs passed the thresholds.")
    return(NULL)
  }
  
  # Print the list to the console
  cat("--- Motifs Passing Thresholds ---\n")
  cat("FDR <", fdr_cutoff, "| Rank <=", top_n_rank, "\n")
  cat("Total TFs:", length(passing_tfs), "\n")
  cat(paste(passing_tfs, collapse = ", "), "\n")
  cat("---------------------------------\n\n")

  # 2. Build the binary matrix
  df_matrix <- df %>%
    filter(TF %in% passing_tfs) %>%
    mutate(
      sample_id = paste(cell_line, dataset, condition, sep = "|"),
      pass = if_else(`q-value/FDR` < fdr_cutoff & Rank <= top_n_rank, 1, 0)
    ) %>%
    group_by(TF, sample_id) %>%
    summarize(pass = max(pass, na.rm = TRUE), .groups = "drop") %>%
    pivot_wider(names_from = sample_id, values_from = pass, values_fill = 0) %>%
    column_to_rownames("TF") %>%
    as.matrix()

  # 3. Extract Metadata for Column Annotations
  col_metadata <- data.frame(sample_id = colnames(df_matrix)) %>%
    separate(sample_id, 
             into = c("cell_line", "dataset", "condition"), 
             sep = "\\|", 
             remove = FALSE) %>%
    column_to_rownames("sample_id")

  # 4. Create Heatmap Annotation
  ha_cols <- annotation_colors[names(annotation_colors) %in% colnames(col_metadata)]
  
  ha <- HeatmapAnnotation(
    df = col_metadata, 
    col = ha_cols,
    annotation_name_side = "left"
  )

  # 5. Render Heatmap with Dynamic Title
  Heatmap(df_matrix, 
          name = "Enriched", 
          col = c("0" = "white", "1" = "firebrick"),
          top_annotation = ha,
          rect_gp = gpar(col = "grey95", lwd = 0.5),
          
          # Dynamic Title including thresholds
          column_title = paste0("HOMER Motif Enrichment Pattern\n(FDR < ", 
                                fdr_cutoff, ", Rank <= ", top_n_rank, ")"),
          
          cluster_rows = TRUE, 
          cluster_columns = TRUE,
          show_column_names = FALSE,
          row_names_gp = gpar(fontsize = 8))
}

# example useage:

# 1. Use with your original experiment data
hm_detailed <- plot_homer_heatmap(final_table, annotation_colors = my_colors)
draw(hm_detailed)

# 2. Use with your 2021 G523 CSV data
hm_2021 <- plot_homer_heatmap(G523_final_table, annotation_colors = my_colors)
draw(hm_2021)

# 3. Use with stricter filters
hm_strict <- plot_homer_heatmap(final_table, fdr_cutoff = 0.01, top_n_rank = 10,
                                annotation_colors = my_colors)
draw(hm_strict)

# heatmap now with TF frequency filtering:
library(tidyverse)
library(ComplexHeatmap)

plot_homer_heatmap_freq <- function(df, 
                               fdr_cutoff = 0.05, 
                               top_n_rank = 60, 
                               min_conditions = 1,
                               annotation_colors = NULL) {
  
  # 1. Identify and Print passing TFs
  passing_tfs <- df %>%
    filter(`q-value/FDR` < fdr_cutoff, Rank <= top_n_rank) %>%
    pull(TF) %>%
    unique() %>%
    sort()

  if(length(passing_tfs) == 0) {
    message("No motifs passed the thresholds.")
    return(NULL)
  }
  
  cat("--- Motifs Passing Thresholds ---\n")
  cat("FDR <", fdr_cutoff, "| Rank <=", top_n_rank, "\n")
  cat("Total TFs:", length(passing_tfs), "\n")
  cat("---------------------------------\n\n")

  # 2. Build the binary matrix
  # Note: Ensure all grouping variables are characters to avoid mapping errors
  df_matrix <- df %>%
    filter(TF %in% passing_tfs) %>%
    mutate(
      sample_id = paste(as.character(cell_line), 
                        as.character(dataset), 
                        as.character(condition), 
                        sep = "|"),
      pass = if_else(`q-value/FDR` < fdr_cutoff & Rank <= top_n_rank, 1, 0)
    ) %>%
    group_by(TF, sample_id) %>%
    summarize(pass = max(pass, na.rm = TRUE), .groups = "drop") %>%
    pivot_wider(names_from = sample_id, values_from = pass, values_fill = 0) %>%
    column_to_rownames("TF") %>%
    as.matrix()

  # 3. Frequency Filter (min_conditions)
  tf_counts <- rowSums(df_matrix)
  keep_tfs <- names(tf_counts[tf_counts >= min_conditions])
  df_matrix <- df_matrix[keep_tfs, , drop = FALSE]

  if(nrow(df_matrix) == 0) {
    message("No TFs pass in ", min_conditions, " or more conditions.")
    return(NULL)
  }

  # 4. Robust Metadata Extraction (Fix for 'sample_id' error)
  # Instead of piping through separate(), we build the dataframe directly
  sample_names <- colnames(df_matrix)
  meta_list <- str_split(sample_names, "\\|")
  
  col_metadata <- data.frame(
    cell_line = sapply(meta_list, `[`, 1),
    dataset   = sapply(meta_list, `[`, 2),
    condition = sapply(meta_list, `[`, 3),
    row.names = sample_names, # This sets the rownames directly
    stringsAsFactors = FALSE
  )

  # 5. Create Heatmap Annotation
  ha_cols <- annotation_colors[names(annotation_colors) %in% colnames(col_metadata)]
  
  ha <- HeatmapAnnotation(
    df = col_metadata, 
    col = ha_cols,
    annotation_name_side = "left"
  )

  # 6. Render Heatmap
  Heatmap(df_matrix, 
          name = "Enriched", 
          col = c("0" = "white", "1" = "firebrick"),
          top_annotation = ha,
          rect_gp = gpar(col = "grey95", lwd = 0.5),
          column_title = paste0("HOMER Motif Enrichment\n(FDR < ", 
                                fdr_cutoff, ", Rank <= ", top_n_rank, 
                                ", Min Conditions >= ", min_conditions, ")"),
          cluster_rows = TRUE, 
          cluster_columns = TRUE,
          show_column_names = FALSE,
          row_names_gp = gpar(fontsize = 7))
}
}

# binary heatmap with rowAnnotation corresponding to matching significant motifsUp or motifsDown from multiome

plot_homer_heatmap_freq_highlight <- function(df, 
                               fdr_cutoff = 0.05, 
                               top_n_rank = 60, 
                               min_conditions = 1,
                               motif_list = NULL,
                               cluster_columns = TRUE,
                               annotation_colors = NULL) {
  
  # 1. Identify TFs passing initial thresholds
  passing_tfs <- df %>%
    filter(`q-value/FDR` < fdr_cutoff, Rank <= top_n_rank) %>%
    pull(TF) %>%
    unique() %>%
    sort()

  if(length(passing_tfs) == 0) {
    message("No motifs passed the initial thresholds (FDR/Rank).")
    return(NULL)
  }

  # 2. Build the binary matrix
  df_matrix <- df %>%
    filter(TF %in% passing_tfs) %>%
    mutate(
      sample_id = paste(cell_line, dataset, condition, sep = "|"),
      pass = if_else(`q-value/FDR` < fdr_cutoff & Rank <= top_n_rank, 1, 0)
    ) %>%
    group_by(TF, sample_id) %>%
    summarize(pass = max(pass, na.rm = TRUE), .groups = "drop") %>%
    pivot_wider(names_from = sample_id, values_from = pass, values_fill = 0) %>%
    column_to_rownames("TF") %>%
    as.matrix()

  # 3. Frequency Filtering
  tf_counts <- rowSums(df_matrix)
  keep_tfs <- names(tf_counts[tf_counts >= min_conditions])
  df_matrix <- df_matrix[keep_tfs, , drop = FALSE]

  if(nrow(df_matrix) == 0) {
    message("No TFs found that pass in ", min_conditions, " or more conditions.")
    return(NULL)
  }

  # --- NEW: HEATMAP SUMMARY PRINTING ---
  cat("==========================================\n")
  cat("      HOMER HEATMAP SUMMARY              \n")
  cat("==========================================\n")
  cat("Thresholds: FDR <", fdr_cutoff, "| Rank <=", top_n_rank, "\n")
  cat("Shared by:  >=", min_conditions, "conditions\n")
  cat("Final Size:", nrow(df_matrix), "TFs x", ncol(df_matrix), "Samples\n")
  
  # Print the top recurring TFs
  cat("\nTop 10 Most Frequent Motifs:\n")
  top_recurring <- sort(tf_counts[keep_tfs], decreasing = TRUE)
  print(head(top_recurring, 10))

  # Summary of the scalable highlights
  if (!is.null(motif_list)) {
    cat("\nNumber of motif_list Matches to df:\n")
    for (list_name in names(motif_list)) {
      matches <- intersect(rownames(df_matrix), motif_list[[list_name]])
      cat("- ", list_name, ": ", length(matches), "/", length(motif_list[[list_name]]), " found\n", sep="")
      if(length(matches) > 0) {
        cat("  Top matches:", paste(head(matches, 50), collapse = ", "), 
            ifelse(length(matches) > 50, "...", ""), "\n")
      }
    }
  }
  cat("==========================================\n\n")

  # 4. Construct Scalable Row Annotations (Highlights)
  row_ha <- NULL
  if (!is.null(motif_list)) {
    row_anno_df <- data.frame(row.names = rownames(df_matrix))
    row_anno_cols <- list()
    list_palette <- c("#4DAF4A", "#984EA3", "#FF7F00", "#FFFF33", "#A65628")
    
    for (i in seq_along(motif_list)) {
      list_name <- names(motif_list)[i]
      row_anno_df[[list_name]] <- ifelse(rownames(df_matrix) %in% motif_list[[i]], "Yes", "No")
      current_col <- list_palette[(i - 1) %% length(list_palette) + 1]
      row_anno_cols[[list_name]] <- c("Yes" = current_col, "No" = "white")
    }
    
    row_ha <- rowAnnotation(
      df = row_anno_df, col = row_anno_cols,
      simple_anno_size = unit(3, "mm"),
      show_annotation_name = TRUE, annotation_name_side = "top",
      annotation_name_gp = gpar(fontsize = 8)
    )
  }

  # 5. Metadata and Final Rendering
  sample_names <- colnames(df_matrix)
  meta_list <- str_split(sample_names, "\\|")
  col_metadata <- data.frame(
    cell_line = sapply(meta_list, `[`, 1),
    dataset   = sapply(meta_list, `[`, 2),
    condition = sapply(meta_list, `[`, 3),
    row.names = sample_names
  )

  ha_top <- HeatmapAnnotation(df = col_metadata, col = annotation_colors, annotation_name_side = "left")

  Heatmap(df_matrix, 
          name = "Significantly Enriched", 
          col = c("0" = "white", "1" = "black"),
          top_annotation = ha_top,
          right_annotation = row_ha, 
          row_names_side = "left",
          rect_gp = gpar(col = "grey95", lwd = 0.5),
          column_title = paste0("HOMER Motif Enrichment\n(FDR < ", 
                                fdr_cutoff, ", Rank <= ", top_n_rank, 
                                ", Min Conditions >= ", min_conditions, ")"),
          heatmap_legend_param = list( # here we rename the 0 and 1 to be readable
            title = "Significantly Enriched",
            at = c(0, 1),
            labels = c("False", "True")
          ),
          cluster_rows = TRUE, 
          cluster_columns = cluster_columns,
          show_column_names = FALSE,
          row_names_gp = gpar(fontsize = 8))
}




```
### Combine G361/G411 pairwise motifs with G523 motifs
```{r combine G361/G411/G523, eval = FALSE}
# Check column types side-by-side
types_final <- sapply(final_table, class)
types_g523  <- sapply(G523_final_table, class)

# Find which columns don't match
mismatches <- types_final[types_final != types_g523[names(types_final)]]
print(mismatches)

# Function to cast columns to match a reference table
align_types <- function(target_df, reference_df) {
  # Get names of columns present in both
  common_cols <- intersect(names(target_df), names(reference_df))
  
  for (col in common_cols) {
    ref_class <- class(reference_df[[col]])[1]
    
    # Cast target column to match reference class
    if (ref_class == "numeric") {
      target_df[[col]] <- as.numeric(target_df[[col]])
    } else if (ref_class == "character") {
      target_df[[col]] <- as.character(target_df[[col]])
    } else if (ref_class == "integer") {
      target_df[[col]] <- as.integer(target_df[[col]])
    }
  }
  return(target_df)
}

# Apply the alignment
G523_final_table <- align_types(G523_final_table, final_table)

# Combine the two tables
G361_G411_G523_final_table <- bind_rows(final_table, G523_final_table)

# Verify the combination
# This will show you the count of rows from each dataset/cell_line
G361_G411_G523_final_table %>%
  group_by(cell_line, dataset) %>%
  tally()

write_csv(G361_G411_G523_final_table, 
          here("data","bulkATAC","2026-04-21_G361_G411_G523_pairwise_HOMER_motifs.csv"),
                      col_names=TRUE)
```

## Compare to human multiome PIMO_up_status pairwise results
```{r binary heatmap of GSC and multiome highlighted}
# load multiome motifs
motifsUp_homer.df_cleaned <-read_csv(
  here(
    "data",
    "human_multiome",
    "motifEnrichment_homer_PIMO_up_status",
    "motifsUp_PIMO_up_status_PIMOup_vs_PIMOdown_homer.csv"
  )
)

motifsDown_homer.df_cleaned <-read_csv(
  here(
    "data",
    "human_multiome",
    "motifEnrichment_homer_PIMO_up_status",
    "motifsDown_PIMO_up_status_PIMOup_vs_PIMOdown_homer.csv"
  )
)
# 332 motifs for multiome, how many match prior to filtering

G361_G411_G523_final_table$TF %>% unique() %>% length # 302
motifsDown_homer.df_cleaned$TF_cleaned %>% unique() %>% length # 320

# compare the 302 vs 320 motifs
intersect(
  G361_G411_G523_final_table$TF %>% unique(),
  motifsDown_homer.df_cleaned$TF_cleaned %>% unique()
) %>% length # 281 of 320 multiome motifs are in 

# 320-281 = 39 missing motifs; can they be repaired
 # what is in the GSC only
setdiff(G361_G411_G523_final_table$TF %>% unique(),
        motifsDown_homer.df_cleaned$TF_cleaned %>% unique())
 # what is in the human multiome only
setdiff(motifsDown_homer.df_cleaned$TF_cleaned %>% unique(),
G361_G411_G523_final_table$TF %>% unique())

shared_sig_motifsUp <- intersect(
  G361_G411_G523_final_table$TF %>% unique(),
  motifsUp_homer.df_cleaned %>% filter(rank <= 60 & (mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned)
)

shared_sig_motifsDown <- intersect(
  G361_G411_G523_final_table$TF %>% unique(),
  motifsDown_homer.df_cleaned %>% filter(rank <= 60 & (mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned)
)

shared_sig_motifsUp %>% length # 49
shared_sig_motifsDown %>% length # 58
# which ones are not in? can we correct them?

# another round of cleaning
motifsUp_homer.df_cleaned_v2 <- homer_motif_cleaning_v2(motifsUp_homer.df_cleaned)
motifsDown_homer.df_cleaned_v2 <- homer_motif_cleaning_v2(motifsDown_homer.df_cleaned)

intersect(
  G361_G411_G523_final_table$TF %>% unique(),
  motifsDown_homer.df_cleaned_v2$TF_cleaned %>% unique()
) %>% length # 302 of 320 multiome motifs are in 

shared_sig_motifsUp <- intersect(
  G361_G411_G523_final_table$TF %>% unique(),
  motifsUp_homer.df_cleaned_v2 %>% filter(rank <= 60 & (mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned)
)

shared_sig_motifsDown <- intersect(
  G361_G411_G523_final_table$TF %>% unique(),
  motifsDown_homer.df_cleaned_v2 %>% filter(rank <= 60 & (mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned)
)

shared_sig_motifsUp %>% length # 58
shared_sig_motifsDown %>% length # 60

# testing threshold filters on multiome;
## 1. rank and mlog10Padj > -log10(0.05)
motifsUp_homer.df_cleaned_v2 %>% filter(rank <= 60 & (mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned) %>% length()
motifsDown_homer.df_cleaned_v2 %>% filter(rank <= 60 & (mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned) %>% length()
## 2. mlog10Padj > -log10(0.05) only
motifsUp_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned) %>% length() # 146
motifsDown_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned) %>% length() # 97

# what threshold did we use for G523 previously in ranked plot

hm <- plot_homer_heatmap_freq_highlight(G361_G411_G523_final_table_filtered, 
                                  fdr_cutoff = 0.05, 
                                  top_n_rank = 60,
                                  min_conditions = 6,
                                  motif_list = list(
                                  PIMOhigh_motifsUp = motifsUp_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned),
                                  PIMOhigh_motifsDown = motifsDown_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned) 
                                  ),
                                  annotation_colors = my_colors)
draw(hm, merge_legend = FALSE)
draw(hm, merge_legend = TRUE)
```

## Export PDF for binary heatmaps
```{r plot and export binary heatmaps, eval = FALSE}

# All three GSCs into one heatmap
pdf(here("data","bulkATAC",
         "2026-04-27_G361_G411_G523_combined_HOMER_motifs_binaryHeatmap.pdf"), 
    width = 10, height = 14)
plot_homer_heatmap(G361_G411_G523_final_table, fdr_cutoff = 0.05, top_n_rank = 10,
                                annotation_colors = my_colors)
plot_homer_heatmap(G361_G411_G523_final_table, fdr_cutoff = 0.05, top_n_rank = 25,
                                annotation_colors = my_colors)
plot_homer_heatmap(G361_G411_G523_final_table, fdr_cutoff = 0.05, top_n_rank = 60,
                                annotation_colors = my_colors)

# G361/G411 but with G523 edgeR only
G361_G411_G523_final_table_filtered_edgeR <- G361_G411_G523_final_table %>%
  filter(!str_detect(condition, "deseq2"))
plot_homer_heatmap(G361_G411_G523_final_table_filtered_edgeR, fdr_cutoff = 0.05, top_n_rank = 10,
                                annotation_colors = my_colors)
plot_homer_heatmap(G361_G411_G523_final_table_filtered_edgeR, fdr_cutoff = 0.05, top_n_rank = 25,
                                annotation_colors = my_colors)
plot_homer_heatmap(G361_G411_G523_final_table_filtered_edgeR, fdr_cutoff = 0.05, top_n_rank = 60,
                                annotation_colors = my_colors)
## with frequency filtering
plot_homer_heatmap_freq(G361_G411_G523_final_table_filtered_edgeR, fdr_cutoff = 0.05, top_n_rank = 60,
                                min_conditions = 1,
                                annotation_colors = my_colors)
plot_homer_heatmap_freq(G361_G411_G523_final_table_filtered_edgeR, fdr_cutoff = 0.05, top_n_rank = 60,
                                min_conditions = 4,
                                annotation_colors = my_colors)
plot_homer_heatmap_freq(G361_G411_G523_final_table_filtered_edgeR, fdr_cutoff = 0.05, top_n_rank = 60,
                                min_conditions = 6,
                                annotation_colors = my_colors)

# G361/G411 but with G523 deseq2 only
G361_G411_G523_final_table_filtered_deseq2 <- G361_G411_G523_final_table %>%
  filter(!str_detect(condition, "edgeR"))
plot_homer_heatmap(G361_G411_G523_final_table_filtered_deseq2, fdr_cutoff = 0.05, top_n_rank = 10,
                                annotation_colors = my_colors)
plot_homer_heatmap(G361_G411_G523_final_table_filtered_deseq2, fdr_cutoff = 0.05, top_n_rank = 25,
                                annotation_colors = my_colors)
plot_homer_heatmap(G361_G411_G523_final_table_filtered_deseq2, fdr_cutoff = 0.05, top_n_rank = 60,
                                annotation_colors = my_colors)
## with frequency filtering
plot_homer_heatmap_freq(G361_G411_G523_final_table_filtered_deseq2, fdr_cutoff = 0.05, top_n_rank = 60,
                                min_conditions = 1,
                                annotation_colors = my_colors)
plot_homer_heatmap_freq(G361_G411_G523_final_table_filtered_deseq2, fdr_cutoff = 0.05, top_n_rank = 60,
                                min_conditions = 4,
                                annotation_colors = my_colors)
plot_homer_heatmap_freq(G361_G411_G523_final_table_filtered_deseq2, fdr_cutoff = 0.05, top_n_rank = 60,
                                min_conditions = 6,
                                annotation_colors = my_colors)
dev.off()

# With rowAnnotation of human_multiome PIMOup_vs_PIMOdown motifs
pdf(here("data","bulkATAC",
         "2026-04-27_G361_G411_G523_combined_HOMER_motifs_vs_human_multiome_PIMOup_vs_PIMOdown_binaryHeatmap.pdf"), 
    width = 10, height = 14)

# G361/G411 but with G523 edgeR
  
plot_homer_heatmap_freq_highlight(G361_G411_G523_final_table_filtered_edgeR, 
                                  fdr_cutoff = 0.05, 
                                  top_n_rank = 60,
                                  min_conditions = 1,
                                  motif_list = list(
                                  PIMOhigh_motifsUp = motifsUp_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned),
                                  PIMOhigh_motifsDown = motifsDown_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned) 
                                  ),
                                  annotation_colors = my_colors) %>% 
  draw(merge_legend = TRUE)

plot_homer_heatmap_freq_highlight(G361_G411_G523_final_table_filtered_edgeR, 
                                  fdr_cutoff = 0.05, 
                                  top_n_rank = 60,
                                  min_conditions = 4,
                                  motif_list = list(
                                  PIMOhigh_motifsUp = motifsUp_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned),
                                  PIMOhigh_motifsDown = motifsDown_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned) 
                                  ),
                                  annotation_colors = my_colors) %>% 
  draw(merge_legend = TRUE)

plot_homer_heatmap_freq_highlight(G361_G411_G523_final_table_filtered_edgeR, 
                                  fdr_cutoff = 0.05, 
                                  top_n_rank = 60,
                                  min_conditions = 6,
                                  motif_list = list(
                                  PIMOhigh_motifsUp = motifsUp_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned),
                                  PIMOhigh_motifsDown = motifsDown_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned) 
                                  ),
                                  annotation_colors = my_colors) %>% 
  draw(merge_legend = TRUE)

# G361/G411 but with G523 deseq2
  
plot_homer_heatmap_freq_highlight(G361_G411_G523_final_table_filtered_deseq2, 
                                  fdr_cutoff = 0.05, 
                                  top_n_rank = 60,
                                  min_conditions = 1,
                                  motif_list = list(
                                  PIMOhigh_motifsUp = motifsUp_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned),
                                  PIMOhigh_motifsDown = motifsDown_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned) 
                                  ),
                                  annotation_colors = my_colors) %>% 
  draw(merge_legend = TRUE)

plot_homer_heatmap_freq_highlight(G361_G411_G523_final_table_filtered_deseq2, 
                                  fdr_cutoff = 0.05, 
                                  top_n_rank = 60,
                                  min_conditions = 4,
                                  motif_list = list(
                                  PIMOhigh_motifsUp = motifsUp_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned),
                                  PIMOhigh_motifsDown = motifsDown_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned) 
                                  ),
                                  annotation_colors = my_colors) %>% 
  draw(merge_legend = TRUE)

plot_homer_heatmap_freq_highlight(G361_G411_G523_final_table_filtered_deseq2, 
                                  fdr_cutoff = 0.05, 
                                  top_n_rank = 60,
                                  min_conditions = 6,
                                  motif_list = list(
                                  PIMOhigh_motifsUp = motifsUp_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned),
                                  PIMOhigh_motifsDown = motifsDown_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned) 
                                  ),
                                  annotation_colors = my_colors) %>% 
  draw(merge_legend = TRUE)

# G361/G411 only
plot_homer_heatmap_freq_highlight(final_table, 
                                  fdr_cutoff = 0.05, 
                                  top_n_rank = 60,
                                  min_conditions = 1,
                                  motif_list = list(
                                  PIMOhigh_motifsUp = motifsUp_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned),
                                  PIMOhigh_motifsDown = motifsDown_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned) 
                                  ),
                                  annotation_colors = my_colors) %>% 
  draw(merge_legend = TRUE)

plot_homer_heatmap_freq_highlight(final_table, 
                                  fdr_cutoff = 0.05, 
                                  top_n_rank = 60,
                                  min_conditions = 4,
                                  motif_list = list(
                                  PIMOhigh_motifsUp = motifsUp_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned),
                                  PIMOhigh_motifsDown = motifsDown_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned) 
                                  ),
                                  annotation_colors = my_colors) %>% 
  draw(merge_legend = TRUE)

plot_homer_heatmap_freq_highlight(final_table, 
                                  fdr_cutoff = 0.05, 
                                  top_n_rank = 60,
                                  min_conditions = 6,
                                  motif_list = list(
                                  PIMOhigh_motifsUp = motifsUp_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned),
                                  PIMOhigh_motifsDown = motifsDown_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned) 
                                  ),
                                  annotation_colors = my_colors) %>% 
  draw(merge_legend = TRUE)

# G523 only edgeR only
G523_filtered <- G523_final_table %>%
  filter(!str_detect(condition, "deseq2"))
plot_homer_heatmap_freq_highlight(G523_filtered, 
                                  fdr_cutoff = 0.05, 
                                  top_n_rank = 60,
                                  min_conditions = 1,
                                  motif_list = list(
                                  PIMOhigh_motifsUp = motifsUp_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned),
                                  PIMOhigh_motifsDown = motifsDown_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned) 
                                  ),
                                  annotation_colors = my_colors) %>% 
  draw(merge_legend = TRUE)

plot_homer_heatmap_freq_highlight(G523_filtered, 
                                  fdr_cutoff = 0.05, 
                                  top_n_rank = 60,
                                  min_conditions = 2,
                                  motif_list = list(
                                  PIMOhigh_motifsUp = motifsUp_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned),
                                  PIMOhigh_motifsDown = motifsDown_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned) 
                                  ),
                                  annotation_colors = my_colors) %>% 
  draw(merge_legend = TRUE)

plot_homer_heatmap_freq_highlight(G523_filtered, 
                                  fdr_cutoff = 0.05, 
                                  top_n_rank = 60,
                                  min_conditions = 4,
                                  motif_list = list(
                                  PIMOhigh_motifsUp = motifsUp_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned),
                                  PIMOhigh_motifsDown = motifsDown_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned) 
                                  ),
                                  annotation_colors = my_colors) %>% 
  draw(merge_legend = TRUE)

 plot_homer_heatmap_freq_highlight(G523_filtered, 
                                   fdr_cutoff = 0.05, 
                                   top_n_rank = 60,
                                   min_conditions = 2,
                                   motif_list = list(
                                   PIMOhigh_motifsUp = motifsUp_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned),
                                   PIMOhigh_motifsDown = motifsDown_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned) 
                                   ),
                                    cluster_columns = FALSE,
                                   annotation_colors = my_colors) %>% 
   draw(merge_legend = TRUE)

# G523 only deseq2 only
G523_filtered_deseq2 <- G523_final_table %>%
  filter(!str_detect(condition, "edgeR"))
plot_homer_heatmap_freq_highlight(G523_filtered_deseq2, 
                                  fdr_cutoff = 0.05, 
                                  top_n_rank = 60,
                                  min_conditions = 1,
                                  motif_list = list(
                                  PIMOhigh_motifsUp = motifsUp_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned),
                                  PIMOhigh_motifsDown = motifsDown_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned) 
                                  ),
                                  annotation_colors = my_colors) %>% 
  draw(merge_legend = TRUE)

plot_homer_heatmap_freq_highlight(G523_filtered_deseq2, 
                                  fdr_cutoff = 0.05, 
                                  top_n_rank = 60,
                                  min_conditions = 2,
                                  motif_list = list(
                                  PIMOhigh_motifsUp = motifsUp_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned),
                                  PIMOhigh_motifsDown = motifsDown_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned) 
                                  ),
                                  annotation_colors = my_colors) %>% 
  draw(merge_legend = TRUE)

plot_homer_heatmap_freq_highlight(G523_filtered_deseq2, 
                                  fdr_cutoff = 0.05, 
                                  top_n_rank = 60,
                                  min_conditions = 4,
                                  motif_list = list(
                                  PIMOhigh_motifsUp = motifsUp_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned),
                                  PIMOhigh_motifsDown = motifsDown_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned) 
                                  ),
                                  annotation_colors = my_colors) %>% 
  draw(merge_legend = TRUE)

 plot_homer_heatmap_freq_highlight(G523_filtered_deseq2, 
                                   fdr_cutoff = 0.05, 
                                   top_n_rank = 60,
                                   min_conditions = 2,
                                   motif_list = list(
                                   PIMOhigh_motifsUp = motifsUp_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned),
                                   PIMOhigh_motifsDown = motifsDown_homer.df_cleaned_v2 %>% filter((mlog10Padj > -log10(0.05))) %>% pull(TF_cleaned) 
                                   ),
                                    cluster_columns = FALSE,
                                   annotation_colors = my_colors) %>% 
   draw(merge_legend = TRUE)

dev.off()


# G361/G411 or G523 binary heatmap of pairwise conditions
pdf(here("data","bulkATAC",
         "2026-04-27_G361_G411_G523_HOMER_motifs_binaryHeatmap.pdf"), 
    width = 10, height = 14)
# G361 and G411
plot_homer_heatmap(final_table, fdr_cutoff = 0.05, top_n_rank = 10,
                                annotation_colors = my_colors)
plot_homer_heatmap(final_table, fdr_cutoff = 0.05, top_n_rank = 25,
                                annotation_colors = my_colors)
plot_homer_heatmap(final_table, fdr_cutoff = 0.05, top_n_rank = 60,
                                annotation_colors = my_colors)
plot_homer_heatmap(final_table, fdr_cutoff = 0.005, top_n_rank = 10,
                                annotation_colors = my_colors)
plot_homer_heatmap(final_table, fdr_cutoff = 0.005, top_n_rank = 25,
                                annotation_colors = my_colors)
plot_homer_heatmap(final_table, fdr_cutoff = 0.005, top_n_rank = 60,
                                annotation_colors = my_colors)
# G523 with deseq2 and edgeR
plot_homer_heatmap(G523_final_table, fdr_cutoff = 0.05, top_n_rank = 10,
                                annotation_colors = my_colors)
plot_homer_heatmap(G523_final_table, fdr_cutoff = 0.05, top_n_rank = 25,
                                annotation_colors = my_colors)
plot_homer_heatmap(G523_final_table, fdr_cutoff = 0.05, top_n_rank = 60,
                                annotation_colors = my_colors)
plot_homer_heatmap(G523_final_table, fdr_cutoff = 0.005, top_n_rank = 10,
                                annotation_colors = my_colors)
plot_homer_heatmap(G523_final_table, fdr_cutoff = 0.005, top_n_rank = 25,
                                annotation_colors = my_colors)
plot_homer_heatmap(G523_final_table, fdr_cutoff = 0.005, top_n_rank = 60,
                                annotation_colors = my_colors)

# G523 with edgeR only (used in supplemental table)
G523_filtered <- G523_final_table %>%
  filter(!str_detect(condition, "deseq2"))
plot_homer_heatmap(G523_filtered, fdr_cutoff = 0.05, top_n_rank = 10,
                                annotation_colors = my_colors)
plot_homer_heatmap(G523_filtered, fdr_cutoff = 0.05, top_n_rank = 25,
                                annotation_colors = my_colors)
plot_homer_heatmap(G523_filtered, fdr_cutoff = 0.05, top_n_rank = 60,
                                annotation_colors = my_colors)
plot_homer_heatmap(G523_filtered, fdr_cutoff = 0.005, top_n_rank = 10,
                                annotation_colors = my_colors)
plot_homer_heatmap(G523_filtered, fdr_cutoff = 0.005, top_n_rank = 25,
                                annotation_colors = my_colors)
plot_homer_heatmap(G523_filtered, fdr_cutoff = 0.005, top_n_rank = 60,
                                annotation_colors = my_colors)

# G523 with deseq2 only (used in supplemental table)
G523_filtered_deseq2 <- G523_final_table %>%
  filter(!str_detect(condition, "edgeR"))
plot_homer_heatmap(G523_filtered_deseq2, fdr_cutoff = 0.05, top_n_rank = 10,
                                annotation_colors = my_colors)
plot_homer_heatmap(G523_filtered_deseq2, fdr_cutoff = 0.05, top_n_rank = 25,
                                annotation_colors = my_colors)
plot_homer_heatmap(G523_filtered_deseq2, fdr_cutoff = 0.05, top_n_rank = 60,
                                annotation_colors = my_colors)
plot_homer_heatmap(G523_filtered_deseq2, fdr_cutoff = 0.005, top_n_rank = 10,
                                annotation_colors = my_colors)
plot_homer_heatmap(G523_filtered_deseq2, fdr_cutoff = 0.005, top_n_rank = 25,
                                annotation_colors = my_colors)
plot_homer_heatmap(G523_filtered_deseq2, fdr_cutoff = 0.005, top_n_rank = 60,
                                annotation_colors = my_colors)
dev.off()
```


# ChromVar Positive-TF Regulators Plot
```{r chromva positive TF regulators plot}
class(chromVarDev_pos_TF_reg)
colnames(chromVarDev_pos_TF_reg)
# convert DFrame to dataframe to tibble
chromVarDev_pos_TF_reg_df <- as_tibble(as.data.frame(chromVarDev_pos_TF_reg))
# remove _number suffix and Add a TF column
chromVarDev_pos_TF_reg_df$TF <- sub("_.*", "", chromVarDev_pos_TF_reg_df$MotifMatrix_homer_name) # remove the _number suffix # add a TF column


chromVarDev_pos_TF_reg_df


# filter based on ArchR default criteria (can also filter based on the TFRegulator == "YES")
chromVarDev_pos_TF_reg_df %>% 
  filter(cor > 0.5 & padj < 0.01 & maxDelta > quantile(maxDelta,0.75))

# quantile threshold used
quantile(chromVarDev_pos_TF_reg_df$maxDelta, 0.75) # 1.133545 

# volcano plot
chromVarDev_pos_TF_reg_df %>% 
  #filter(maxDelta >= quantile(maxDelta, 0.75, na.rm = TRUE)) %>% 
  ggplot(
    aes(x = cor, y = maxDelta, color = TFRegulator)
  ) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "black") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  geom_point() +
  # Manually set the colors here
  scale_color_manual(values = c("YES" = "red", "NO" = "grey")) +
  labs(
    x = "Correlation to Gene Score",
    y = "Mean TF Motif Delta",
    title = "Positive TF Regulators (Top Quartile)",
    subtitle = "Human Multiome: homer PIMO_up_status"
  ) + 
  theme_minimal()

# Volcano plot with labels
chromVarDev_pos_TF_reg_df %>% 
  ggplot(
    aes(x = cor, y = maxDelta, color = TFRegulator)
  ) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "black") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  geom_point() +
  # Labeling only the "YES" points
  geom_text_repel(
    data = . %>% filter(TFRegulator == "YES"),
    aes(label = GeneExpressionMatrix_name),
    size = 3,
    max.overlaps = Inf,      # Ensures all "YES" labels show up
    box.padding = 0.5,       # Adds space around labels
    point.padding = 0.3,     # Adds space between point and label
    show.legend = FALSE      # Prevents an 'a' from appearing in the legend
  ) +
  scale_color_manual(values = c("YES" = "red", "NO" = "grey70")) +
  labs(
    x = "Correlation to Gene Score",
    y = "Mean TF Motif Delta",
    title = "Positive TF Regulators (Top Quartile)",
    subtitle = "Human Multiome: homer PIMO_up_status"
  ) + 
  theme_minimal()



View(chromVarDev_pos_TF_reg_df)
```
## custom plotting function for Positive TF regulators
```{r custom function for postive TF regulators}
plot_tf_regulators <- function(data, 
                               cor_thresh = 0.5, 
                               padj_thresh = 0.01, 
                               delta_quantile = 0.75,
                               title = "Positive TF Regulators") {
  
  # 1. Calculate the specific maxDelta value for the chosen quantile
  cutoff_val <- quantile(data$maxDelta, delta_quantile, na.rm = TRUE)
  
  # 2. Create the TF_Regulator column and filter/label logic
  plot_df <- data %>%
    mutate(TF_Regulator = ifelse(
      cor > cor_thresh & 
      padj < padj_thresh & 
      maxDelta >= cutoff_val, 
      "YES", "NO"
    ))
  
  # 3. Generate the plot
  ggplot(plot_df, aes(x = cor, y = maxDelta, color = TF_Regulator)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
    geom_point(alpha = 0.7) +
    # Label only the "YES" regulators
    geom_text_repel(
      data = . %>% filter(TF_Regulator == "YES"),
      aes(label = GeneExpressionMatrix_name),
      size = 3.5,
      # 'direction = "both"' is default; "y" forces them to stay at their x-pos 
      # but move up/down. "x" forces them to move left/right.
      direction = "both", 
      # Increase force to push labels further away from the data/origin
      force = 3,
      # nudge_x can push all labels slightly right if your cor is > 0
      nudge_y = 0.5,
      box.padding = 0.5,
      point.padding = 0.3,
      max.overlaps = Inf,
      show.legend = FALSE
    ) +
    scale_color_manual(values = c("YES" = "black", "NO" = "grey80")) +
    labs(
      x = "Correlation to Gene Score",
      y = "Mean TF Motif Delta",
      title = title,
      subtitle = paste0("Thresholds: cor > ", cor_thresh, 
                        ", padj < ", padj_thresh, 
                        ", maxDelta > ", delta_quantile*100, "th percentile")
    ) + 
    theme_minimal()
}

# Using default values
plot_tf_regulators(chromVarDev_pos_TF_reg_df, title = "Human Multiome: homer PIMO_up_status")

# using Relaxed values
plot_tf_regulators(
  chromVarDev_pos_TF_reg_df, 
  cor_thresh = 0.4, 
  padj_thresh = 0.05, 
  delta_quantile = 0.75,
  title = "Human Multiome: homer PIMO_up_status"
)

```
## export to pdf
```{r export to pdf tf regulator}

pdf(here("Plots","2025-12-18_human_multiome_PIMO_up_status_PIMOup_vs_PIMOdown_Positive-TF-regulators-Motifs-Enriched_homer_labelled.pdf"))
# Using default values
plot_tf_regulators(chromVarDev_pos_TF_reg_df, title = "Human Multiome: homer PIMO_up_status PIMOup_vs_PIMOdown")

# using Relaxed values
plot_tf_regulators(
  chromVarDev_pos_TF_reg_df, 
  cor_thresh = 0.4, 
  padj_thresh = 0.05, 
  delta_quantile = 0.75,
  title = "Human Multiome: homer PIMO_up_status PIMOup_vs_PIMOdown"
)
dev.off()

```