---
title: "multiome_Supplemental_Tables"
author: "Ronald Wu"
date: "2026-03-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Install Packages
```{r install packages, eval = FALSE}
install.packages("writexl")
```

# Load Packages
```{r load packages}
library(tidyverse)
library(writexl)
library(here)
```

# Dataset description

Three specific datasets:
1. human_multiome
+ PeakSet
  + PIMO_up_status
  + k3_program (cNMF scPOS-MPs)
  + meta_module (nomura states)
+ motifEnrichment
  +
  +
  +
2. gaiti_multiome
3. mouse_multiome

table contents for each dataset
a) peak set or consensus peakset (matches MA plot numbers)
b) DA peaks for groupBy=PIMO_up_status
c) Motif enrichment for groupBy=PIMO_up_status

# Load peakSet (consensus peaks)
In ArchR, the generation of a consensus peak set is slightly different from traditional bulk ATAC-seq. Specifically, here an iterative overlap procedure is used to produce a reproducible peak set for each group contained in a specified groupBy column of interest in the metadata, such as clusters, PIMO_up_status, others.

In ArchR, only a single peakSet can be active, thus if multiple peak set are made for different groupBy of interest, addPeakSet() can be used to assign a new peakSet GRanges object to the ArchRProject object.
## human_multiome
```{r loading human_multiome peaksets}
# load GRanges peakSet:

# human_multiome:
  # 1. PIMO_up_status
peakSet_PIMO_up_status <-
  readRDS(
    here(
      "data",
      "human_multiome",
      "human_multiome_harmony_merged_malig_peak",
      "PeakCalls",
      "PeakSet_gr_PIMO_up_status.rds"
    )
  )
  # 2. k3_program
peakSet_k3_program <-
  readRDS(
    here(
      "data",
      "human_multiome",
      "human_multiome_hmmp_k3_program",
      "PeakCalls",
      "PeakSet_gr_k3_program.rds"
    )
  )
  # 3. meta_module (nomura states)
peakSet_meta_module <-
  readRDS(
    here(
      "data",
      "human_multiome",
      "human_multiome_hmmp_nomura",
      "PeakCalls",
      "PeakSet_gr_meta_module.rds"
    )
  )
```
## Gaiti_multiome
```{r loading Gaiti_multiome peaksets}
# gaiti_multome
# 1. PIMO_up_status
peakSet_PIMO_up_status_Gaiti <-
  readRDS(
    here(
      "data",
      "Gaiti_multiome",
      "Gaiti_multiome_harmony_merged_malig_peak",
      "PeakCalls",
      "PeakSet_gr_PIMO_up_status.rds"
    )
  )
# 2. PIMO_Region (PIMO_up_status with Region.Annotation)
peakSet_PIMO_Region_Gaiti <-
  readRDS(
    here(
      "data",
      "Gaiti_multiome",
      "Gaiti_multiome_harmony_merged_malig_peak",
      "PeakCalls",
      "PeakSet_gr_PIMO_Region.rds"
    )
  )
# 3. Region.annotation
peakSet_Region.annotation_Gaiti <-
  readRDS(
    here(
      "data",
      "Gaiti_multiome",
      "Gaiti_multiome_harmony_merged_malig_peak",
      "PeakCalls",
      "PeakSet_gr_Region.annotation.rds"
    )
  )
```
## mouse_multiome
```{r loading mouse_multiome peaksets}
# mouse_multiome

  # 1. PIMO_up_status
peakSet_PIMO_up_status_mouse <-
  readRDS(
    here(
      "data",
      "mouse_multiome",
      "mouse_multiome_harmony_merged_malig_peak_subset",
      "PeakCalls",
      "PeakSet_gr_PIMO_up_status.rds"
    )
  )
```

## EDA on peaksets
```{r eda on peaksets}
library(GenomicRanges)
gr_eda <-function(gr){
cat("total number of ranges: ", length(gr),"\n")        # Total number of ranges
cat("basic summary of object: ", summary(gr),"\n")       # Basic summary of the object
seqinfo(gr)   # Information about the chromosomes (lengths, circularity)

table(seqnames(gr))
}

# call function
gr_eda(peakSet_PIMO_up_status)

peakSet_PIMO_up_status
peakSet_k3_program
peakSet_meta_module
peakSet_PIMO_up_status_mouse
```
# Load marker peaks and marker test
## human_multiome:
```{r load markerPeaks/markersTest}
# For PIMOup vs PIMOdown (useGroups=PIMOup, bgdGroups = PIMOdown)
# SE
markersTest <- readRDS(
  here(
    "data",
    "human_multiome",
    "human_multiome_harmony_merged_malig_peak",
    "PeakCalls",
    "markersTest_PIMO_up_status_PIMOup_vs_PIMOdown.rds"
  )
)
# GR
markersTest_gr <- readRDS(
  here(
    "data",
    "human_multiome",
    "human_multiome_harmony_merged_malig_peak",
    "PeakCalls",
    "markersTest_GR_PIMO_up_status_PIMOup_vs_PIMOdown.rds"
  )
)
```


```{r eda on markerspeaks/markersTest}
# EDA on markerTest peaks
markersTest # dim: 179499 1
markersTest_gr$PIMOup # 178910

edaPeaks <- function(gr){
cat("number of ranges:", length(gr),"\n")

up <- subset(gr, FDR <= 0.1 & Log2FC >= 0.5) %>%
  length()
down <- subset(gr, FDR <= 0.1 & Log2FC <= -0.5) %>%
  length()
cat("threshold", "FDR <= 0.1 & Log2FC >= 0.5\n")
cat("number of up/open peaks:", up,"\n")
cat("threshold", "FDR <= 0.1 & Log2FC <= -0.5\n")
cat("number of down/closed peaks:", down,"\n")
}



edaPeaks(markersTest_gr$PIMOup)

# Differentially Accessible Peaks
sapply(assays(markersTest), function(x) sum(is.na(x)))
```
markersTest vs markersTest_gr:
179499 vs 178910
-why is 589 peak difference? FDR has 589 NAs!, which is removed in the gr.
## Gaiti_multiome
The markersTest files were originally made incorrectly with the peakCalling script, and run_05b_Gaiti_multiome_pairwiseTesting.sh was used to assign the proper peakSet to generate the correct pairwise results.
```{r load markerPeaks/markersTest gaiti}
# For PIMOup_TC_vs_PIMOdown_TC (useGroups=PIMOup_TC, bgdGroups = PIMOdown_TC)
# SE

markersTest_PIMOup_TC_vs_PIMOdown_TC <- readRDS(
  here(
    "data",
    "Gaiti_multiome",
    "Gaiti_multiome_harmony_merged_malig_peak",
    "PeakCalls",
    "2026-04-06",
    "markersTest_PIMO_Region_PIMOup_TC_vs_PIMOdown_TC.rds"
  )
)
# GR
markersTest_gr_PIMOup_TC_vs_PIMOdown_TC <- readRDS(
  here(
    "data",
    "Gaiti_multiome",
    "Gaiti_multiome_harmony_merged_malig_peak",
    "PeakCalls",
    "2026-04-06",
    "markersTest_GR_PIMO_Region_PIMOup_TC_vs_PIMOdown_TC.rds"
  )
)
sapply(assays(markersTest_PIMOup_TC_vs_PIMOdown_TC), function(x) sum(is.na(x)))
edaPeaks(markersTest_gr_PIMOup_TC_vs_PIMOdown_TC$PIMOup)
```

# Load motifEnrichment: motifsUp and motifsDown
## human_multiome:
### PIMO_up_status (pairwise)
```{r load PIMO_up_status (pairwise)}
motifsUp_homer.df_cleaned <- read_csv(
  here(
    "data",
    "human_multiome",
    "motifEnrichment_homer_PIMO_up_status",
    "motifsUp_PIMO_up_status_PIMOup_vs_PIMOdown_homer.csv"
  )
) %>% # additional cleaning for excel export
  dplyr::select(c(-TF, -TF_fixed)) %>%
  rename(TF = TF_cleaned) %>%
  relocate(TF, .before = 1)

motifsDown_homer.df_cleaned <- read_csv(
  here(
    "data",
    "human_multiome",
    "motifEnrichment_homer_PIMO_up_status",
    "motifsDown_PIMO_up_status_PIMOup_vs_PIMOdown_homer.csv"
  )
) %>% # additional cleaning for excel export
  dplyr::select(c(-TF, -TF_fixed)) %>%
  rename(TF = TF_cleaned) %>%
  relocate(TF, .before = 1)
```
### k3_program
Note: Extra column called `group` that has program_1, program_2, program_3
```{r load k3_program}

k3_motifsUp.df_cleaned <- read_csv(
  here(
    "data",
    "human_multiome",
    "human_multiome_hmmp_k3_program",
    "motifEnrichment_homer_k3_program",
    "motifsUp_k3program_homer.csv"
  )
) %>% # additional cleaning for excel export
  dplyr::select(c(-TF, -TF_fixed)) %>%
  rename(TF = TF_cleaned) %>%
  relocate(TF, .before = 1)

k3_motifsDown.df_cleaned <-
  read_csv(
  here(
    "data",
    "human_multiome",
    "human_multiome_hmmp_k3_program",
    "motifEnrichment_homer_k3_program",
    "motifsDown_k3program_homer.csv"
  )
) %>% # additional cleaning for excel export
  dplyr::select(c(-TF, -TF_fixed)) %>%
  rename(TF = TF_cleaned) %>%
  relocate(TF, .before = 1)
```

### meta_module (nomura states)
Note: Extra column called `group` that has all the 10nomura states
```{r load meta_module nomura states}

nomura_motifsUp.df_cleaned <- read_csv(
  here(
    "data",
    "human_multiome",
    "human_multiome_hmmp_nomura",
    "motifEnrichment_homer_meta_module",
    "motifsUp_meta_module_homer.csv"
  )
) %>% # additional cleaning for excel export
  dplyr::select(c(-TF, -TF_fixed)) %>%
  rename(TF = TF_cleaned) %>%
  relocate(TF, .before = 1)

nomura_motifsDown.df_cleaned <- read_csv(
  here(
    "data",
    "human_multiome",
    "human_multiome_hmmp_nomura",
    "motifEnrichment_homer_meta_module",
    "motifsDown_meta_module_homer.csv"
  )
) %>% # additional cleaning for excel export
  dplyr::select(c(-TF, -TF_fixed)) %>%
  rename(TF = TF_cleaned) %>%
  relocate(TF, .before = 1)
```

## Gaiti_multiome:
### PIMO_Region (pairwise)
```{r load PIMO_Region from Gaiti_multiome}

motifsUp_PIMOupTCvsPIMOdownTC.df_cleaned <- read_csv(
  here(
    "data",
    "Gaiti_multiome",
    "motifEnrichment_homer_PIMO_Region_PIMOup_TC_vs_PIMOdown_TC",
    "motifsUp_PIMORegion_PIMOupTCvsPIMOdownTC_homer.csv"
  )
) %>% # additional cleaning for excel export
  dplyr::select(c(-TF, -TF_fixed)) %>%
  rename(TF = TF_cleaned) %>%
  relocate(TF, .before = 1)

motifsDown_PIMOupTCvsPIMOdownTC.df_cleaned <- read_csv(
  here(
    "data",
    "Gaiti_multiome",
    "motifEnrichment_homer_PIMO_Region_PIMOup_TC_vs_PIMOdown_TC",
    "motifsDown_PIMORegion_PIMOupTCvsPIMOdownTC_homer.csv"
  )
) %>% # additional cleaning for excel export
  dplyr::select(c(-TF, -TF_fixed)) %>%
  rename(TF = TF_cleaned) %>%
  relocate(TF, .before = 1)
```

# peakSet to excel

```{r peakSet to excel sheet}
#' Export a list of GRanges objects and additional data frames to a single Excel file
#'
#' @param gr_list A named list of GRanges objects
#' @param extra_sheets A named list of standard data frames (optional)
#' @param file_path The path where the Excel file will be saved
#' @param workbook_description A plain text string explaining the contents of the Excel file
export_master_excel <- function(gr_list, extra_sheets = NULL, file_path, workbook_description = "Supplemental multiome data.") {
  
  # 1. Validate GRanges Input
  if (!is.list(gr_list) || !all(vapply(gr_list, inherits, logical(1), "GRanges"))) {
    stop("gr_list must be a list containing ONLY GRanges objects.")
  }
  
  # 2. Helper function to ensure lists have valid names
  ensure_names <- function(lst, prefix) {
    if (is.null(names(lst))) names(lst) <- paste0(prefix, seq_along(lst))
    empty_names <- names(lst) == ""
    if (any(empty_names)) names(lst)[empty_names] <- paste0(prefix, which(empty_names))
    return(lst)
  }
  gr_list <- ensure_names(gr_list, "GRanges_")
  
  # 3. Create Row 1: The Workbook Overview / Description
  desc_df <- data.frame(
    Sheet_Name = "ALL SHEETS",
    Data_Type = "Workbook Overview",
    Total_Rows_or_Ranges = NA,
    Total_Columns = NA,
    Key_Details = workbook_description,
    stringsAsFactors = FALSE
  )
  
  # 4. Create the Summary Rows for GRanges
  gr_summary_list <- lapply(names(gr_list), function(name) {
    gr <- gr_list[[name]]
    data.frame(
      Sheet_Name = name,
      Data_Type = "Genomic Ranges",
      Total_Rows_or_Ranges = length(gr),
      Total_Columns = ncol(mcols(gr)),
      Key_Details = paste0("Chrs: ", paste(head(seqlevelsInUse(gr), 5), collapse = ", "), 
                           " | Avg Width: ", round(mean(width(gr)), 2)),
      stringsAsFactors = FALSE
    )
  })
  gr_summary_df <- do.call(rbind, gr_summary_list)
  
  # 5. Create the Summary Rows for extra_sheets (if any)
  if (!is.null(extra_sheets)) {
    if (!is.list(extra_sheets) || !all(vapply(extra_sheets, is.data.frame, logical(1)))) {
      stop("extra_sheets must be a named list containing ONLY data frames.")
    }
    extra_sheets <- ensure_names(extra_sheets, "Sheet_")
    
    extra_summary_list <- lapply(names(extra_sheets), function(name) {
      df <- extra_sheets[[name]]
      data.frame(
        Sheet_Name = name,
        Data_Type = "Data Frame",
        Total_Rows_or_Ranges = nrow(df),
        Total_Columns = ncol(df),
        Key_Details = paste0("Cols: ", paste(head(colnames(df), 5), collapse = ", ")),
        stringsAsFactors = FALSE
      )
    })
    extra_summary_df <- do.call(rbind, extra_summary_list)
  } else {
    extra_summary_df <- NULL
  }
  
  # 6. Combine all summaries into one Master Summary DF
  master_summary_df <- rbind(desc_df, gr_summary_df, extra_summary_df)
  
  # 7. Convert all GRanges to standard Data Frames
  data_list <- lapply(gr_list, function(gr) {
    if (!is.null(names(gr))) {
      gr$range_name <- names(gr)
      names(gr) <- NULL
    }
    as.data.frame(gr)
  })
  
  # 8. Combine Summary, GRanges DFs, and Extra DFs into the final sheets list
  sheets_list <- c(list(Master_Summary = master_summary_df), data_list)
  if (!is.null(extra_sheets)) {
    sheets_list <- c(sheets_list, extra_sheets)
  }
  
  # 9. Safety check: Excel limits sheet names to 31 characters
  names(sheets_list) <- substr(names(sheets_list), 1, 31)
  
  # 10. Write to Excel
  write_xlsx(sheets_list, path = file_path)
  
  # Calculate extra sheets length safely for the message
  extra_len <- ifelse(is.null(extra_sheets), 0, length(extra_sheets))
  message(sprintf("Successfully exported %d GRanges objects and %d additional sheets to %s", 
                  length(gr_list), extra_len, file_path))
}
```
## export supplemental table 5 human multiome
```{r useage, eval = FALSE}
# optional: Package the motifs into a named list

motif_sheets <- list(
  "PIMO_up_status_motifsUp" = motifsUp_homer.df_cleaned,
  "PIMO_up_status_motifsDown" = motifsDown_homer.df_cleaned,
  "k3_program_motifsUp" = k3_motifsUp.df_cleaned,
  "k3_program_motifsDown" = k3_motifsDown.df_cleaned,
  "nomura_motifsUp" = nomura_motifsUp.df_cleaned,
  "nomura_motifsDown" = nomura_motifsUp.df_cleaned
)

# 3. Package your GRanges objects into a named list 
# (Using your existing PIMO_up_status as the base)
my_gr_list <- list(
  "PIMO_up_status_peakSet" = peakSet_PIMO_up_status,
  "PIMO_up_status_DAR_up" = subset(markersTest_gr$PIMOup, FDR <= 0.1 &
                                     Log2FC >= 0.5),
  "PIMO_up_status_DAR_down" = subset(markersTest_gr$PIMOup, FDR <= 0.1 &
                                       Log2FC <= -0.5),
  "k3_program_peakSet" = peakSet_k3_program,
  "nomura_peakSet" = peakSet_meta_module
)

# 4. Export the consolidated workbooks

# human_multiome:
export_master_excel(
  gr_list = my_gr_list,
  extra_sheets = motif_sheets,
  file_path = here(
    "figures", 
    "supplemental_table", 
    "Supplementary_table_5.xlsx"
  ),
  workbook_description = "This workbook contains consensus peak sets and enriched motifs for the human multiome dataset. The associated tables were generated for each group of interest: PIMO_up_status, k3_program, and nomura"
)



```

## export supplementary table 6 Gaiti multiome
```{r useage supplementary table 6, eval = FALSE}
motif_sheets <- list(
  "PIMOupTCvsPIMOdownTCmotifsUp" = motifsUp_PIMOupTCvsPIMOdownTC.df_cleaned,
  "PIMOupTCvsPIMOdownTCmotifsDown" = motifsDown_PIMOupTCvsPIMOdownTC.df_cleaned
)

# gaiti_multiome
export_master_excel(
  gr_list = list(
    "PIMO_Region_peakSet" = peakSet_PIMO_Region_Gaiti,
    "PIMO_Region_DAR_up" = subset(markersTest_gr_PIMOup_TC_vs_PIMOdown_TC$PIMOup, FDR <= 0.1 &
                                     Log2FC >= 0.5),
    "PIMO_Region_DAR_down" = subset(markersTest_gr_PIMOup_TC_vs_PIMOdown_TC$PIMOup, FDR <= 0.1 &
                                       Log2FC <= -0.5),
    "PIMO_up_status_peakSet" = peakSet_PIMO_up_status_Gaiti,
    "Region.annotation_peakSet" = peakSet_Region.annotation_Gaiti
  ),
  extra_sheets = motif_sheets,
  file_path = here(
    "figures", 
    "supplemental_table", 
    "Supplementary_table_6.xlsx"
  ),
  workbook_description = "This workbook contains consensus peak sets and enriched motifs for the Wu et al dataset (Gaiti multiome). The associated tables were generated for each group of interest: PIMO_Region, PIMO_up_status, and Region.annotation"
)

```