---
title: "multiome_Supplemental_Tables"
author: "Ronald Wu"
date: "2026-03-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Install Packages
```{r install packages, eval = FALSE}
install.packages("writexl")
```

# Load Packages
```{r load packages}
library(tidyverse)
library(writexl)
library(here)
```

# Dataset description

Three specific datasets:
1. human_multiome
+ PeakSet
  + PIMO_up_status
  + k3_program (cNMF scPOS-MPs)
  + meta_module (nomura states)
+ motifEnrichment
  +
  +
  +
2. gaiti_multiome
3. mouse_multiome

table contents for each dataset
a) peak set or consensus peakset (matches MA plot numbers)
b) DA peaks for groupBy=PIMO_up_status
c) Motif enrichment for groupBy=PIMO_up_status

# Load peakSet (consensus peaks)
In ArchR, the generation of a consensus peak set is slightly different from traditional bulk ATAC-seq. Specifically, here an iterative overlap procedure is used to produce a reproducible peak set for each group contained in a specified groupBy column of interest in the metadata, such as clusters, PIMO_up_status, others.

In ArchR, only a single peakSet can be active, thus if multiple peak set are made for different groupBy of interest, addPeakSet() can be used to assign a new peakSet GRanges object to the ArchRProject object.
## human_multiome
```{r loading human_multiome peaksets}
# load GRanges peakSet:

# human_multiome:
  # 1. PIMO_up_status
peakSet_PIMO_up_status <-
  readRDS(
    here(
      "data",
      "human_multiome",
      "human_multiome_harmony_merged_malig_peak",
      "PeakCalls",
      "PeakSet_gr_PIMO_up_status.rds"
    )
  )
  # 2. k3_program
peakSet_k3_program <-
  readRDS(
    here(
      "data",
      "human_multiome",
      "human_multiome_hmmp_k3_program",
      "PeakCalls",
      "PeakSet_gr_k3_program.rds"
    )
  )
  # 3. meta_module (nomura states)
peakSet_meta_module <-
  readRDS(
    here(
      "data",
      "human_multiome",
      "human_multiome_hmmp_nomura",
      "PeakCalls",
      "PeakSet_gr_meta_module.rds"
    )
  )
```
## Gaiti_multiome
```{r loading Gaiti_multiome peaksets}
# gaiti_multome
# 1. PIMO_up_status
peakSet_PIMO_up_status_Gaiti <-
  readRDS(
    here(
      "data",
      "Gaiti_multiome",
      "Gaiti_multiome_harmony_merged_malig_peak",
      "PeakCalls",
      "PeakSet_gr_PIMO_up_status.rds"
    )
  )
# 2. PIMO_Region (PIMO_up_status with Region.Annotation)
peakSet_PIMO_Region_Gaiti <-
  readRDS(
    here(
      "data",
      "Gaiti_multiome",
      "Gaiti_multiome_harmony_merged_malig_peak",
      "PeakCalls",
      "PeakSet_gr_PIMO_Region.rds"
    )
  )
# 3. Region.annotation
peakSet_Region.annotation_Gaiti <-
  readRDS(
    here(
      "data",
      "Gaiti_multiome",
      "Gaiti_multiome_harmony_merged_malig_peak",
      "PeakCalls",
      "PeakSet_gr_Region.annotation.rds"
    )
  )
```
## mouse_multiome
```{r loading mouse_multiome peaksets}
# mouse_multiome

  # 1. PIMO_up_status
peakSet_PIMO_up_status_mouse <-
  readRDS(
    here(
      "data",
      "mouse_multiome",
      "mouse_multiome_harmony_merged_malig_peak_subset",
      "PeakCalls",
      "PeakSet_gr_PIMO_up_status.rds"
    )
  )
```

## EDA on peaksets
```{r eda on peaksets}
library(GenomicRanges)
gr_eda <-function(gr){
cat("total number of ranges: ", length(gr),"\n")        # Total number of ranges
cat("basic summary of object: ", summary(gr),"\n")       # Basic summary of the object
seqinfo(gr)   # Information about the chromosomes (lengths, circularity)

table(seqnames(gr))
}

# call function
gr_eda(peakSet_PIMO_up_status)

peakSet_PIMO_up_status
peakSet_k3_program
peakSet_meta_module
peakSet_PIMO_up_status_mouse
```
# Load marker peaks and marker test
## human_multiome:
```{r load markerPeaks/markersTest}
# For PIMOup vs PIMOdown (useGroups=PIMOup, bgdGroups = PIMOdown)
# SE
markersTest <- readRDS(
  here(
    "data",
    "human_multiome",
    "human_multiome_harmony_merged_malig_peak",
    "PeakCalls",
    "markersTest_PIMO_up_status_PIMOup_vs_PIMOdown.rds"
  )
)
# GR
markersTest_gr <- readRDS(
  here(
    "data",
    "human_multiome",
    "human_multiome_harmony_merged_malig_peak",
    "PeakCalls",
    "markersTest_GR_PIMO_up_status_PIMOup_vs_PIMOdown.rds"
  )
)

# For k3_program: (non-pairwise testing; each program vs all others)

hmmp_k3_markersPeaks <- readRDS(
  here(
    "data",
    "human_multiome",
    "human_multiome_hmmp_k3_program",
    "peakCalls",
    "markersPeaks_k3_program.rds"
  )
)

# GR
hmmp_k3_markersPeaks_GR <- readRDS(
  here(
    "data",
    "human_multiome",
    "human_multiome_hmmp_k3_program",
    "peakCalls",
    "markersPeaks_GR_k3_program.rds"
  )
)
```

## Gaiti_multiome
The markersTest files were originally made incorrectly with the peakCalling script, and run_05b_Gaiti_multiome_pairwiseTesting.sh was used to assign the proper peakSet to generate the correct pairwise results.
```{r load markerPeaks/markersTest gaiti}
# For PIMOup_TC_vs_PIMOdown_TC (useGroups=PIMOup_TC, bgdGroups = PIMOdown_TC)
# SE

markersTest_PIMOup_TC_vs_PIMOdown_TC <- readRDS(
  here(
    "data",
    "Gaiti_multiome",
    "Gaiti_multiome_harmony_merged_malig_peak",
    "PeakCalls",
    "2026-04-06",
    "markersTest_PIMO_Region_PIMOup_TC_vs_PIMOdown_TC.rds"
  )
)
# GR
markersTest_gr_PIMOup_TC_vs_PIMOdown_TC <- readRDS(
  here(
    "data",
    "Gaiti_multiome",
    "Gaiti_multiome_harmony_merged_malig_peak",
    "PeakCalls",
    "2026-04-06",
    "markersTest_GR_PIMO_Region_PIMOup_TC_vs_PIMOdown_TC.rds"
  )
)
sapply(assays(markersTest_PIMOup_TC_vs_PIMOdown_TC), function(x) sum(is.na(x)))
edaPeaks(markersTest_gr_PIMOup_TC_vs_PIMOdown_TC$PIMOup)


# For PIMOup_TC_vs_PIMOdown_TC (useGroups=PIMOup_TC, bgdGroups = PIMOdown_TC)
# GR
markersTest_gr_PIMOup_EB_vs_PIMOdown_EB <- readRDS(
  here(
    "data",
    "Gaiti_multiome",
    "Gaiti_multiome_harmony_merged_malig_peak",
    "PeakCalls",
    "2026-04-06",
    "markersTest_GR_PIMO_Region_PIMOup_EB_vs_PIMOdown_EB.rds"
  )
)
```

## eda on marker peaks or marker test
```{r eda on markerspeaks/markersTest}
# EDA on markerTest peaks
markersTest # dim: 179499 1
markersTest_gr$PIMOup # 178910

edaPeaks <- function(gr){
cat("number of ranges:", length(gr),"\n")

up <- subset(gr, FDR <= 0.1 & Log2FC >= 0.5) %>%
  length()
down <- subset(gr, FDR <= 0.1 & Log2FC <= -0.5) %>%
  length()
cat("threshold", "FDR <= 0.1 & Log2FC >= 0.5\n")
cat("number of up/open peaks:", up,"\n")
cat("threshold", "FDR <= 0.1 & Log2FC <= -0.5\n")
cat("number of down/closed peaks:", down,"\n")
}



edaPeaks(markersTest_gr$PIMOup)

# Differentially Accessible Peaks
sapply(assays(markersTest), function(x) sum(is.na(x)))

# EDA on markerPeaks from k3_program
hmmp_k3_markersPeaks # dim: 58477 3
hmmp_k3_markersPeaks_GR$program_1
  
edaPeaks(hmmp_k3_markersPeaks_GR$program_1) 
edaPeaks(hmmp_k3_markersPeaks_GR$program_2)
edaPeaks(hmmp_k3_markersPeaks_GR$program_3)

# EDA on markerTest for PIMO_region
edaPeaks(markersTest_gr_PIMOup_TC_vs_PIMOdown_TC$PIMOup_TC)
edaPeaks(markersTest_gr_PIMOup_EB_vs_PIMOdown_EB$PIMOup_EB)
```
markersTest vs markersTest_gr:
179499 vs 178910
-why is 589 peak difference? FDR has 589 NAs!, which is removed in the gr.

# Load motifEnrichment: motifsUp and motifsDown
## human_multiome:
### PIMO_up_status (pairwise)
```{r load PIMO_up_status (pairwise)}
motifsUp_homer.df_cleaned <- read_csv(
  here(
    "data",
    "human_multiome",
    "motifEnrichment_homer_PIMO_up_status",
    "motifsUp_PIMO_up_status_PIMOup_vs_PIMOdown_homer.csv"
  )
) %>% # rename and move the cleaned column to the start and keep only relevent columns
dplyr::select(TF = TF_cleaned, mlog10Padj, rank, TF_family)

motifsDown_homer.df_cleaned <- read_csv(
  here(
    "data",
    "human_multiome",
    "motifEnrichment_homer_PIMO_up_status",
    "motifsDown_PIMO_up_status_PIMOup_vs_PIMOdown_homer.csv"
  )
) %>% # rename and move the cleaned column to the start and keep only relevent columns
dplyr::select(TF = TF_cleaned, mlog10Padj, rank, TF_family)
```
### k3_program
Note: Extra column called `group` that has program_1, program_2, program_3
```{r load k3_program}

k3_motifsUp.df_cleaned <- read_csv(
  here(
    "data",
    "human_multiome",
    "human_multiome_hmmp_k3_program",
    "motifEnrichment_homer_k3_program",
    "motifsUp_k3program_homer.csv"
  )
) %>% # rename and move the cleaned column to the start and keep only relevent columns
dplyr::select(TF = TF_cleaned, group, mlog10Padj, rank, TF_family)

k3_motifsDown.df_cleaned <-
  read_csv(
  here(
    "data",
    "human_multiome",
    "human_multiome_hmmp_k3_program",
    "motifEnrichment_homer_k3_program",
    "motifsDown_k3program_homer.csv"
  )
) %>% # rename and move the cleaned column to the start and keep only relevent columns
dplyr::select(TF = TF_cleaned, group, mlog10Padj, rank, TF_family)
```

### meta_module (nomura states)
Note: Extra column called `group` that has all the 10nomura states
```{r load meta_module nomura states}

nomura_motifsUp.df_cleaned <- read_csv(
  here(
    "data",
    "human_multiome",
    "human_multiome_hmmp_nomura",
    "motifEnrichment_homer_meta_module",
    "motifsUp_meta_module_homer.csv"
  )
) %>% # rename and move the cleaned column to the start and keep only relevent columns
dplyr::select(TF = TF_cleaned, group, mlog10Padj, rank, TF_family)

nomura_motifsDown.df_cleaned <- read_csv(
  here(
    "data",
    "human_multiome",
    "human_multiome_hmmp_nomura",
    "motifEnrichment_homer_meta_module",
    "motifsDown_meta_module_homer.csv"
  )
) %>% # rename and move the cleaned column to the start and keep only relevent columns
dplyr::select(TF = TF_cleaned, group, mlog10Padj, rank, TF_family)
```

## Gaiti_multiome:
### PIMO_Region (pairwise)
```{r load PIMO_Region from Gaiti_multiome}

motifsUp_PIMOupTCvsPIMOdownTC.df_cleaned <- read_csv(
  here(
    "data",
    "Gaiti_multiome",
    "motifEnrichment_homer_PIMO_Region_PIMOup_TC_vs_PIMOdown_TC",
    "motifsUp_PIMORegion_PIMOupTCvsPIMOdownTC_homer.csv"
  )
) %>% # rename and move the cleaned column to the start and keep only relevent columns
dplyr::select(TF = TF_cleaned, mlog10Padj, rank, TF_family)

motifsDown_PIMOupTCvsPIMOdownTC.df_cleaned <- read_csv(
  here(
    "data",
    "Gaiti_multiome",
    "motifEnrichment_homer_PIMO_Region_PIMOup_TC_vs_PIMOdown_TC",
    "motifsDown_PIMORegion_PIMOupTCvsPIMOdownTC_homer.csv"
  )
) %>% # rename and move the cleaned column to the start and keep only relevent columns
dplyr::select(TF = TF_cleaned, mlog10Padj, rank, TF_family)

motifsUp_PIMOupEBvsPIMOdownEB.df_cleaned <- read_csv(
  here(
    "data",
    "Gaiti_multiome",
    "motifEnrichment_homer_PIMO_Region_PIMOup_EB_vs_PIMOdown_EB",
    "motifsUp_PIMORegion_PIMOupEBvsPIMOdownEB_homer.csv"
  )
) %>% # rename and move the cleaned column to the start and keep only relevent columns
dplyr::select(TF = TF_cleaned, mlog10Padj, rank, TF_family)

motifsDown_PIMOupEBvsPIMOdownEB.df_cleaned <- read_csv(
  here(
    "data",
    "Gaiti_multiome",
    "motifEnrichment_homer_PIMO_Region_PIMOup_EB_vs_PIMOdown_EB",
    "motifsDown_PIMORegion_PIMOupEBvsPIMOdownEB_homer.csv"
  )
) %>% # rename and move the cleaned column to the start and keep only relevent columns
dplyr::select(TF = TF_cleaned, mlog10Padj, rank, TF_family)
```

# peakSet to excel

```{r peakSet to excel sheet}
#' Export a list of GRanges objects and additional data frames to a single Excel file
#'
#' @param gr_list A named list of GRanges objects
#' @param extra_sheets A named list of standard data frames (optional)
#' @param file_path The path where the Excel file will be saved
#' @param workbook_description A plain text string explaining the contents of the Excel file
export_master_excel <- function(gr_list, extra_sheets = NULL, file_path, workbook_description = "Supplemental multiome data.") {
  
  # 1. Validate GRanges Input
  if (!is.list(gr_list) || !all(vapply(gr_list, inherits, logical(1), "GRanges"))) {
    stop("gr_list must be a list containing ONLY GRanges objects.")
  }
  
  # 2. Helper function to ensure lists have valid names
  ensure_names <- function(lst, prefix) {
    if (is.null(names(lst))) names(lst) <- paste0(prefix, seq_along(lst))
    empty_names <- names(lst) == ""
    if (any(empty_names)) names(lst)[empty_names] <- paste0(prefix, which(empty_names))
    return(lst)
  }
  gr_list <- ensure_names(gr_list, "GRanges_")
  
  # 3. Create Row 1: The Workbook Overview / Description
  desc_df <- data.frame(
    Sheet_Name = "ALL SHEETS",
    Data_Type = "Workbook Overview",
    Total_Rows_or_Ranges = NA,
    Total_Columns = NA,
    Key_Details = workbook_description,
    stringsAsFactors = FALSE
  )
  
  # 4. Create the Summary Rows for GRanges
  gr_summary_list <- lapply(names(gr_list), function(name) {
    gr <- gr_list[[name]]
    data.frame(
      Sheet_Name = name,
      Data_Type = "Genomic Ranges",
      Total_Rows_or_Ranges = length(gr),
      Total_Columns = ncol(mcols(gr)),
      Key_Details = paste0("Chrs: ", paste(head(seqlevelsInUse(gr), 5), collapse = ", "), 
                           " | Avg Width: ", round(mean(width(gr)), 2)),
      stringsAsFactors = FALSE
    )
  })
  gr_summary_df <- do.call(rbind, gr_summary_list)
  
  # 5. Create the Summary Rows for extra_sheets (if any)
  if (!is.null(extra_sheets)) {
    if (!is.list(extra_sheets) || !all(vapply(extra_sheets, is.data.frame, logical(1)))) {
      stop("extra_sheets must be a named list containing ONLY data frames.")
    }
    extra_sheets <- ensure_names(extra_sheets, "Sheet_")
    
    extra_summary_list <- lapply(names(extra_sheets), function(name) {
      df <- extra_sheets[[name]]
      data.frame(
        Sheet_Name = name,
        Data_Type = "Data Frame",
        Total_Rows_or_Ranges = nrow(df),
        Total_Columns = ncol(df),
        Key_Details = paste0("Cols: ", paste(head(colnames(df), 5), collapse = ", ")),
        stringsAsFactors = FALSE
      )
    })
    extra_summary_df <- do.call(rbind, extra_summary_list)
  } else {
    extra_summary_df <- NULL
  }
  
  # 6. Combine all summaries into one Master Summary DF
  master_summary_df <- rbind(desc_df, gr_summary_df, extra_summary_df)
  
  # 7. Convert all GRanges to standard Data Frames
  data_list <- lapply(gr_list, function(gr) {
    if (!is.null(names(gr))) {
      gr$range_name <- names(gr)
      names(gr) <- NULL
    }
    as.data.frame(gr)
  })
  
  # 8. Combine Summary, GRanges DFs, and Extra DFs into the final sheets list
  sheets_list <- c(list(Master_Summary = master_summary_df), data_list)
  if (!is.null(extra_sheets)) {
    sheets_list <- c(sheets_list, extra_sheets)
  }
  
  # 9. Safety check: Excel limits sheet names to 31 characters
  names(sheets_list) <- substr(names(sheets_list), 1, 31)
  
  # 10. Write to Excel
  write_xlsx(sheets_list, path = file_path)
  
  # Calculate extra sheets length safely for the message
  extra_len <- ifelse(is.null(extra_sheets), 0, length(extra_sheets))
  message(sprintf("Successfully exported %d GRanges objects and %d additional sheets to %s", 
                  length(gr_list), extra_len, file_path))
}
```

## export supplemental table 9 human multiome
```{r useage, eval = FALSE}
# optional: Package the motifs into a named list

motif_sheets <- list(
  "PIMO_up_status_motifsUp" = motifsUp_homer.df_cleaned,
  "PIMO_up_status_motifsDown" = motifsDown_homer.df_cleaned,
  "k3_program_motifsUp" = k3_motifsUp.df_cleaned,
  "k3_program_motifsDown" = k3_motifsDown.df_cleaned
  #"nomura_motifsUp" = nomura_motifsUp.df_cleaned,
  #"nomura_motifsDown" = nomura_motifsUp.df_cleaned
)

# Package GRanges objects into a named list 

my_gr_list <- list(
  "PIMO_up_status_peakSet" = peakSet_PIMO_up_status,
  "PIMO_up_status_DAR_up" = subset(markersTest_gr$PIMOup, FDR <= 0.1 &
                                     Log2FC >= 0.5),
  "PIMO_up_status_DAR_down" = subset(markersTest_gr$PIMOup, FDR <= 0.1 &
                                       Log2FC <= -0.5),
  "k3_program_peakSet" = peakSet_k3_program,
  "k3_program_1_DARup" = subset(hmmp_k3_markersPeaks_GR$program_1, 
                                FDR<=0.1 & Log2FC >= 0.5),
  "k3_program_1_DARdown" = subset(hmmp_k3_markersPeaks_GR$program_1, 
                                FDR<=0.1 & Log2FC <= -0.5),
  "k3_program_2_DARup" = subset(hmmp_k3_markersPeaks_GR$program_2, 
                                FDR<=0.1 & Log2FC >= 0.5),
  "k3_program_2_DARdown" = subset(hmmp_k3_markersPeaks_GR$program_2, 
                                FDR<=0.1 & Log2FC <= -0.5),
  "k3_program_3_DARup" = subset(hmmp_k3_markersPeaks_GR$program_3, 
                                FDR<=0.1 & Log2FC >= 0.5),
  "k3_program_3_DARdown" = subset(hmmp_k3_markersPeaks_GR$program_3, 
                                FDR<=0.1 & Log2FC <= -0.5)
# "nomura_peakSet" = peakSet_meta_module
)

# Export the consolidated workbooks

# human_multiome:
export_master_excel(
  gr_list = my_gr_list,
  extra_sheets = motif_sheets,
  file_path = here(
    "figures", 
    "supplemental_table", 
    "Supplementary_table_9.xlsx"
  ),
  workbook_description = "This workbook contains consensus peak sets, differentially accessible regions (DARs), and enriched motifs for the human multiome dataset. The associated tables were generated for each group of interest: PIMO_up_status and k3_program. For k3_program, program_1, program_2, and program_3 refer to the EGFR/HIF(RTK), Connective and Reactive programs respectively as mentioned in the main text. For DARs, only significant peak regions are shown if any."
)

```

## export supplementary table 10 Gaiti multiome
```{r useage supplementary table 10, eval = FALSE}
motif_sheets <- list(
  "PIMOupTCvsPIMOdownTCmotifsUp" = motifsUp_PIMOupTCvsPIMOdownTC.df_cleaned,
  "PIMOupTCvsPIMOdownTCmotifsDown" = motifsDown_PIMOupTCvsPIMOdownTC.df_cleaned,
  "PIMOupEBvsPIMOdownEBmotifsUp" = motifsUp_PIMOupEBvsPIMOdownEB.df_cleaned,
  "PIMOupEBvsPIMOdownEBmotifsDown" = motifsDown_PIMOupEBvsPIMOdownEB.df_cleaned
)

# gaiti_multiome
export_master_excel(
  gr_list = list(
    "PIMO_Region_peakSet" = peakSet_PIMO_Region_Gaiti,
    "PIMOupTCvsPIMOdownTC_DAR_up" = subset(markersTest_gr_PIMOup_TC_vs_PIMOdown_TC$PIMOup_TC, FDR <= 0.1 &
                                     Log2FC >= 0.5),
    "PIMOupTCvsPIMOdownTC_DAR_down" = subset(markersTest_gr_PIMOup_TC_vs_PIMOdown_TC$PIMOup_TC, FDR <= 0.1 &
                                       Log2FC <= -0.5),
    "PIMOupEBvsPIMOdownEB_DAR_up" = subset(markersTest_gr_PIMOup_EB_vs_PIMOdown_EB$PIMOup_EB, FDR <= 0.1 &
                                     Log2FC >= 0.5),
    "PIMOupEBvsPIMOdownEB_DAR_down" = subset(markersTest_gr_PIMOup_EB_vs_PIMOdown_EB$PIMOup_EB, FDR <= 0.1 &
                                       Log2FC <= -0.5),
    "PIMO_up_status_peakSet" = peakSet_PIMO_up_status_Gaiti,
    "Region.annotation_peakSet" = peakSet_Region.annotation_Gaiti
  ),
  extra_sheets = motif_sheets,
  file_path = here(
    "figures", 
    "supplemental_table", 
    "Supplementary_table_10.xlsx"
  ),
  workbook_description = "This workbook contains consensus peak sets, differentially accessible regions (DARs), and enriched motifs for the Wu et al dataset (Gaiti multiome). The associated tables were generated for each group of interest: PIMO_Region, PIMO_up_status, and Region.annotation. The PIMO_Region group is made combining PIMO_up_status and Region.annotation. For PIMO_region DARs, only the significant peak regions are shown for the indicated pairwise comparison. PT region is excluded due to lack of significant DARs."
)

```