---
title: "ArchR_1.0.3_H4H_installation"
author: "Ronald Wu"
date: '2025-03-31'
output: html_document
---

# Introduction
This document describes the installation of ArchR 1.0.3 on the H4H cluster. The installation is based on the renv lock file provided by the ArchR team, which is compatible with R 4.4. 
On the H4H cluster, the R version is R 4.4.1, which is not exactly the R/4.4 that is expected from the lock file, however, we will test if this is still okay.

We previously installed Arch 1.0.2 on the cluster and became stuck at the macs peak calling. Hopefully, this has been resolved.
##  Cluster commands
```{bash cluster commands}
module load R/4.4.1
module load R/4.1.0
R
```


## install renv
Here we install renv on the cluster in our personal library. The precompiled renv lock file will be used to manage dependencies.
```{R install renv}
install.packages("renv")
# personal library: ‘/cluster/home/rwu/R/x86_64-conda-linux-gnu-library/4.4’
```

##  Install ArchR 1.0.3 on H4H using precompiled renv to manage dependencies
```{r install ArchR, eval = FALSE}
library(renv)
dir.create(path = "./ArchR103", showWarnings = FALSE)
setwd("./ArchR103") # for 4.1.0

# download ArchR renv lock file for R 4.1.0
download.file(url = "https://pub-9ae435458ecc412abbbc9420a502ec38.r2.dev/renv_4_1.lock", destfile = "./renv.lock")

# restore the project from the lock file
## this will take a while to download and install
renv::init()
```

```{r install ArchR, eval = FALSE}
library(renv)
dir.create(path = "./ArchR103_4", showWarnings = FALSE)
setwd("./ArchR103_4") # for 4.4.1

# download ArchR renv lock file for R 4.4.1
download.file(url = "https://pub-9ae435458ecc412abbbc9420a502ec38.r2.dev/renv.lock", destfile = "./renv.lock")

# restore the project from the lock file
## this will take a while to download and install
renv::init()
```

# custom renv::install("XML")
Since we got error messages about the XML package during renv::init(), we will install the XML package manually.

```{r custom renv::install("XML"), eval = FALSE}
renv::install("XML") # These packages will be installed into "~/R/x86_64-conda-linux-gnu-library/4.4".
renv::install("gert") # gert 2.1.5 will be installed
renv::install("sf")
```

# install units; 

Had to use conda to install udunits2, which is a dependency for the units package. For some reason, renv could not access the udunits2 library, even though it was installed in the /usr/lib64, and the 
header was found in the /usr/include/udunits2 dir.
## install udunits2 with conda first
```{bash conda install udunits2}
# Create a new conda environment in your home directory
conda create -p ~/myenv
# Activate the new environment
conda activate ~/myenv
# Install udunits2 in your personal environment
conda install -c conda-forge udunits2

```

```{r install units}
config <- c(units="--with-udunits2-lib=/cluster/home/rwu/myenv/lib --with-udunits2-include=/cluster/home/rwu/myenv/include")
options(configure.args=config)
renv::install("units")
```

# Install libxml2
```{bash libxml2 installation}
wget https://gitlab.gnome.org/GNOME/libxml2/-/archive/v2.14.0/libxml2-v2.14.0.tar.gz

tar -xvzf libxml2-v2.14.0.tar.gz
cd libxml2-v2.14.0
./configure --prefix=/cluster/home/rwu/bin/libxml2-v2.14.0 --disable-static --with-history


./configure --prefix=$HOME/.local --with-python=/usr/bin/python3.9

```

# Install sf dependencies
Claude thinks we can't find libproj.so.22 based on the error messages after running renv::init().

```{bash sf dependencies}
# install sf dependencies
conda install -c conda-forge gdal geos proj
```

Double check if R can find these files.
```{bash sf dependencies}
conda activate ~/myenv
conda list -p $CONDA_PREFIX proj
conda list -p $CONDA_PREFIX gdal
conda list -p $CONDA_PREFIX geos

find $CONDA_PREFIX -type f -name "*.so*" | grep proj 
#/cluster/home/rwu/myenv/lib/libproj.so.25.9.6.0

# tell R where to find libraries and other environment variables
export LD_LIBRARY_PATH=/cluster/home/rwu/myenv/lib:$LD_LIBRARY_PATH
export PROJ_LIB=/cluster/home/rwu/myenv/share/proj
export PKG_CONFIG_PATH=/cluster/home/rwu/myenv/lib/pkgconfig:$PKG_CONFIG_PATH
```
# packages in environment at /cluster/tools/software/rocky9/R/4.4.1:
#
# Name                    Version                   Build  Channel
proj                      9.6.0                h0054346_0    conda-forge
gdal                      3.10.2          py313h49478cc_5    conda-forge
libgdal-core              3.10.2               hae73b24_5    conda-forge
geos                      3.13.1               h97f6797_0    conda-forge

```{r try sf install now}
Sys.setenv(LD_LIBRARY_PATH = paste0("/cluster/home/rwu/myenv/lib:", Sys.getenv("LD_LIBRARY_PATH")))
Sys.setenv(PROJ_LIB = "/cluster/home/rwu/myenv/share/proj")
Sys.setenv(PKG_CONFIG_PATH = paste0("/cluster/home/rwu/myenv/lib/pkgconfig:", Sys.getenv("PKG_CONFIG_PATH")))

renv::install("sf", configure.args = "--with-proj-lib=/cluster/home/rwu/myenv/lib --with-proj-include=/cluster/home/rwu/myenv/include")
```

# Install ArchR
Here we will install ArchR after loading and running a R session in the ArchR103_4 directory. The renv environment will be loaded first, since we previously initialized it.

```{r install ArchR}
# in build node (with internet connection)
library(devtools)
library(BiocManager)
devtools::install_github("GreenleafLab/ArchR", ref="master", repos = BiocManager::repositories())
```
# Loading ArchR
```{bash init first time}
# navigate to working directory
cd ArchR103_4
module load R/4.4.1

# `r studio` will not work, so we have to use R in the terminal
# load renv and ArchR
library(renv)
renv::init()
library(BiocManager) # note warning about no internet connection
renv::init()
library(ArchR)

path.package("ArchR") # "/cluster/home/rwu/ArchR103_4/renv/library/linux-rocky-9.4/R-4.4/x86_64-conda-linux-gnu/ArchR"
```


# libflexiblas.so.3.0.0 error
Got an error when loading ArchR, one of the matrix package dependencies cannot be found. 
libflexiblas.so.3: cannot open shared object file:
```{r libflexiblas.so.3.0.0 error}
# just install the R package
renv:install("flexiblas")

conda install -c conda-forge openblas
```

cluster/home/rwu/myenv/lib/libopenblas.so
/cluster/home/rwu/myenv/lib/libopenblas.so.0